
Here, we view how each asset moves in the economies they are in.

## Structure

**NCE**: 

Here, we are using Fama's (1970) Efficient Market Hypothesis (EMH) which states that prices fully reflect all available information [2]. Because information arrives as unpredictable, exogenous shocks, price updates are independent and identically distributed ($i.i.d.$). This reduces the asset's structural state to a trivial, memoryless coordinate point on a flat Euclidean space.

Log returns are used as empirical evidence of such state [a1]. 

With Log returns, we can use a rolling window of mean, variance, skewneses, and kurtosis.


In short, we can compress it down these:
 1. Raw Input: Adjusted Closing Price on `UAL`, `PAL`, `BTC-USD`, and `SPY`.
 2. Enriched State: Log Return.
 3. Compression: Rolling Window of mean, variance, skewness, and kurtosis.

**Objective**:

**Code**:
```{r price-structure-df-nce}

rolling_window_size <- 21
asset_list <- c("UAL", "PAL", "BTC", "SPY")

# 1. Dynamic Rolling Moment Generation
price_structure_df <- master_df %>%
  # Rename up front to simplify downstream dynamic column naming
  select(
    Date, 
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj
  ) %>%
  # Apply all 4 functions to all 4 assets dynamically using R 4.1+ lambdas
  mutate(
    across(
      all_of(asset_list),
      list(
        roll_mean = \(x) rollapplyr(x, rolling_window_size, mean, fill = NA),
        roll_var  = \(x) rollapplyr(x, rolling_window_size, var, fill = NA),
        roll_skew = \(x) rollapplyr(x, rolling_window_size, skewness, fill = NA),
        roll_kurt = \(x) rollapplyr(x, rolling_window_size, kurtosis, fill = NA)
      ),
      .names = "{.col}_{.fn}" # Dynamically names them e.g., UAL_roll_mean
    )
  ) %>%
  # Remove incomplete windows
  filter(!is.na(UAL_roll_mean))

print(price_structure_df)


# 2. Dynamic Plotting Loop (Bug Fixed)
price_structure_plot_list <- list()

for (asset in asset_list) {
  
  # Select Date and ONLY the columns matching the current loop's asset
  plot_df <- price_structure_df %>%
    select(Date, starts_with(paste0(asset, "_"))) %>%
    pivot_longer(
      cols = -Date, 
      names_to = "Metric", 
      values_to = "Value"
    )
  
  p_structure <- ggplot(plot_df, aes(x = Date, y = Value, color = Metric)) +
    geom_line(alpha = 0.7) +
    facet_wrap(~Metric, scales = "free_y", ncol = 1) +
    theme_minimal() +
    labs(
      title = paste(asset, "Structural State: Rolling Moments (21-Day)"),
      y = "Moment Value"
    )
  
  price_structure_plot_list[[asset]] <- ggplotly(p_structure, height=1000)
}

# Render the list of interactive plots
tagList(price_structure_plot_list)
```

**Output**:

**Financial Interpretation**:

**CAS**: Treats the shape of the return distribution itself is the object of interest.
 
 
This follows from Mandelbrot's work [5], which states that financial returns are non-Gaussian and follow a stable Paretian distribution where variance is infinite or non-convergent over local windows. By lifting the returns into a Jet Bundle (position, velocity, curvature), the system recognizes that two assets sharing identical log returns can be on entirely different paths depending on their structural velocity (momentum) and curvature (volatility acceleration).

 
 1. Raw Input: Adjusted Closing Price on `UAL`, `PAL`, `BTC-USD`, and `SPY`.
 2. Enriched State: Jet Bundle Lift of Low Return, with its Velocity, and Curvature.
 3. Compression: Rolling Window, of 21 days (Days in a month for financial trading), of its mean, variance, skewness, and kurtosis.


**Objective**:

**Code**:

**Output**:

**Financial Interpretation**:

```{r price-structure-df-cas}
rolling_window_size <- 21
logadj_cols <- names(master_df)[str_detect(names(master_df), "\\.LogAdj$")]

# --- 1. JET BUNDLE LIFT GENERATION ---
jet_bundle_raw <- master_df %>%
  select(Date, all_of(logadj_cols)) %>%
  rename_with(~ str_replace(.x, "\\.LogAdj$", ".Position"), all_of(logadj_cols)) %>%
  mutate(
    across(ends_with(".Position"), ~ .x - lag(.x), 
           .names = "{str_remove(.col, '\\\\.Position')}.Velocity"),
    across(ends_with(".Position"), ~ .x - 2 * lag(.x) + lag(.x, 2), 
           .names = "{str_remove(.col, '\\\\.Position')}.Acceleration")
  )


# --- 2. MULTI-DIMENSIONAL ROLLING MOMENT COMPUTATION ---
rolled_phase_space <- jet_bundle_raw %>%
  pivot_longer(
    cols = -Date,
    names_to = c("Asset", "Dimension"),
    names_sep = "\\."
  ) %>%
  group_by(Asset, Dimension) %>%
  mutate(
    mean = rollapplyr(value, rolling_window_size, mean, fill = NA),
    var  = rollapplyr(value, rolling_window_size, var, fill = NA),
    skew = rollapplyr(value, rolling_window_size, skewness, fill = NA),
    kurt = rollapplyr(value, rolling_window_size, kurtosis, fill = NA)
  ) %>%
  ungroup() %>%
  filter(!is.na(mean)) %>%
  select(-value)


# --- 3. RESHAPE & CALCULATE LEAD (NEXT POINT) ACCELERATION ---
phase_space_4d <- rolled_phase_space %>%
  pivot_wider(
    names_from = Dimension,
    values_from = c(mean, var, skew, kurt),
    names_glue = "{Dimension}_{.value}"
  ) %>%
  filter(Date >= as.Date("2019-01-01") & Date <= as.Date("2025-12-31")) %>%
  mutate(Asset = str_remove(Asset, "^`|`$")) %>%
  # Group by Asset so we don't lead values across different assets
  group_by(Asset) %>%
  mutate(
    # Shifting the future acceleration variance back to the current row
    Next_Acceleration_Var = lead(Acceleration_var, 1)
  ) %>%
  ungroup()


# --- 4. GENERATE 3D PLOTS WITH LEAD-COLORED SEGMENTS ---
all_assets_4d_plots <- list()

for (current_asset in unique(phase_space_4d$Asset)) {
  
  asset_data <- phase_space_4d %>% filter(Asset == current_asset)
  
  p_4d <- plot_ly(asset_data, 
          x = ~Date, 
          y = ~Position_mean,  
          z = ~Velocity_mean,  
          
          # Map color to the next point's acceleration variance
          color = ~Next_Acceleration_Var, 
          colors = viridis(256, option = "magma"),
          type = 'scatter3d', 
          mode = 'lines+markers',
          
          # Markers setup
          marker = list(
            size = 3, 
            opacity = 0.85
          ),
          
          # Line setup: No 'color = black' override so it maps to color scale
          line = list(
            width = 3.5, # Slightly thicker lines to show off the segment colors
            opacity = 0.9
          )
  ) %>%
    layout(
      title = paste(current_asset, "System Phase Trajectory (Colored by Next Step Accel)"),
      scene = list(
        xaxis = list(
          title = 'Date (1D)',
          type = 'date',
          range = c("2019-01-01", "2025-12-31")
        ),
        yaxis = list(title = '21d Mean Position'),
        zaxis = list(title = '21d Mean Velocity'),
        
        aspectmode = "manual",
        aspectratio = list(x = 3, y = 1, z = 1)
      ),
      margin = list(l = 0, r = 0, b = 0, t = 40)
    )
  
  all_assets_4d_plots[[current_asset]] <- p_4d
}

htmltools::tagList(all_assets_4d_plots)
```


## Dynamics

**NCE**: Returns are draws from a fixed Gaussian, so risk scales predictably with thr square root of time.
 
 1. Transition (Conservative): Modeled via Black-Scholes Diffusion where price evolution is a continuous random walk driven by standard Brownian motion ($dW_t$).
 2. Grouped Context of Transitions: Portfolio risk across multiple time horizons ($T$) aggregates linearly according to the standard square-root-of-time scaling rule: $\sigma_T = \sigma_1 \times \sqrt{T}$. The temporal progression contains zero memory or feedback loops.
 
 
```{r rolling-dynamic-plot}
price_dynamics_df <- master_df %>%
  select(Date, UAL.LogAdj, PAL.LogAdj, `BTC-USD.LogAdj`, SPY.LogAdj) %>%
  
  # Mutate to build longer-horizon compound paths (Rolling sums of daily log returns)
  mutate(
    # UAL Multi-horizon paths
    UAL_5d_return  = rollapplyr(UAL.LogAdj, 5, sum, fill = NA),
    UAL_21d_return = rollapplyr(UAL.LogAdj, 21, sum, fill = NA),
    
    # PAL Multi-horizon paths
    PAL_5d_return  = rollapplyr(PAL.LogAdj, 5, sum, fill = NA),
    PAL_21d_return = rollapplyr(PAL.LogAdj, 21, sum, fill = NA),
    
    # BTC Multi-horizon paths
    BTC_5d_return  = rollapplyr(`BTC-USD.LogAdj`, 5, sum, fill = NA),
    BTC_21d_return = rollapplyr(`BTC-USD.LogAdj`, 21, sum, fill = NA),
    
    # SPY Multi-horizon paths
    SPY_5d_return  = rollapplyr(SPY.LogAdj, 5, sum, fill = NA),
    SPY_21d_return = rollapplyr(SPY.LogAdj, 21, sum, fill = NA)
  )
# --- DYNAMICS STEP ---
# 1. Compute empirical variance for each horizon (uncorrected by T)
empirical_summary <- price_dynamics_df %>%
  summarise(
    `1-Day`  = var(PAL.LogAdj, na.rm = TRUE),
    `5-Day`  = var(PAL_5d_return, na.rm = TRUE),
    `21-Day` = var(PAL_21d_return, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Horizon", values_to = "Empirical_Variance")

# 2. Extract the base 1-day baseline variance (our variance unit scale sigma^2)
base_var <- empirical_summary %>% 
  filter(Horizon == "1-Day") %>% 
  pull(Empirical_Variance)

# 3. Inject the NCE Black-Scholes Linear Grouped Progression Scale
scaling_df <- empirical_summary %>%
  mutate(
    # Map Horizon strings to their integer T parameters
    T_days = case_when(
      Horizon == "1-Day"  ~ 1,
      Horizon == "5-Day"  ~ 5,
      Horizon == "21-Day" ~ 21
    ),
    # NCE Definition: Variance scales linearly with time (sigma_1^2 * T)
    NCE_Theoretical_Variance = base_var * T_days
  ) %>%
  # Pivot to long format so ggplot can create separate color layers
  pivot_longer(
    cols = c(Empirical_Variance, NCE_Theoretical_Variance), 
    names_to = "Type", 
    values_to = "VarianceValue"
  ) %>%
  mutate(
    Type = case_when(
      Type == "Empirical_Variance"         ~ "Empirical (Actual Data)",
      Type == "NCE_Theoretical_Variance"   ~ "NCE: Black-Scholes Theoretical Line"
    )
  )

# 4. Generate the Visualization highlighting the grouping context deviation
p_dynamics <- ggplot(scaling_df, aes(x = factor(T_days), y = VarianceValue, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "darkred")) +
  theme_minimal() +
  labs(
    title = "PAL Dynamics: Empirical vs. Black-Scholes Linear Risk Scaling",
    subtitle = "NCE requires empirical columns to exactly match theoretical anchors",
    x = "Time Horizon (Days)",
    y = "Total Return Variance",
    fill = "Framework Layer"
  )

ggplotly(p_dynamics)
```
 
**CAS**: Variance scaling across horizons deviates from that straight line — evidence of memory, clustering, or regime persistence rather than pure noise.
 
 1. Transition (Diffusion): Modeled as a continuous Flow on a Manifold splitting into a Symplectic core (conservative supply/demand matching) and Contact geometry (dissipative, thermodynamic features such as liquidation cascades and market-maker risk-premium bleed).
 2. Grouped Context of Transitions: Follows a multi-scale variance scaling rule: $\text{Var}(r_{t+T}) \propto T^{2H}$, where the Hurst Exponent ($H \neq 0.5$) captures temporal memory. Transitions group into persistent regimes where high-volatility events cluster endogenously (Mandelbrot 1963).

```{r price-movement-cas-dynamics}

cas_dynamics_df <- jet_bundle_raw %>%
  select(Date, PAL.Position, PAL.Velocity, PAL.Acceleration) %>%
  mutate(
    # Position multi-horizon paths
    PAL_pos_5d  = rollapplyr(PAL.Position, 5, sum, fill = NA),
    PAL_pos_21d = rollapplyr(PAL.Position, 21, sum, fill = NA),
    
    # Velocity multi-horizon paths
    PAL_vel_5d  = rollapplyr(PAL.Velocity, 5, sum, fill = NA),
    PAL_vel_21d = rollapplyr(PAL.Velocity, 21, sum, fill = NA),
    
    # Curvature multi-horizon paths
    PAL_cur_5d  = rollapplyr(PAL.Acceleration, 5, sum, fill = NA),
    PAL_cur_21d = rollapplyr(PAL.Acceleration, 21, sum, fill = NA)
  )

# --- CONSOLIDATE AND GENERATE SCALING MATRIX ---
# Gather empirical variances across every coordinate space tier
cas_empirical <- cas_dynamics_df %>%
  summarise(
    # Position Horizons
    pos_1d  = var(PAL.Position, na.rm = TRUE),
    pos_5d  = var(PAL_pos_5d, na.rm = TRUE),
    pos_21d = var(PAL_pos_21d, na.rm = TRUE),
    
    # Velocity Horizons
    vel_1d  = var(PAL.Velocity, na.rm = TRUE),
    vel_5d  = var(PAL_vel_5d, na.rm = TRUE),
    vel_21d = var(PAL_vel_21d, na.rm = TRUE),
    
    # Curvature Horizons
    cur_1d  = var(PAL.Acceleration, na.rm = TRUE),
    cur_5d  = var(PAL_cur_5d, na.rm = TRUE),
    cur_21d = var(PAL_cur_21d, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Key", values_to = "Empirical_Variance") %>%
  separate(Key, into = c("Space", "Horizon"), sep = "_")

# Map integer T scales and baseline parameters to compute NCE theoretical projections
cas_scaling_df <- cas_empirical %>%
  mutate(
    T_days = case_when(
      Horizon == "1d"  ~ 1,
      Horizon == "5d"  ~ 5,
      Horizon == "21d" ~ 21
    ),
    Space_Label = case_when(
      Space == "pos" ~ "Position Space (Returns)",
      Space == "vel" ~ "Velocity Space (Momentum)",
      Space == "cur" ~ "Acceleration Space (Curvature)"
    )
  ) %>%
  # Dynamically fetch the 1-Day baseline variance for each respective space tier
  group_by(Space) %>%
  mutate(
    NCE_Theoretical_Variance = Empirical_Variance[Horizon == "1d"] * T_days
  ) %>%
  ungroup() %>%
  # Reshape for multi-layered side-by-side column mapping
  pivot_longer(
    cols = c(Empirical_Variance, NCE_Theoretical_Variance),
    names_to = "Framework_Type",
    values_to = "VarianceValue"
  ) %>%
  mutate(
    Framework_Type = case_when(
      Framework_Type == "Empirical_Variance"       ~ "Empirical (Actual CAS Dynamics)",
      Framework_Type == "NCE_Theoretical_Variance" ~ "NCE Benchmark (Linear Scaling)"
    )
  )
plotly_plots <- list()

# --- PLOT CAS DYNAMICS SEPARATELY BY SPACE TIER ---
for (current_space in unique(cas_scaling_df$Space_Label)) {
  
  space_dynamics_data <- cas_scaling_df %>% filter(Space_Label == current_space)
  
  p_cas_dynamics <- ggplot(space_dynamics_data, aes(x = factor(T_days), y = VarianceValue, fill = Framework_Type)) +
    geom_col(position = "dodge", alpha = 0.8) +
    scale_fill_manual(values = c("darkblue", "darkgrey")) +
    theme_minimal() +
    labs(
      title = paste("CAS Dynamics Scaling:", current_space),
      subtitle = "Significant differences between actual and benchmark indicate memory structures",
      x = "Time Horizon (Days)",
      y = "Total Coordinate Variance",
      fill = "Model Hypothesis"
    )
  
  plotly_plots[[current_space]] <- ggplotly(p_cas_dynamics)
}
tagList(plotly_plots)
```

```{r}

cas_true_dynamics_df <- cas_scaling_df %>%
  # Filter to look at unique empirical horizons to calculate H
  filter(Framework_Type == "Empirical (Actual CAS Dynamics)") %>%
  group_by(Space_Label) %>%
  mutate(
    # Estimate 2H via a local log-linear regression slope: ln(Var) vs ln(T)
    # Since Var = C * T^(2H) -> log(Var) = log(C) + 2H*log(T)
    log_T = log(T_days),
    log_Var = log(VarianceValue),
    
    # Extract the slope (2H) using basic linear algebra syntax
    two_H = cov(log_Var, log_T) / var(log_T),
    Hurst_Est = two_H / 2
  ) %>%
  ungroup()

# 1. Isolate the base 1-Day empirical variance for each Space tier to prevent recycling errors
base_variance_map <- cas_scaling_df %>%
  filter(Framework_Type == "Empirical (Actual CAS Dynamics)", Horizon == "1d") %>%
  select(Space, Base_Var = VarianceValue)

# 2. Join the base variance map and the estimated Hurst exponents back to the scaling dataset
plotting_matrix_df <- cas_scaling_df %>%
  left_join(base_variance_map, by = "Space") %>%
  left_join(
    cas_true_dynamics_df %>% select(Space, Hurst_Est) %>% unique(),
    by = "Space"
  )

# 3. Create the distinct CAS Fractional scaling curves using your exact formula: Var_1d * T^(2H)
cas_lines_df <- plotting_matrix_df %>%
  filter(Framework_Type == "NCE Benchmark (Linear Scaling)") %>%
  mutate(
    Framework_Type = "CAS Fractional Target (T^{2H})",
    # Override NCE linear scaling with Mandelbrot's power law profile
    VarianceValue = Base_Var * (T_days^(2 * Hurst_Est))
  )

# 4. Bind the new fractional curve back into your master visual comparison matrix
plotting_matrix_df <- bind_rows(cas_scaling_df, cas_lines_df) %>%
  left_join(
    unique(cas_true_dynamics_df[, c("Space_Label", "Hurst_Est")]), 
    by = "Space_Label", 
    suffix = c("", ".y")
  ) %>%
  # Fill missing Hurst_Est text descriptors generated by the bind_rows step
  group_by(Space) %>%
  fill(Hurst_Est, Space_Label, .direction = "updown") %>%
  ungroup()

plotly_plots_2 <- list()

# --- PLOT THE THREE-WAY PARADIGM COMPARISON ---
for (current_space in unique(plotting_matrix_df$Space_Label)) {
  
  space_data <- plotting_matrix_df %>% filter(Space_Label == current_space)
  h_val <- round(unique(space_data$Hurst_Est), 3)
  
  p_cas_true <- ggplot(space_data, aes(x = factor(T_days), y = VarianceValue, fill = Framework_Type)) +
    geom_col(position = "dodge", alpha = 0.85) +
    scale_fill_manual(values = c("darkblue", "darkorange", "darkgrey")) +
    theme_minimal() +
    labs(
      title = paste("True CAS Dynamics Scaling:", current_space),
      subtitle = paste0("Empirical Memory Profile vs. Linear Benchmarks (Estimated Hurst H = ", h_val, ")"),
      x = "Time Horizon (Days)",
      y = "Total Coordinate Variance",
      fill = "Model Class"
    )
  
  plotly_plots_2[[current_space]] <- ggplotly(p_cas_true)
}

tagList(plotly_plots_2)
```

```{r}
# 1. Isolate clean continuous log returns for your target asset (e.g., PAL)
returns_vector <- master_df %>%
  filter(!is.na(PAL.LogAdj)) %>%
  pull(PAL.LogAdj)

# 2. Compute a complete continuous spectrum of empirical variances across horizons 1 to 21
horizons <- 1:21
continuous_empirical_var <- sapply(horizons, function(k) {
  k_day_returns <- rollapplyr(returns_vector, k, sum, fill = NA)
  var(k_day_returns, na.rm = TRUE)
})

# 3. Extract base 1-day variance baseline unit
base_var <- continuous_empirical_var[1]

# 4. Programmatically estimate the exact Hurst Exponent (H) via log-linear fit
log_T <- log(horizons)
log_Var <- log(continuous_empirical_var)
fit_hurst <- lm(log_Var ~ log_T)
estimated_H <- as.numeric(coef(fit_hurst)[2] / 2)

# 5. Build the complete geometric coordinate dataset mapping the three trajectories
dynamics_manifold_df <- tibble(
  Horizon_T = horizons,
  Empirical_Path = continuous_empirical_var,
  NCE_Linear_Path = base_var * horizons,                     # Var(T) = sigma^2 * T
  CAS_Fractal_Path = base_var * (horizons^(2 * estimated_H)) # Var(T) = sigma^2 * T^(2H)
) %>%
  # Pivot into long format for line geom mapping
  pivot_longer(
    cols = -Horizon_T,
    names_to = "Paradigm_Layer",
    values_to = "Total_Variance"
  ) %>%
  mutate(
    Paradigm_Layer = case_when(
      Paradigm_Layer == "Empirical_Path"   ~ "Actual Empirical Path (Market Data)",
      Paradigm_Layer == "NCE_Linear_Path"  ~ "NCE / Black-Scholes Path (Linear)",
      Paradigm_Layer == "CAS_Fractal_Path" ~ paste0("CAS Fractal Curve (Hurst H = ", round(estimated_H, 3), ")")
    )
  )

# 6. Generate the Geometric Line Plot
p_dynamics_curves <- ggplot(dynamics_manifold_df, 
                            aes(x = Horizon_T, y = Total_Variance, color = Paradigm_Layer, linetype = Paradigm_Layer)) +
  geom_line(size = 1.2, alpha = 0.9) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = c("darkblue", "darkorange", "red")) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
  theme_minimal() +
  labs(
    title = "PAL Dynamics Manifold: Multi-Scale Variance Scaling Curves",
    subtitle = "Visualizing the geometric path difference between memoryless diffusion and complex systems",
    x = "Time Horizon Scaling (Days, T)",
    y = "Total Aggregated Return Variance",
    color = "Theoretical Framework",
    linetype = "Theoretical Framework"
  ) +
  theme(legend.position = "bottom")

# Render interactively for your integrated HTML report
ggplotly(p_dynamics_curves)
```

## Inference

**NCE**: Checks if short-horizon volatility is sufficient to price any longer-horizon risk. 
 
 1. Gathering Evidence: Goodness of Fit test (Jarque-Bera) for normality, and the Lo-MacKinlay Variance Ratio Test ($VR(k)$) to evaluate variance scaling across multiple horizons ($k$).
 2. Decision Rule: 
    - **IF** Jarque-Bera fails to reject the null hypothesis of normality ($p > 0.05$) **AND** $VR(k) \approx 1$ across all designated horizons ($k$):
      - Then we accept the NCE framework. Daily volatility is sufficient to price longer-horizon risk, confirming a flat, memoryless Euclidean price space.
    - **IF** Jarque-Bera rejects normality ($p \le 0.05$) **OR** $VR(k)$ significantly drifts away from $1$ as $k$ increases:
      - Then the NCE assumption fails. Short-horizon volatility calculations systematically misprice or understate long-horizon structural risk.
 3. Verdict: `TBD`

```{r inference-}

# --- INFERENCE STEP ---
# Pick an asset to test dynamically (or loop through them all)
target_asset <- "PAL"

# Gather vectors from your completed dynamics/structure frames
daily_returns  <- na.omit(price_dynamics_df[[paste0(target_asset, ".LogAdj")]])
ret_5d         <- na.omit(price_dynamics_df[[paste0(target_asset, "_5d_return")]])
ret_21d        <- na.omit(price_dynamics_df[[paste0(target_asset, "_21d_return")]])

# Execute NCE Significance Tests
jb_test   <- jarque.bera.test(daily_returns)
p_val_jb  <- jb_test$p.value

# Form the Variance Ratios to check against 1
vr_5d  <- var(ret_5d) / (5 * var(daily_returns))
vr_21d <- var(ret_21d) / (21 * var(daily_returns))

# Execute CAS Significance Tests
# Note:Hurst Exponents via Rescaled Range, and Hill Estimators for heavy tails.
hurst_val <- pracma::hurstexp(daily_returns, display = FALSE)$Hs
# tail_index <- custom_hill_estimator(daily_returns) 

# 4. Print Outputs to clear the TBD fields
cat("--- INFERENCE REPORT FOR:", target_asset, "---\n",
    "Jarque-Bera p-value: ", p_val_jb, "\n",
    "Variance Ratio (5-Day Target = 1): ", vr_5d, "\n",
    "Variance Ratio (21-Day Target = 1): ", vr_21d, "\n",
    "Hurst Exponent (Target = 0.5): ", hurst_val, "\n")
```


**CAS**: Check if that the tail risk is systematically underpriced by that approach, because rare, large moves happen more often than a Gaussian model would predict.
 
 1. Evidence:
   - Hurst Exponent ($H$) via Rescaled Range Analysis to evaluate long-range dependency and clustering ($H \neq 0.5$).
   - Hill’s Estimator for Tail Index ($\alpha$) to compute the power-law decay exponent of the return distribution's extreme tails.
 2. Decision Rule:
    - **IF** the estimated Hurst exponent deviates significantly from equilibrium ($H > 0.5$ for persistence or $H < 0.5$ for anti-persistence) **AND** Hill's tail index indicates heavy tails ($\alpha < 3$, where variance or kurtosis begins to fail standard convergence limits):
      - Then we validate the CAS paradigm. The asset demonstrates structural memory and fat tails, proving that return distributions cannot be compressed into a memoryless scalar without ignoring catastrophic systemic tail risk.
    - **IF** $H \approx 0.5$ **AND** $\alpha \ge 3$ (approaching Gaussian tail decay thresholds):
      - Then the asset behaves like a standard equilibrium system. Deviations are transient noise, and the traditional random walk approximation remains robust.
 3. Verdict: `TBD`
 
```{r cas-inference-engine, echo=TRUE, message=FALSE, warning=FALSE}
# ==============================================================================
# CAS INFERENCE ENGINE: EVIDENCE, DECISION RULE, & PROGRAMMATIC VERDICT
# ==============================================================================

# 1. Isolate target asset log returns (e.g., PAL) and strip NA values
returns_vector <- master_df %>%
  filter(!is.na(PAL.LogAdj)) %>%
  pull(PAL.LogAdj)

# ------------------------------------------------------------------------------
# STEP 1: GATHERING EVIDENCE (STATISTICAL COMPUTATION)
# ------------------------------------------------------------------------------

### A. Hurst Exponent (H) via Log-Variance Multi-Scale Regression
# We compute variance across horizons 1 to 21 to capture the exact memory decay profile
horizons <- 1:21
horizon_variances <- sapply(horizons, function(k) {
  # Generate k-day rolling compounded paths
  k_day_returns <- zoo::rollapplyr(returns_vector, k, sum, fill = NA)
  var(k_day_returns, na.rm = TRUE)
})

# Log-linear fit: ln(Var) = ln(C) + 2H*ln(T)
log_T <- log(horizons)
log_Var <- log(horizon_variances)
fit_hurst <- lm(log_Var ~ log_T)
estimated_H <- as.numeric(coef(fit_hurst)[2] / 2)


### B. Hill's Estimator for Tail Index (alpha)
# We isolate the upper extreme tail (top 5% positive tail shocks) to extract the power law decay
tail_threshold <- 0.05 
sorted_returns <- sort(returns_vector, decreasing = TRUE)
n_tail <- floor(length(sorted_returns) * tail_threshold)

# Extract extreme threshold anchor
x_threshold <- sorted_returns[n_tail + 1]
tail_shocks <- sorted_returns[1:n_tail]

# Hill Estimator Formula: 1 / [ (1/k) * Sum( ln(x_i / x_{k+1}) ) ]
hill_alpha <- 1 / (mean(log(tail_shocks / x_threshold)))

# Print Computed Evidence Summary Table
cat("===========================================\n",
    "      CAS EVIDENCE GATHERING MATRIX        \n",
    "===========================================\n",
    "Target Asset Component: PAL\n",
    "Estimated Hurst Exponent (H) : ", round(estimated_H, 4), "\n",
    "Hill's Tail Decay Index (α)  : ", round(hill_alpha, 4), "\n",
    "-------------------------------------------\n\n")


# ------------------------------------------------------------------------------
# STEP 2 & 3: APPLICATION OF DECISION RULE & VERDICT SPECIFICATION
# ------------------------------------------------------------------------------

# Establish strict structural thresholds based on your framework rules
hurst_lower_bound <- 0.45
hurst_upper_bound <- 0.55
alpha_fat_tail_threshold <- 3.0

if ((estimated_H > hurst_upper_bound || estimated_H < hurst_lower_bound) && hill_alpha < alpha_fat_tail_threshold) {
  
  # --- BRANCH 1: VALIDATE COMPLEX ADAPTIVE SYSTEMS PARADIGM ---
  cas_verdict <- "VALIDATED (REJECT NEOCLASSICAL EQUILIBRIUM)"
  
  cas_interpretation <- paste0(
    "VERDICT ANALYSIS: The CAS paradigm is empirically confirmed.\n",
    "Reasoning: The asset demonstrates strong anomalous scaling (H = ", round(estimated_H, 3), 
    if_else(estimated_H < 0.5, " indicating anti-persistent mean-reversion", " indicating persistent trend-reinforcement"), 
    ") coupled with extreme fat-tailed decay (α = ", round(hill_alpha, 3), ").\n",
    "Systemic Impact: Because α < 3, the traditional Gaussian variance boundaries are violated.\n",
    "Compressing this asset into standard local variance metrics or flat i.i.d assumptions will systematically\n",
    "underprice options structures and risk profiles, rendering standard Black-Scholes risk models blind\n",
    "to endogenous, structural liquidation cascades."
  )
  
} else {
  
  # --- BRANCH 2: VALIDATE NEOCLASSICAL EQUILIBRIUM ---
  cas_verdict <- "REJECTED (NEOCLASSICAL RANDOM WALK ROBUST)"
  
  cas_interpretation <- paste0(
    "VERDICT ANALYSIS: The Neoclassical Equilibrium framework holds robustly.\n",
    "Reasoning: The asset's scaling footprint closely approximates memoryless Brownian motion (H = ", round(estimated_H, 3), 
    ") and the distribution's tail profile safely matches or exceeds normal exponential decay structures (α = ", round(hill_alpha, 3), ").\n",
    "Systemic Impact: Shocks are transient, exogenous, and uncorrelated. Traditional Value-at-Risk (VaR)\n",
    "and square-root-of-time scaling mechanics provide a dependable framework for structural risk budgeting."
  )
}

# Output programmatically derived Verdict and Detailed Summary
cat("======================================================================\n", 
    "FINAL PROGRAMMATIC INFERENCE VERDICT:", cas_verdict, "\n",
    "======================================================================\n",
    cas_interpretation, "\n")
```
 
## Synthesis
The Price Movement analysis serves as our most granular lens, examining individual asset vectors in isolation. If the empirical verdicts yield deviations from normality and square-root scaling, it proves that "price" is not a simple, self-contained coordinate point. Instead, assets contain intrinsic path dependency and memory, prompting us to step outward into our next lens: how these unique asset states interact with one another under varying regimes.

## Footnote:
[a1] https://ccsenet.org/journal/index.php/ijsp/article/view/17911 
