
Here, we view how the relationships of the assets moves in the economies they are in.

## Structure

### Correlation and Covariance Matrix
 
 Here, we utilize Markowitz’s Modern Portfolio Theory (MPT) and the foundations of Sharpe’s Capital Asset Pricing Model (CAPM). Risk is fully quantified by the variance-covariance matrix ($\Sigma$), assuming structural errors or shifts are merely random, zero-mean Gaussian noise around a permanent equilibrium.
 
```{r}
returns_only <- master_df %>% 
  select(
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj
  ) %>% 
  drop_na()

cov_matrix <- cov(returns_only)
cor_matrix <- cor(returns_only)

cat("### Covariance Matrix (Sigma)\n")
print(round(cov_matrix, 6))

cat("\n### Correlation Matrix (Rho)\n")
print(round(cor_matrix, 6))

ggcorrplot(
  cor_matrix, 
  hc.order = FALSE,      # No fancy clustering/reordering
  type = "full",         # Full square matrix so the axes line up perfectly
  lab = TRUE,            # Display the correlation coefficients
  lab_size = 4,
  colors = c("#20639B", "white", "#D14949"),
  title = "Standard Static Correlation Matrix"
)
```

**CAS**: 
 
This builds on Bouchaud’s Law of Endogenous Correlation Breakdown[b1], which proves that cross-asset dependencies are non-stationary, state-dependent, and governed by systemic feedback loops. Rather than viewing correlation shifts as statistical noise, the system recognizes them as distinct geometric structures. High-volatility matrices represent entirely separate topological charts on the market manifold that cannot be smoothly mapped using a single, global Euclidean average.
 
 1. Raw Input: Daily log returns matrix for UAL, PAL, BTC-USD, SPY, and the daily conditioning regime marker VIX.
 2. Enriched State: Conditiona covariance matrices $\Sigma_t \vert{} \text{VIX}_t$, explicitly split into separate charts.
 3. Compression: Partitioned states mapped into distinct manifolds: Low-VIX Chart ($\text{VIX} \le 25\text{th percentile}$) and High-VIX Chart ($\text{VIX} \ge 75\text{th percentile}$).


CAS's state is that same matrix conditioned on regime — a High-VIX matrix and a Low-VIX matrix, treated as genuinely different objects, not noise around one "true" matrix.

```{r}
# 1. Prepare data with VIX as our dynamic conditioning regime marker
cas_prepared <- master_df %>%
  select(
    Date,
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj,
    VIX = VIX.Adjusted # Assuming VIX.Close is in your master_df
  ) %>%
  drop_na()

vix_25 <- quantile(cas_prepared$VIX, 0.25, na.rm = TRUE)
vix_75 <- quantile(cas_prepared$VIX, 0.75, na.rm = TRUE)
low_vix_returns <- cas_prepared %>% filter(VIX <= vix_25) %>% select(UAL, PAL, BTC, SPY)
high_vix_returns <- cas_prepared %>% filter(VIX >= vix_75) %>% select(UAL, PAL, BTC, SPY)
cor_low_vix  <- cor(low_vix_returns)
cor_high_vix <- cor(high_vix_returns)

# Plot them side-by-side to visualize the "Endogenous Correlation Breakdown"
library(gridExtra)

plot_low <- ggcorrplot(
  cor_low_vix, 
  type = "full", 
  lab = TRUE, 
  title = paste("Low-VIX Chart (VIX <=", round(vix_25, 1), ")"),
  colors = c("#20639B", "white", "#D14949")
)
plot_high <- ggcorrplot(
  cor_high_vix, 
  type = "full", 
  lab = TRUE, 
  title = paste("High-VIX Chart (VIX >=", round(vix_75, 1), ")"),
  colors = c("#20639B", "white", "#D14949")
)

grid.arrange(plot_low, plot_high, ncol = 2)
```


## Dynamics

### Invariance of Risk

Assumes correlations are stationary and static — the relationship between UAL and SPY today is the same relationship in a crash.  Diversification is permanent. Therefore, the evolution in the asset relationship variance transitions over time are deterministic. When looked as a whole, multiple assets' evolution are modeled as permanently coupled geometric Brownian motions. Co-movements scale symmetrically across all economic cycles, treating structural changes as impossible.

###

Correlations are regime-dependent and tend to converge upward precisely during stress where diversification "disappears" when you need it most.
 
 1. Transitions (Diffusion): Systemic risk behaves like a thermodynamic fluid following a Non-equilibrium Dissipation. In calm regimes or low-VIX, the flow is Symplectic (information-conserving, asset pairs preserve their unique orthogonal vectors). When a threshold is crossed (e.g., a high-VIX shock), margin calls and liquidations force a massive energy transfer (entropy bleed).
 2. Grouped Context of Transitions: Asset transitions are grouped into distinct market volatility regimes. Under high-volatility environments (conditioned by a high `VIX` threshold), individual asset behaviors lose their unique degrees of freedom, and correlations across completely different asset classes (e.g., `UAL`, `PAL`, and `BTC-USD`) shift simultaneously. 


```{r}
# --- HELPER FUNCTION: CORRELATION TO MST ---
build_mst_df <- function(cor_matrix, layout_seed = 42) {
  dist_matrix <- sqrt(2 * (1 - cor_matrix))
  g_full <- graph_from_adjacency_matrix(
    dist_matrix, 
    mode = "undirected", 
    weighted = TRUE, 
    diag = FALSE
  )
  
  # 3. Extract the Minimum Spanning Tree (MST) using Kruskal's algorithm
  g_mst <- minimum.spanning.tree(g_full, weights = E(g_full)$weight)
  set.seed(layout_seed)
  mst_net <- ggnetwork(g_mst, layout = layout_with_kk(g_mst))
  return(mst_net)
}

# --- GENERATE REGIME-SPECIFIC MSTs ---
mst_low  <- build_mst_df(cor_low_vix)
mst_high <- build_mst_df(cor_high_vix)

# --- PLOT THE MSTs ---
plot_mst <- function(mst_data, title_text) {
  ggplot(mst_data, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges(color = "grey50", alpha = 0.7, size = 1) +
    # Edge labels show the physical "distance" (closer to 0 means higher correlation)
    geom_edgetext(aes(label = round(weight, 2)), color = "darkblue", fill = "white", size = 4) +
    geom_nodes(color = "firebrick", size = 12) +
    geom_nodetext(aes(label = name), color = "white", size = 4) +
    theme_blank() +
    labs(title = title_text)
}

grid.arrange(
  plot_mst(mst_low, "Low-VIX MST"),
  plot_mst(mst_high, "High-VIX MST"),
  ncol = 2
)

# --- 1. SETUP PARAMETERS ---
roll_window <- 21
dates <- unique(cas_prepared$Date)[roll_window:nrow(cas_prepared)]

network_metrics <- data.frame(
  Date = as.Date(character()), 
  Mean_Distance = numeric(), 
  Diameter = numeric()
)

# 2. RUN ROLLING NETWORK FILTER WITH NA SAFETY ---
for (i in roll_window:nrow(cas_prepared)) {
  # Extract window returns
  window_data <- cas_prepared[(i - roll_window + 1):i, c("UAL", "PAL", "BTC", "SPY")]
  window_date <- cas_prepared$Date[i]
  window_data_clean <- na.omit(window_data)
  has_variance <- all(sapply(window_data_clean, sd) > 0)
  
  if (nrow(window_data_clean) >= 10 && has_variance) {
    # Compute correlation
    r_matrix <- cor(window_data_clean)
    
    # Map correlation to distance: d = sqrt(2 * (1 - r))
    d_matrix <- sqrt(2 * (1 - r_matrix))
    d_matrix[is.na(d_matrix)] <- 2
    
    # Build the MST
    g_full <- graph_from_adjacency_matrix(d_matrix, mode = "undirected", weighted = TRUE, diag = FALSE)
    g_mst  <- minimum.spanning.tree(g_full, weights = E(g_full)$weight)
    
    # Calculate topology metrics
    mean_dist <- mean(E(g_mst)$weight) # Average closeness of the system
    diameter_val <- diameter(g_mst, directed = FALSE, weights = E(g_mst)$weight) # Furthest distance
    
    network_metrics <- rbind(network_metrics, data.frame(
      Date = window_date,
      Mean_Distance = mean_dist,
      Diameter = diameter_val
    ))
  }
}

network_evolution <- network_metrics %>%
  left_join(select(cas_prepared, Date, VIX), by = "Date")

ggplot(network_evolution, aes(x = Date)) +
  geom_line(aes(y = Mean_Distance, color = "Network Mean Distance (Compression)"), size = 1) +
  geom_line(aes(y = VIX / 100, color = "VIX (Fear Index, Scaled)"), size = 1) +
  scale_y_continuous(
    name = "Network Mean Distance",
    sec.axis = sec_axis(~ . * 100, name = "VIX Level")
  ) +
  scale_color_manual(values = c("forestgreen", "darkred")) +
  theme_minimal() +
  labs(
    title = "Rolling Network Compression vs. Market Fear",
    x = "Date",
    color = "Metric"
  ) +
  theme(legend.position = "bottom")
```

```{r}
# --- 1. SETUP PARAMETERS ---
roll_window <- 21
assets <- c("UAL", "PAL", "BTC", "SPY")

# Initialize expanded data frame to capture both MST and H1 TDA metrics
tda_network_metrics <- data.frame(
  Date = as.Date(character()), 
  Mean_Distance = numeric(), 
  Diameter = numeric(),
  H1_Loop_Count = numeric(),       # Number of multi-asset cycles formed
  Max_H1_Persistence = numeric(),  # Lifespan of the most robust cycle
  Total_H1_Persistence = numeric() # Aggregate structural fragmentation metric
)

# --- 2. RUN ROLLING FILTRATION LOOP ---
for (i in roll_window:nrow(cas_prepared)) {
  
  # Extract localized window returns
  window_data <- cas_prepared[(i - roll_window + 1):i, assets]
  window_date <- cas_prepared$Date[i]
  
  # Clean NA data and verify variance boundaries
  window_data_clean <- na.omit(window_data)
  has_variance <- all(sapply(window_data_clean, sd) > 0)
  
  if (nrow(window_data_clean) >= 10 && has_variance) {
    
    # Compute correlation and map to metric space distance
    r_matrix <- cor(window_data_clean)
    d_matrix <- sqrt(2 * (1 - r_matrix))
    d_matrix[is.na(d_matrix)] <- 2 # Max distance boundary override
    
    # --- MST SKELETON CALCULATIONS ---
    g_full    <- graph_from_adjacency_matrix(d_matrix, mode = "undirected", weighted = TRUE, diag = FALSE)
    g_mst     <- minimum.spanning.tree(g_full, weights = E(g_full)$weight)
    mean_dist <- mean(E(g_mst)$weight) 
    diameter_val <- diameter(g_mst, directed = FALSE, weights = E(g_mst)$weight) 
    
    # --- TDA PERSISTENT HOMOLOGY CALCULATIONS (H1) ---
    # Compute the persistence diagram via Vietoris-Rips filtration over the distance matrix
    vr_diag <- ripsDiag(
      X = d_matrix,
      maxdimension = 1,           # Track H0 (clusters) and H1 (loops)
      maxscale = 2.0,             # Max theoretical distance in this metric space
      dist = "arbitrary",         # Inform package we are passing a raw distance matrix
      printProgress = FALSE
    )

    diag_matrix <- vr_diag$diagram
    h1_features <- diag_matrix[diag_matrix[, "dimension"] == 1, , drop = FALSE]
    
    if (nrow(h1_features) > 0) {
      # Persistence = Death Scale - Birth Scale
      persistence_lifespans <- h1_features[, "Death"] - h1_features[, "Birth"]
      
      h1_count <- nrow(h1_features)
      max_h1_pers <- max(persistence_lifespans)
      total_h1_pers <- sum(persistence_lifespans)
    } else {
      h1_count <- 0
      max_h1_pers <- 0
      total_h1_pers <- 0
    }
    
    # Append combined metrics
    tda_network_metrics <- rbind(tda_network_metrics, data.frame(
      Date = window_date,
      Mean_Distance = mean_dist,
      Diameter = diameter_val,
      H1_Loop_Count = h1_count,
      Max_H1_Persistence = max_h1_pers,
      Total_H1_Persistence = total_h1_pers
    ))
  }
}

# Merge topology metrics with market fear index
tda_evolution <- tda_network_metrics %>%
  left_join(select(cas_prepared, Date, VIX), by = "Date")

# --- 3. VISUALIZE MULTI-ASSET LOOP DESTRUCTION VS VIX ---
ggplot(tda_evolution, aes(x = Date)) +
  geom_line(aes(y = Total_H1_Persistence, color = "Total H1 Loop Persistence (Fragmentation)"), size = 1) +
  geom_line(aes(y = VIX / 100, color = "VIX Level (Scaled)"), size = 1, linetype = "dashed") +
  scale_y_continuous(
    name = "Total H1 Persistence Length",
    sec.axis = sec_axis(~ . * 100, name = "VIX Level")
  ) +
  scale_color_manual(values = c("darkorchid", "darkred")) +
  theme_minimal() +
  labs(
    title = "H1 Loop Lifespans vs. Market Stress",
    subtitle = "Tracking how systemic shocks plug structural holes in the asset manifold",
    x = "Date",
    color = "Metric"
  ) +
  theme(legend.position = "bottom")
```

During the massive volatility spikes in early 2020 (COVID-19 crash) and early 2025, the purple line collapses and pins itself to exactly 0. When the VIX spikes, institutional de-risking, margin calls, and forced liquidations act as a singular, dominant macro force. This force connects all assets (UAL, PAL, BTC-USD, SPY) simultaneously. The global risk factor instantly "plugs" the structural holes in the asset manifold. Because every asset becomes highly correlated with every other asset, multi-dimensional loops cannot form or survive.

Look closely at late 2023 through mid-2024. The VIX drops to its lowest historical baseline in the sample (around 10–15). Simultaneously, the purple line erupts with its densest concentration of high spikes. In a quiet market, the asset manifold is highly fragmented. Assets are no longer bound by a single, overriding macro driver, allowing long-lived topological loops ($H_1$ cycles) to open and persist across wide distance scales. Sector-specific dynamics, idiosyncratic asset behaviors, and independent capital flows are dominating the system. The market is healthy, decoupled, and operating with high degrees of freedom. During these purple spikes, traditional portfolio diversification is performing at maximum efficiency.


## Inference

**NCE**: The Neoclassical paradigm infers that because asset relationships are governed by a stationary covariance matrix, the physical structure of the market network is constant. A diversified portfolio's risk reduction remains stable and can be reliably trusted for long-term strategic asset allocation and Value-at-Risk (VaR) planning.
 
 1. Gathering Evidence: We test the null hypothesis that the covariance structures of the low-stress and high-stress regimes are identical:
  - H0: Sigma_Low-VIX = Sigma_High-VIX
  - If the underlying covariance is stationary, the rolling MST Mean Distance (Dt) must behave as a stationary white noise process fluctuating tightly around a constant, time-invariant mean: H0: Dt = Mean_D + e_t, where e_t ~ N(0, sigma^2). We test this via an Augmented Dickey-Fuller (ADF) test on the rolling MST distance series.
 2. Formal Hypotheses & Decision Rule
      - Null Hypothesis (H0): Structural Stationarity (Rigid Market Skeleton)
        - Jennrich Test fails to reject correlation homogeneity (Jennrich > 0.05) AND
        - Rolling MST Mean Distance is stationary (ADF <= 0.05) AND
        - MST distance is uncorrelated with (VIX corr > 0.05)
        - Conclusion: Stationarity holds. The portfolio shield is static and rigid.
        - The market covariance matrix and network topology are stationary, regime-invariant, and uncorrelated with market stress. A static historical covariance matrix provides a permanently valid portfolio shield.
      - Alternative Hypothesis (H1): Structural Non-Stationarity (Dynamic Regime Shifts)
        - Jennrich Test rejects homogeneity (Jennrich <= 0.05) OR
        - Rolling MST Mean Distance exhibits a unit root (ADF > 0.05) OR
        - MST distance significantly correlates with VIX (cor <= 0.05).
        - The market covariance structure or network skeleton is non-stationary, exhibiting regime dependencies or structural drift. Static risk models will severely miscalculate risk during market environment shifts.
    

```{r}
# Ensure clean, filtered matrices and series
Sigma_Low  <- cov(low_vix_returns)
Sigma_High <- cov(high_vix_returns)
R_Low      <- cor(low_vix_returns)
R_High     <- cor(high_vix_returns)

n_low  <- nrow(low_vix_returns)
n_high <- nrow(high_vix_returns)

# Test 1: Jennrich Test for Correlation Matrix Equality
jennrich_test <- psych::cortest.jennrich(R1 = R_Low, R2 = R_High, n1 = n_low, n2 = n_high)

nce_result <- paste0(
  "--- 1. Jennrich Test for Correlation Matrix Equality ---\n",
  "Null Hypothesis (H0) : Correlation matrices are equal across regimes (R_Low = R_High)\n",
  "Chi-Square Statistic : ", round(jennrich_test$chi, 4), "\n",
  "Degrees of Freedom   : ", jennrich_test$df, "\n",
  "p-value              : ", format.pval(jennrich_test$p, digits = 4), "\n"
)

if (jennrich_test$p <= 0.05) {
  nce_result <- paste0(nce_result, "VERDICT              : REJECT H0. Correlation matrix is non-stationary and regime-dependent.\n\n")
} else {
  nce_result <- paste0(nce_result, "VERDICT              : FAIL TO REJECT H0. Correlation matrix is stationary.\n\n")
}

# Test 2: Augmented Dickey-Fuller (ADF) Test on MST Mean Distance
rolling_distances <- na.omit(network_evolution$Mean_Distance)
adf_result <- urca::ur.df(rolling_distances, type = "drift", selectlags = "AIC")
df_stat    <- adf_result@teststat[1, 1]
lag_param  <- adf_result@lags
p_val_adf  <- urca::punitroot(df_stat, trend = "c", statistic = "t")

nce_result <- paste0(
  nce_result,
  "--- 2. ADF Test on Rolling MST Mean Distance ---\n",
  "Null Hypothesis (H0) : Series has a unit root (Non-stationary)\n",
  "Dickey-Fuller Stat   : ", round(df_stat, 4), "\n",
  "Lag Parameter        : ", lag_param, "\n"
)

if (length(p_val_adf) == 0 || is.na(p_val_adf)) {
  nce_result <- paste0(
    nce_result, 
    "p-value              : NA (Inconclusive due to series properties)\n",
    "VERDICT              : INCONCLUSIVE. Could not resolve stationarity.\n\n"
  )
} else {
  nce_result <- paste0(nce_result, "p-value              : ", format.pval(p_val_adf, digits = 4), "\n")
  if (p_val_adf <= 0.05) {
    nce_result <- paste0(nce_result, "VERDICT              : REJECT H0. STATIONARY. The rolling network skeleton fluctuates around a constant mean.\n\n")
  } else {
    nce_result <- paste0(nce_result, "VERDICT              : FAIL TO REJECT H0. NON-STATIONARY. The network skeleton exhibits structural drift/memory.\n\n")
  }
}

# Test 3: Correlation Test between Rolling MST Mean Distance and VIX
cor_test <- cor.test(network_evolution$Mean_Distance, network_evolution$VIX, use = "complete.obs")
cor_stat  <- cor_test$statistic
cor_p_val <- cor_test$p.value
cor_est   <- cor_test$estimate

nce_result <- paste0(
  nce_result,
  "--- 3. Correlation Test: Rolling MST Mean Distance vs. VIX ---\n",
  "Null Hypothesis (H0) : MST Mean Distance and VIX are uncorrelated (rho = 0)\n",
  "Correlation (rho)    : ", round(cor_est, 4), "\n",
  "t-Statistic          : ", round(cor_stat, 4), "\n",
  "p-value              : ", format.pval(cor_p_val, digits = 4), "\n"
)

if (cor_p_val <= 0.05) {
  nce_result <- paste0(nce_result, "VERDICT              : REJECT H0. Dynamic coupling detected (MST Distance correlates with VIX).\n\n")
} else {
  nce_result <- paste0(nce_result, "VERDICT              : FAIL TO REJECT H0. UNCORRELATED. Network geometry is independent of market stress.\n\n")
}

cat(nce_result)
```

 3. Verdict: `TBD`
 
**CAS**: The Complex Adaptive Systems paradigm infers that because joint asset transitions are governed by regime-dependent contact geometry, a portfolio’s effective diversification benefit will systematically shrink exactly during the deep market drawdowns an investor cares about avoiding.
 

1. Gathering Evidence:
  -  Asymmetric Correlation Co-Movement Test measuring the mean difference in cross-asset correlations (Delta rho) and the distance (Frobenius Norm) between Sigma_Low and Sigma_High during distinct VIX regimes.
  - Topological Compression Test (Pearson's r): We test for endogenous network collapse by calculating the correlation between the rolling MST Mean Distance (Dt) and the market fear gauge (Vt). Under CAS, as stress rises, the network skeleton contracts:
* H0: Cor(Dt, Vt) = 0
* H1: Cor(Dt, Vt) < 0

2. Formal Hypotheses & Decision Rule
  -  Null Hypothesis (H0): Linear Network Geometry (Stable Skeleton)
    - Frobenius matrix distance is not statistically significant AND
    - Pearson's correlation shows no significant negative relationship (p >= 0.05) AND
    - MST network diameter remains stable during VIX spikes.
    - Conclusion: The asset ecosystem follows a simpler, linear geometry. Network metrics remain stable or show random, uncorrelated fluctuations during volatility spikes, proving correlation shifts are temporary statistical noise rather than systemic phase changes.
  - Alternative Hypothesis (H1): Endogenous Topological Phase Transition (Dynamic Network Collapse)
    - Frobenius matrix distance is statistically significant AND
    - Pearson's correlation shows a strong, significant negative relationship (Cor(Dt, Vt) < 0 with p < 0.05) AND
    - MST network diameter contracts during VIX spikes.
    - Conclusion: The market undergoes an endogenous topological phase transition under stress. The asset network actively collapses, causing correlation to break down and diversification to vanish at peak volatility.
    

```{r}
# Test 1: Frobenius Norm of Covariance Matrix Difference
# H0: ||Sigma_Low - Sigma_High||_F = 0 (Covariance matrix distance is zero)
frobenius_norm <- norm(Sigma_Low - Sigma_High, type = "F")

frob_out <- paste0(
  "--- 1. Frobenius Norm of Covariance Differences ---\n",
  "Null Hypothesis (H0) : Covariance matrices are identical (||Sigma_Low - Sigma_High||_F = 0)\n",
  "Frobenius Distance   : ", round(frobenius_norm, 6), "\n"
)

if (frobenius_norm > 0) {
  frob_out <- paste0(frob_out, "VERDICT              : REJECT H0. Non-zero matrix distance detected between stress regimes.\n\n")
} else {
  frob_out <- paste0(frob_out, "VERDICT              : FAIL TO REJECT H0. Covariance matrices are identical.\n\n")
}

# Test 2: Topological Network Compression Test (Pearson's r)
# H0: Cor(MST Mean Distance, VIX) >= 0 (No compression under stress)
# H1: Cor(MST Mean Distance, VIX) < 0 (Compression under stress)
pearson_test <- cor.test(
  network_evolution$Mean_Distance, 
  network_evolution$VIX, 
  alternative = "less", 
  method = "pearson"
)

frob_out <- paste0(
  frob_out,
  "--- 2. Pearson Test: Network Distance vs. Market Fear (VIX) ---\n",
  "Null Hypothesis (H0) : Network distance is not negatively correlated with VIX (rho >= 0)\n",
  "Correlation (r)      : ", round(pearson_test$estimate, 4), "\n",
  "t-Statistic          : ", round(pearson_test$statistic, 4), "\n",
  "p-value              : ", format.pval(pearson_test$p.value, digits = 4), "\n"
)

if (pearson_test$p.value <= 0.05 && pearson_test$estimate < -0.3) {
  frob_out <- paste0(frob_out, "VERDICT              : REJECT H0. Massive, statistically significant network contraction during fear spikes.\n\n")
} else if (pearson_test$p.value <= 0.05) {
  frob_out <- paste0(frob_out, "VERDICT              : REJECT H0. Statistically significant network contraction detected.\n\n")
} else {
  frob_out <- paste0(frob_out, "VERDICT              : FAIL TO REJECT H0. No significant negative relationship between network structure and VIX.\n\n")
}

# Test 3: MST Diameter Contraction Comparison
# H0: High-VIX MST Diameter >= Low-VIX MST Diameter (No diameter contraction)
dist_low  <- sqrt(2 * (1 - R_Low))
dist_high <- sqrt(2 * (1 - R_High))

g_low  <- minimum.spanning.tree(graph_from_adjacency_matrix(dist_low, mode = "undirected", weighted = TRUE, diag = FALSE))
g_high <- minimum.spanning.tree(graph_from_adjacency_matrix(dist_high, mode = "undirected", weighted = TRUE, diag = FALSE))

diameter_low  <- diameter(g_low, directed = FALSE, weights = E(g_low)$weight)
diameter_high <- diameter(g_high, directed = FALSE, weights = E(g_high)$weight)

frob_out <- paste0(
  frob_out, 
  "--- 3. MST Network Diameter Comparison ---\n",
  "Null Hypothesis (H0) : High-VIX diameter >= Low-VIX diameter (No contraction under stress)\n",
  "Low-VIX MST Diameter : ", round(diameter_low, 4), "\n",
  "High-VIX MST Diameter: ", round(diameter_high, 4), "\n"
)

if (diameter_high < diameter_low) {
  percent_contraction <- ((diameter_low - diameter_high) / diameter_low) * 100
  frob_out <- paste0(frob_out, "VERDICT              : REJECT H0. Network diameter contracted by ", round(percent_contraction, 2), "% under stress.\n\n")
} else {
  frob_out <- paste0(frob_out, "VERDICT              : FAIL TO REJECT H0. Network diameter did not contract under stress.\n\n")
}

cat(frob_out)
```


 3. Verdict: Null Hypothesis Rejected with Nuance. The traditional "Bouchaud's Correlation Breakdown" model assumes all assets collapse together during a panic. However, data reveals a far more complex adaptive phenomenon. Under stress, the network does not uniformly contract; it stretches (Diameter increases from 2.65 to 3.30). While cyclical equities and airlines group tightly together, highly speculative assets like Bitcoin decouple entirely and drift to the extreme periphery of the market manifold. Diversification does not simply disappear—it structurally morphs, leaving standard equity hedges broken while expanding the distance to alternative assets.
 

## Synthesis

By widening our lens from isolated prices to joint relationships, we can test whether cross-asset correlations behave like fixed, immutable laws (NCE) or dynamic structures that warp based on market stress (CAS). If the tests show that correlations converge toward $1$ during high-VIX environments, it proves that diversification is highly unstable. This motivates our final zoom-out: examining the underlying macroeconomic yield curve that anchors these assets.

This result also has a direct, NCE-native consequence for portfolio construction, via Grinold's (1989) Fundamental Law of Active Management [3]: expected risk-adjusted skill scales as $\text{IR} \approx \text{IC} \times \sqrt{\text{Breadth}}$, where Breadth is the number of *independent* return sources a portfolio can draw on. A four-asset book only earns the diversification credit its Fundamental Law "breadth" term promises if those four return streams are close to independent. The MST-diameter and Frobenius-distance results above show that this independence is regime-dependent, not constant: during calm markets, `UAL`, `PAL`, `BTC-USD`, and `SPY` behave close enough to four separate bets to justify treating Breadth $\approx 4$; during high-VIX regimes, the network stretches and partially decouples rather than uniformly collapsing (per the Verdict above), so the *effective* breadth an allocator can rely on shifts asset-by-asset instead of shrinking uniformly to $1$. In other words, the Fundamental Law's Breadth term is not a fixed input in this dataset - it is itself a random variable driven by the same VIX-conditioned network dynamics tested here, which is exactly the quantity Section 07's allocation should condition on rather than assume.


## Footnote:

[b2] https://arxiv.org/abs/1009.2928