
Here, we view how the relationships of the assets moves in the economies they are in.

## Structure

### Correlation and Covariance Matrix
 
 Here, we utilize Markowitz’s Modern Portfolio Theory (MPT) and the foundations of Sharpe’s Capital Asset Pricing Model (CAPM). Risk is fully quantified by the variance-covariance matrix ($\Sigma$), assuming structural errors or shifts are merely random, zero-mean Gaussian noise around a permanent equilibrium.
 
 1. **Raw Input**: Co-moving daily log returns matrix for UAL, PAL, BTC-USD, and SPY.
 2. **Enriched State**: A static sample covariance matrix ($\Sigma$) calculated across the entire dataset.
 3. **Compression**: A single cross-asset correlation scalar ($\rho_{i,j}$) per asset pair.
 
```{r}
# --- 1. Standard Data Extraction ---
returns_only <- master_df %>% 
  select(
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj
  ) %>% 
  drop_na()

# --- 2. Calculate Standard Matrices ---
# Just pure, unadulterated base R matrices
cov_matrix <- cov(returns_only)
cor_matrix <- cor(returns_only)

# --- 3. Print Standard Text Outputs ---
# These will render as regular, readable code blocks in your Rmd
cat("### Covariance Matrix (Sigma)\n")
print(round(cov_matrix, 6))

cat("\n### Correlation Matrix (Rho)\n")
print(round(cor_matrix, 6))

ggcorrplot(
  cor_matrix, 
  hc.order = FALSE,      # No fancy clustering/reordering
  type = "full",         # Full square matrix so the axes line up perfectly
  lab = TRUE,            # Display the correlation coefficients
  lab_size = 4,
  colors = c("#20639B", "white", "#D14949"),
  title = "Standard Static Correlation Matrix"
)
```

**CAS**: 
 
This builds on Bouchaud’s Law of Endogenous Correlation Breakdown[b1], which proves that cross-asset dependencies are non-stationary, state-dependent, and governed by systemic feedback loops. Rather than viewing correlation shifts as statistical noise, the system recognizes them as distinct geometric structures. High-volatility matrices represent entirely separate topological charts on the market manifold that cannot be smoothly mapped using a single, global Euclidean average.
 
 1. Raw Input: Daily log returns matrix for UAL, PAL, BTC-USD, SPY, and the daily conditioning regime marker VIX.
 2. Enriched State: Conditional, rolling covariance matrices $\Sigma_t \vert{} \text{VIX}_t$, explicitly split into separate geometric charts.
 3. Compression: Partitioned states mapped into distinct manifolds: Low-VIX Chart ($\text{VIX} \le 25\text{th percentile}$) and High-VIX Chart ($\text{VIX} \ge 75\text{th percentile}$).


CAS's state is that same matrix conditioned on regime — a High-VIX matrix and a Low-VIX matrix, treated as genuinely different objects, not noise around one "true" matrix.

```{r}
# 1. Prepare data with VIX as our dynamic conditioning regime marker
cas_prepared <- master_df %>%
  select(
    Date,
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj,
    VIX = VIX.Adjusted # Assuming VIX.Close is in your master_df
  ) %>%
  drop_na()

# 2. Compute the conditioning thresholds (25th and 75th percentiles of VIX)
vix_25 <- quantile(cas_prepared$VIX, 0.25, na.rm = TRUE)
vix_75 <- quantile(cas_prepared$VIX, 0.75, na.rm = TRUE)

# 3. Partition the dataset into separate topological charts
low_vix_returns <- cas_prepared %>% filter(VIX <= vix_25) %>% select(UAL, PAL, BTC, SPY)
high_vix_returns <- cas_prepared %>% filter(VIX >= vix_75) %>% select(UAL, PAL, BTC, SPY)

# 4. Calculate the separate, conditional correlation objects
cor_low_vix  <- cor(low_vix_returns)
cor_high_vix <- cor(high_vix_returns)

# 5. Plot them side-by-side to visualize the "Endogenous Correlation Breakdown"
library(gridExtra)

plot_low <- ggcorrplot(
  cor_low_vix, 
  type = "full", 
  lab = TRUE, 
  title = paste("Low-VIX Chart (VIX <=", round(vix_25, 1), ")"),
  colors = c("#20639B", "white", "#D14949")
)

plot_high <- ggcorrplot(
  cor_high_vix, 
  type = "full", 
  lab = TRUE, 
  title = paste("High-VIX Chart (VIX >=", round(vix_75, 1), ")"),
  colors = c("#20639B", "white", "#D14949")
)

# Render side-by-side comparison
grid.arrange(plot_low, plot_high, ncol = 2)
```


## Dynamics

**NCE**: 

Assumes correlations are stationary and static — the relationship between UAL and SPY today is the same relationship in a crash.  Diversification is permanent.

 1. Transitions (Conservative): Portfolio variance transitions over time are deterministic.
 2. Grouped Context of Transitions: Multiple assets' cross-dynamics are modeled as permanently coupled geometric Brownian motions. Co-movements scale symmetrically across all economic cycles, treating structural changes as impossible.

**CAS**: Correlations are regime-dependent and tend to converge upward precisely during stress where diversification "disappears" when you need it most.
 
 1. Transitions (Diffusion): Systemic risk behaves like a thermodynamic fluid following a Non-equilibrium Dissipation. In calm regimes or low-VIX, the flow is Symplectic (information-conserving, asset pairs preserve their unique orthogonal vectors). When a threshold is crossed (e.g., a high-VIX shock), margin calls and liquidations force a massive energy transfer (entropy bleed).
 2. Grouped Context of Transitions: Asset transitions are grouped into distinct market volatility regimes. Under high-volatility environments (conditioned by a high `VIX` threshold), individual asset behaviors lose their unique degrees of freedom, and correlations across completely different asset classes (e.g., `UAL`, `PAL`, and `BTC-USD`) shift simultaneously. 


```{r}
# --- HELPER FUNCTION: CORRELATION TO MST ---
build_mst_df <- function(cor_matrix, layout_seed = 42) {
  # 1. Convert correlation to distance: d = sqrt(2 * (1 - rho))
  dist_matrix <- sqrt(2 * (1 - cor_matrix))
  
  # 2. Create a fully connected graph from the distance matrix
  g_full <- graph_from_adjacency_matrix(
    dist_matrix, 
    mode = "undirected", 
    weighted = TRUE, 
    diag = FALSE
  )
  
  # 3. Extract the Minimum Spanning Tree (MST) using Kruskal's algorithm
  g_mst <- minimum.spanning.tree(g_full, weights = E(g_full)$weight)
  
  # 4. Format into a clean dataframe for ggplot/ggnetwork visualization
  set.seed(layout_seed)
  mst_net <- ggnetwork(g_mst, layout = layout_with_kk(g_mst))
  return(mst_net)
}

# --- GENERATE REGIME-SPECIFIC MSTs ---
mst_low  <- build_mst_df(cor_low_vix)
mst_high <- build_mst_df(cor_high_vix)

# --- PLOT THE MSTs ---
plot_mst <- function(mst_data, title_text) {
  ggplot(mst_data, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges(color = "grey50", alpha = 0.7, size = 1) +
    # Edge labels show the physical "distance" (closer to 0 means higher correlation)
    geom_edgetext(aes(label = round(weight, 2)), color = "darkblue", fill = "white", size = 3) +
    geom_nodes(color = "firebrick", size = 8) +
    geom_nodetext(aes(label = name), color = "white", fontface = "bold", size = 3) +
    theme_blank() +
    labs(title = title_text)
}

grid.arrange(
  plot_mst(mst_low, "Low-VIX MST (Tranquil System Skeleton)"),
  plot_mst(mst_high, "High-VIX MST (Stressed System Skeleton)"),
  ncol = 2
)
```

```{r}
# --- 1. SETUP PARAMETERS ---
roll_window <- 21
dates <- unique(cas_prepared$Date)[roll_window:nrow(cas_prepared)]

# Initialize empty data frame to collect metrics
network_metrics <- data.frame(
  Date = as.Date(character()), 
  Mean_Distance = numeric(), 
  Diameter = numeric()
)

# 2. RUN ROLLING NETWORK FILTER WITH NA SAFETY ---
for (i in roll_window:nrow(cas_prepared)) {
  # Extract window returns
  window_data <- cas_prepared[(i - roll_window + 1):i, c("UAL", "PAL", "BTC", "SPY")]
  window_date <- cas_prepared$Date[i]
  # Remove any rows within the rolling window that contain NAs
  window_data_clean <- na.omit(window_data)
  
  # Ensure we still have enough data points to compute a valid correlation matrix
  # And ensure none of the columns have zero standard deviation (which causes NaN correlations)
  has_variance <- all(sapply(window_data_clean, sd) > 0)
  
  if (nrow(window_data_clean) >= 10 && has_variance) {
    # Compute correlation
    r_matrix <- cor(window_data_clean)
    
    # Map correlation to distance: d = sqrt(2 * (1 - r))
    d_matrix <- sqrt(2 * (1 - r_matrix))
    
    # Final sanity check: If any cell in our distance matrix is still NA/NaN, replace it with 2 (max distance)
    d_matrix[is.na(d_matrix)] <- 2
    
    # Build the MST
    g_full <- graph_from_adjacency_matrix(d_matrix, mode = "undirected", weighted = TRUE, diag = FALSE)
    g_mst  <- minimum.spanning.tree(g_full, weights = E(g_full)$weight)
    
    # Calculate topology metrics
    mean_dist <- mean(E(g_mst)$weight) # Average closeness of the system
    diameter_val <- diameter(g_mst, directed = FALSE, weights = E(g_mst)$weight) # Furthest distance
    
    network_metrics <- rbind(network_metrics, data.frame(
      Date = window_date,
      Mean_Distance = mean_dist,
      Diameter = diameter_val
    ))
  }
}

network_evolution <- network_metrics %>%
  left_join(select(cas_prepared, Date, VIX), by = "Date")

ggplot(network_evolution, aes(x = Date)) +
  geom_line(aes(y = Mean_Distance, color = "Network Mean Distance (Compression)"), size = 1) +
  geom_line(aes(y = VIX / 100, color = "VIX (Fear Index, Scaled)"), size = 1) +
  scale_y_continuous(
    name = "Network Mean Distance",
    sec.axis = sec_axis(~ . * 100, name = "VIX Level")
  ) +
  scale_color_manual(values = c("forestgreen", "darkred")) +
  theme_minimal() +
  labs(
    title = "Systemic Phase Transitions: Rolling Network Compression vs. Market Fear",
    subtitle = "Safe from NA errors",
    x = "Date",
    color = "Metric"
  ) +
  theme(legend.position = "bottom")
```

## Inference

**NCE**: The Neoclassical paradigm infers that because asset relationships are governed by a stationary covariance matrix, the physical structure of the market network is constant. A diversified portfolio's risk reduction remains stable and can be reliably trusted for long-term strategic asset allocation and Value-at-Risk (VaR) planning.
 
 1. Gathering Evidence: We test the null hypothesis that the covariance structures of the low-stress and high-stress regimes are identical:
    - $$H_0: \mathbf{\Sigma}_{\text{Low-VIX}} = \mathbf{\Sigma}_{\text{High-VIX}}$$
    - If the underlying covariance is stationary, the rolling MST Mean Distance ($D_t$) must behave as a stationary white noise process fluctuating tightly around a constant, time-invariant mean: $$H_0: D_t = \bar{D} + \epsilon_t \quad \text{where} \quad \epsilon_t \sim N(0, \sigma^2)$$. We test this via an Augmented Dickey-Fuller (ADF) test on the rolling MST distance series.
 2. Decision Rule:
    - **IF** the Jennrich test fails to reject covariance homogeneity ($p > 0.05$) AND the rolling MST Mean Distance series is stationary and completely uncorrelated with the VIX ($p_{\text{cor}} > 0.05$):
    - Then NCE is verified. The market skeleton is rigid and static. A single historical covariance matrix is sufficient to build permanently diversified portfolios.
    - **IF** the Jennrich test rejects homogeneity ($p \le 0.05$) OR the rolling MST metrics show structural breaks and systematic drift:
    - Then the NCE assumption of a stable portfolio shield is broken. Covariance is non-stationary, meaning static risk models will severely miscalculate risk during environmental shifts.
    

```{r}
# Ensure we have clean, filtered matrices and series
Sigma_Low  <- cov(low_vix_returns)
Sigma_High <- cov(high_vix_returns)
R_Low      <- cor(low_vix_returns)
R_High     <- cor(high_vix_returns)

n_low  <- nrow(low_vix_returns)
n_high <- nrow(high_vix_returns)
# Test 1: Jennrich Test for Correlation Matrix Equality
# H0: R_Low = R_High (Correlation structures are identical across regimes)
jennrich_test <- psych::cortest.jennrich(R1 = R_Low, R2 = R_High, n1 = n_low, n2 = n_high)

cat("--- 1. Jennrich Test for Correlation Matrix Equality ---\n")
cat("Chi-Square Statistic :", round(jennrich_test$chi, 4), "\n")
cat("Degrees of Freedom   :", jennrich_test$df, "\n")
cat("p-value              :", format.pval(jennrich_test$p, digits = 4), "\n")

if (jennrich_test$p <= 0.05) {
  cat("VERDICT: REJECT H0. The correlation matrix is non-stationary and regime-dependent.\n\n")
} else {
  cat("VERDICT: FAIL TO REJECT H0. The correlation matrix is stationary.\n\n")
}

# Test 2: Augmented Dickey-Fuller (ADF) Test on MST Mean Distance
# H0: The rolling MST Mean Distance has a unit root (is non-stationary)
# we clean NA values for the time series test
rolling_distances <- na.omit(network_evolution$Mean_Distance)
adf_result <- tseries::adf.test(rolling_distances, alternative = "stationary")

cat("--- 2. ADF Test on Rolling MST Mean Distance ---\n")
cat("Dickey-Fuller Stat   :", round(adf_result$statistic, 4), "\n")
cat("Lag Parameter        :", adf_result$parameter, "\n")

# Safe check for NA or empty p-value
p_val_adf <- adf_result$p.value # Note: tseries uses p.value, not p

if (length(p_val_adf) == 0 || is.na(p_val_adf)) {
  cat("p-value              : NA (Inconclusive due to series properties)\n")
  cat("VERDICT: INCONCLUSIVE. The test could not resolve stationarity.\n\n")
} else {
  cat("p-value              :", format.pval(p_val_adf, digits = 4), "\n")
  if (p_val_adf <= 0.05) {
    cat("VERDICT: STATIONARY. The rolling network skeleton fluctuates around a constant mean.\n\n")
  } else {
    cat("VERDICT: NON-STATIONARY. The network skeleton exhibits structural drift/memory.\n\n")
  }
}
```

 3. Verdict: `TBD`
 
**CAS**: The Complex Adaptive Systems paradigm infers that because joint asset transitions are governed by regime-dependent contact geometry, a portfolio’s effective diversification benefit will systematically shrink exactly during the deep market drawdowns an investor cares about avoiding.
 
 1. Gathering Evidence: 
    - Asymmetric Correlation Co-Movement Test measuring the mean difference in cross-asset correlations ($\Delta \rho_{i,j}$) and the distance (Frobenius Norm) between $\Sigma_{\text{Low}}$ and $\Sigma_{\text{High}}$ during distinct `VIX` regimes.
    - Topological Compression Test (Pearson's $r$): We test for endogenous network collapse by calculating the correlation between the rolling MST Mean Distance ($D_t$) and the market fear gauge ($V_t$). Under CAS, as stress rises, the network skeleton contracts:
    - $$H_0: \text{Cor}(D_t, V_t) = 0$$
    - $$H_1: \text{Cor}(D_t, V_t) < 0$$
 2. Decision Rule:
    - **IF** the Frobenius matrix distance is statistically significant, AND Pearson’s correlation shows a strong, significant negative relationship ($\text{Cor}(D_t, V_t) \ll 0$ with $p < 0.05$), AND the MST network diameter contracts during VIX spikes:
      - The market undergoes an endogenous topological phase transition under stress. The asset network actively collapses, causing correlation to break down and diversification to vanish at peak volatility.
    - **IF** the network metrics remain stable or show random, uncorrelated fluctuations during volatility spikes:
      - Then the asset ecosystem follows a simpler, linear geometry. Correlation shifts are temporary statistical noise rather than systemic phase changes.
      

```{r}
# Test 1: Frobenius Norm of Covariance Matrix Difference
# Measures the absolute distance between the two regime charts in state-space
frobenius_norm <- norm(Sigma_Low - Sigma_High, type = "F")

cat("--- 1. Frobenius Norm of Covariance Differences ---\n")
cat("Frobenius Distance (||Sigma_Low - Sigma_High||_F):", round(frobenius_norm, 6), "\n\n")

# Test 2: Topological Network Compression Test (Pearson's r)
# H0: Correlation(MST Mean Distance, VIX) >= 0 (No compression under stress)
# H1: Correlation(MST Mean Distance, VIX) < 0 (Compression under stress)
pearson_test <- cor.test(
  network_evolution$Mean_Distance, 
  network_evolution$VIX, 
  alternative = "less", 
  method = "pearson"
)

cat("--- 2. Pearson Test: Network Distance vs. Market Fear (VIX) ---\n")
cat("Correlation (r)      :", round(pearson_test$estimate, 4), "\n")
cat("t-statistic          :", round(pearson_test$statistic, 4), "\n")
cat("p-value              :", format.pval(pearson_test$p, digits = 4), "\n")

if (pearson_test$p <= 0.05 && pearson_test$estimate < -0.3) {
  cat("VERDICT: VALIDATED. Massive, statistically significant network contraction during fear spikes.\n\n")
} else {
  cat("VERDICT: REJECTED. No meaningful relationship between network structure and market fear.\n\n")
}

# Test 3: MST Diameter Contraction Comparison
# Extract the underlying MST objects for low and high VIX regimes to calculate diameter
dist_low  <- sqrt(2 * (1 - R_Low))
dist_high <- sqrt(2 * (1 - R_High))

g_low  <- minimum.spanning.tree(graph_from_adjacency_matrix(dist_low, mode = "undirected", weighted = TRUE, diag = FALSE))
g_high <- minimum.spanning.tree(graph_from_adjacency_matrix(dist_high, mode = "undirected", weighted = TRUE, diag = FALSE))

diameter_low  <- diameter(g_low, directed = FALSE, weights = E(g_low)$weight)
diameter_high <- diameter(g_high, directed = FALSE, weights = E(g_high)$weight)

cat("--- 3. MST Network Diameter Comparison ---\n")
cat("Low-VIX MST Diameter :", round(diameter_low, 4), "\n")
cat("High-VIX MST Diameter:", round(diameter_high, 4), "\n")

if (diameter_high < diameter_low) {
  percent_contraction <- ((diameter_low - diameter_high) / diameter_low) * 100
  cat("VERDICT: VALIDATED. Network diameter contracted by", round(percent_contraction, 2), "% under stress.\n\n")
} else {
  cat("VERDICT: REJECTED. Network diameter did not contract under stress.\n\n")
}
```


 3. Verdict: Null Hypothesis Rejected with Nuance. The traditional "Bouchaud's Correlation Breakdown" model assumes all assets collapse together during a panic. However, data reveals a far more complex adaptive phenomenon. Under stress, the network does not uniformly contract; it stretches (Diameter increases from 2.65 to 3.30). While cyclical equities and airlines group tightly together, highly speculative assets like Bitcoin decouple entirely and drift to the extreme periphery of the market manifold. Diversification does not simply disappear—it structurally morphs, leaving standard equity hedges broken while expanding the distance to alternative assets.
 

## Synthesis

By widening our lens from isolated prices to joint relationships, we can test whether cross-asset correlations behave like fixed, immutable laws (NCE) or dynamic structures that warp based on market stress (CAS). If the tests show that correlations converge toward $1$ during high-VIX environments, it proves that diversification is highly unstable. This motivates our final zoom-out: examining the underlying macroeconomic yield curve that anchors these assets.

This result also has a direct, NCE-native consequence for portfolio construction, via Grinold's (1989) Fundamental Law of Active Management [3]: expected risk-adjusted skill scales as $\text{IR} \approx \text{IC} \times \sqrt{\text{Breadth}}$, where Breadth is the number of *independent* return sources a portfolio can draw on. A four-asset book only earns the diversification credit its Fundamental Law "breadth" term promises if those four return streams are close to independent. The MST-diameter and Frobenius-distance results above show that this independence is regime-dependent, not constant: during calm markets, `UAL`, `PAL`, `BTC-USD`, and `SPY` behave close enough to four separate bets to justify treating Breadth $\approx 4$; during high-VIX regimes, the network stretches and partially decouples rather than uniformly collapsing (per the Verdict above), so the *effective* breadth an allocator can rely on shifts asset-by-asset instead of shrinking uniformly to $1$. In other words, the Fundamental Law's Breadth term is not a fixed input in this dataset - it is itself a random variable driven by the same VIX-conditioned network dynamics tested here, which is exactly the quantity Section 07's allocation should condition on rather than assume.


## Footnote:

[b2] https://arxiv.org/abs/1009.2928