
Here, we view the entire context of the market and economy.

## Structure

**NCE**: via CAPM, collapses the entire term structure into one risk-free rate — a single number standing in for "the cost of money."
 
 1. Raw Input: Daily DGS10, DGS2, or monthly FEDFUNDS.
 2. Enriched State: A single risk-free scalar ($r_f$) selected based on the closest maturity to the asset's cash flows (e.g., using DGS2 for a 2-year horizon contract).
 3. Compression: The chosen flat scalar applied as a uniform baseline denominator in linear valuation equations.

```{r}
# --- 1. DYNAMIC PARAMETER ESTIMATION FROM DATA ---
# Extract daily log returns from your master dataframe
daily_returns <- master_df %>%
  select(
    Date,
    SPY = SPY.LogAdj,
    PAL = PAL.LogAdj,
    UAL = UAL.LogAdj,
    BTC = `BTC-USD.LogAdj`
  ) %>%
  drop_na() %>%
  select(-Date)

# Define trading frequency scaling factor (Annualized)
scale_factor <- 252

# A. Calculate Dynamic Expected Annualized Returns (mu)
# Using mean daily log returns annualized
mu <- colMeans(daily_returns) * scale_factor

# B. Calculate Dynamic Annualized Volatilities (vol)
vol <- apply(daily_returns, 2, sd) * sqrt(scale_factor)

# C. Calculate Dynamic Correlation Matrix
corr_matrix <- cor(daily_returns)

# D. Calculate Dynamic Covariance Matrix (Sigma)
# Sigma = D %*% R %*% D where D is diagonal of standard deviations
D <- diag(vol)
Sigma <- D %*% corr_matrix %*% D

# --- 2. SOLVE ANALYTIC EFFICIENT FRONTIER WITH EMPIRICAL DATA ---
rf <- 0.045  # Standard policy/risk-free rate (4.5%)
inv_Sigma <- solve(Sigma)

ones <- rep(1, length(mu))
A <- as.numeric(t(ones) %*% inv_Sigma %*% ones)
B <- as.numeric(t(ones) %*% inv_Sigma %*% mu)
C <- as.numeric(t(mu) %*% inv_Sigma %*% mu)
D_val <- A * C - B^2

# Generate empirical frontier curve points
target_returns <- seq(min(mu) - 0.05, max(mu) + 0.10, length.out = 500)
frontier_points <- data.frame(Return = target_returns) %>%
  mutate(
    Variance = (A * Return^2 - 2 * B * Return + C) / D_val,
    SD = sqrt(Variance)
  )

# --- 3. DISSOLVE TANGENCY PORTFOLIO (M) ---
excess_return <- mu - rf * ones
w_tangency <- inv_Sigma %*% excess_return
w_tangency <- w_tangency / sum(w_tangency)

ret_m <- as.numeric(t(w_tangency) %*% mu)
sd_m <- as.numeric(sqrt(t(w_tangency) %*% Sigma %*% w_tangency))
sharpe_m <- (ret_m - rf) / sd_m

# --- 4. RISK DECOMPOSITION FOR UAL ---
ual_ret <- mu["UAL"]
ual_sd  <- vol["UAL"]
sd_a <- (ual_ret - rf) / sharpe_m

# --- 5. RENDER THE DYNAMIC PLOT ---
ggplot() +
  geom_path(data = frontier_points, aes(x = SD, y = Return), 
            color = "black", size = 1.1) +
  geom_abline(intercept = rf, slope = sharpe_m, 
              color = "blue", size = 1, linetype = "solid") +
  
  # Horizontal drop lines
  annotate("segment", x = 0, xend = ual_sd, y = ual_ret, yend = ual_ret, 
           color = "grey50", linetype = "dashed") +
  annotate("segment", x = sd_a, xend = sd_a, y = 0, yend = ual_ret, 
           color = "grey50", linetype = "dotted") +
  annotate("segment", x = ual_sd, xend = ual_sd, y = 0, yend = ual_ret, 
           color = "grey50", linetype = "dotted") +
  
  # Risk Brackets at the bottom of the plot
  annotate("segment", x = 0.01, xend = sd_a - 0.01, y = 0.01, yend = 0.01, 
           arrow = arrow(ends = "both", angle = 90, length = unit(.1, "cm")), color = "#D14949", size = 0.8) +
  annotate("text", x = sd_a / 2, y = 0.02, label = "Systematic\nRisk", 
           color = "#D14949", size = 3, fontface = "bold") +
  
  annotate("segment", x = sd_a + 0.01, xend = ual_sd - 0.01, y = 0.01, yend = 0.01, 
           arrow = arrow(ends = "both", angle = 90, length = unit(.1, "cm")), color = "#20639B", size = 0.8) +
  annotate("text", x = (sd_a + ual_sd) / 2, y = 0.02, label = "Diversifiable\n(Idiosyncratic) Risk", 
           color = "#20639B", size = 3, fontface = "bold") +
  
  # Reference Nodes
  annotate("point", x = 0, y = rf, color = "black", size = 4, shape = 18) +
  annotate("text", x = 0.01, y = rf + 0.005, label = "r[f]", parse = TRUE, fontface = "bold", hjust = 0) +
  
  annotate("point", x = sd_m, y = ret_m, color = "darkgreen", size = 4) +
  annotate("text", x = sd_m - 0.015, y = ret_m + 0.01, label = "M (Tangency Portfolio)", 
           color = "darkgreen", fontface = "bold", hjust = 1) +
  
  annotate("point", x = sd_a, y = ual_ret, color = "black", size = 3) +
  annotate("text", x = sd_a + 0.01, y = ual_ret - 0.01, label = "A", fontface = "bold") +
  
  # Asset labels (Dynamically positioned at their real calculated risk/return points)
  geom_point(data = data.frame(SD = vol, Return = mu, Asset = names(mu)), 
             aes(x = SD, y = Return, color = Asset), size = 4) +
  geom_text(data = data.frame(SD = vol, Return = mu, Asset = names(mu)), 
            aes(x = SD, y = Return, label = Asset), vjust = -1.2, fontface = "bold", size = 3.5) +
  
  scale_color_manual(values = c("SPY" = "#3CAEA3", "PAL" = "#F6D55C", "UAL" = "#20639B", "BTC" = "#D14949")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_cartesian(xlim = c(0, max(vol) * 2.5), ylim = c(min(mu), max(mu) * 2.5)) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Capital Allocation Line & Empirical Efficient Frontier",
    subtitle = "Calculated Dynamically from Real-World Asset Return Matrix",
    x = "Portfolio Total Risk (Standard Deviation)",
    y = "Expected Return (E[R])"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```


**NCE (extended — factor decomposition):** The single-scalar CAPM view above is the *degenerate* case of a richer, but still classical, equilibrium-econometrics model: the Litterman & Scheinkman (1991) three-factor decomposition of the yield curve [4]. This is a linear, orthogonal (PCA-derived) re-coordinatization of the same term structure - it relaxes "one rate" to "three factors," but it is still a static, reversible, equilibrium description with no feedback, heterogeneous agents, or path-dependent mechanism attached. That is why it belongs on the NCE side of this report rather than the CAS side: richer linear algebra is not, by itself, evidence of a complex adaptive system. (See the Footnote at the end of this section for what a genuinely CAS-consistent account of the yield curve would need to look like instead.)

 1. Raw Input: Co-moving rates across the term spectrum (FEDFUNDS, DGS2, and DGS10) matched with macroeconomic inflation context (CPIAUCSL).
 2. Enriched State: A 3-dimensional coordinate state vector mapping the classical Litterman-Scheinkman yield curve factors directly:
   - $\text{Level}_t = \frac{\text{FEDFUNDS}_t + \text{DGS2}_t + \text{DGS10}_t}{3}$ (Overall altitude of interest rates)
   - $\text{Slope}_t = \text{DGS10}_t - \text{FEDFUNDS}_t$ (The long-to-short macro spread)
   - $\text{Curvature}_t = (\text{DGS10}_t - \text{DGS2}_t) - (\text{DGS2}_t - \text{FEDFUNDS}_t) = \text{DGS10}_t - 2\cdot\text{DGS2}_t + \text{FEDFUNDS}_t$ (The structural bend/twist of the medium-term belly relative to the anchors)
 3. Compression: Projecting this 3D coordinate $[L_t, S_t, C_t]$ onto distinct, discrete macro-regime charts (e.g., Bear Flattener, Bull Steeper, or Inverted Belly).

This directly invokes Litterman & Scheinkman's (1991) three-factor principal-component result, which finds that over 95% of term-structure variation is captured by these three orthogonal geometric transformations (Level, Slope, and Curvature) [4].

```{r}
# --- 1. DATA ENRICHMENT ---
macro_enriched <- master_df %>%
  select(Date, FEDFUNDS, DGS2, DGS10, CPIAUCSL) %>%
  # Carry forward monthly CPI to daily observations if dataframe is daily
  mutate(CPIAUCSL = zoo::na.locf(CPIAUCSL, na.rm = FALSE)) %>%
  drop_na() %>%
  mutate(
    
    # Calculate Nominal Factors
    Level = (FEDFUNDS + DGS2 + DGS10) / 3,
    Slope = DGS10 - FEDFUNDS,
    Curvature = DGS10 - (2 * DGS2) + FEDFUNDS,
  ) %>%
  drop_na()

# --- 2. PANEL A: RAW TERM STRUCTURE ---
raw_long <- macro_enriched %>%
  select(Date, FEDFUNDS, DGS2, DGS10) %>%
  pivot_longer(-Date, names_to = "Maturity", values_to = "Rate")

p_raw <- ggplot(raw_long, aes(x = Date, y = Rate, color = Maturity)) +
  geom_line(size = 0.8, alpha = 0.8) +
  scale_color_manual(values = c("FEDFUNDS" = "#20639B", "DGS2" = "#3CAEA3", "DGS10" = "#F6D55C")) +
  theme_minimal(base_size = 11) +
  labs(
    title = "Raw Term Structure Deformations",
    y = "Interest Rate (%)",
    x = NULL
  ) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

# --- 3. PANEL B: LITTERMAN-SCHEINKMAN FACTORS & INFLATION STATE ---
# Expanded selection to include both real level transformations and the inflation rate
factors_long <- macro_enriched %>%
  select(Date, Level, Slope, Curvature) %>%
  pivot_longer(-Date, names_to = "Coordinate", values_to = "Value")

# Enforce a logical ordering for the facet layout
factors_long$Coordinate <- factor(
  factors_long$Coordinate, 
  levels = c("Level", "Slope", "Curvature")
)

p_factors <- ggplot(factors_long, aes(x = Date, y = Value, color = Coordinate)) +
  geom_line(size = 0.8, alpha = 0.8) +
  scale_color_manual(values = c(
    "Level"         = "#D14949", # Nominal baseline
    "Slope"         = "#3F51B5", # Spread
    "Curvature"     = "#4CAF50"  # Twist

  )) +
  facet_wrap(~Coordinate, scales = "free_y", ncol = 1) +
  theme_minimal(base_size = 11) +
  labs(
    title = "Litterman-Scheinkman components",
    subtitle = "",
    y = "Coordinate Value / Rate (%)",
    x = "Date"
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold"), 
    strip.text = element_text(face = "bold")
  )

# --- 4. RENDER STRUCTURE PLOT ---
grid.arrange(p_raw, p_factors, ncol = 2)
```

## Dynamics

**NCE**: Assumes shifts in "the" rate are parallel and exogenous — everything moves together. 
 
 1. Transition (Conservative): The curve transitions uniformly. Short-rate models treat the entire curve as a single string pulled by a central mean-reverting anchor.
 2. Grouped Context of Transitions: Macroeconomic adjustments are uniform. A shift in central bank policy alters the short-term rate, and that change transfers linearly up the maturity spectrum, maintaining rigid, parallel tracking across `FEDFUNDS`, `DGS2`, and `DGS10`.

**NCE (extended — factor decomposition):** Allows the curve to twist (short end rising while the long end holds, or vice versa), which the single-scalar CAPM model above can't represent, but which is still just a richer *linear* decomposition of the same equilibrium object.
 
 1. Transition (Conservative, higher-dimensional): The curve moves via non-parallel but still linear factor rotation. For example, during an aggressive central bank tightening cycle, FEDFUNDS shoots up, DGS2 spikes violently due to medium-term policy stress, while DGS10 lags behind due to long-term economic deceleration.
 2. Grouped Transitions: Systemic rate movements are grouped by their geometric deformations (Litterman & Scheinkman, 1991 [4]). Movements are re-expressed along the curve's Level, Slope, and Curvature axes and act as macroeconomic filters that alter the valuation dynamics of specific assets (e.g., separating capital between pure liquidity proxies like `BTC-USD` and debt-laden operational equities like `UAL`) - but the *mechanism* generating those movements here is still an orthogonal factor rotation, not agent feedback or path dependence.


```{r}
# --- DATA PREPARATION & MACRO FACTOR DECOMPOSITION ---
# Extract assets, interest rates, and inflation context into a unified matrix
macro_enriched <- master_df %>%
  select(
    Date,
    FEDFUNDS, 
    DGS2, 
    DGS10,
    CPIAUCSL,
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj
  ) %>%
  # Carry forward monthly CPI to daily observations to prevent data gaps
  mutate(CPIAUCSL = zoo::na.locf(CPIAUCSL, na.rm = FALSE)) %>%
  drop_na() %>%
  mutate(
    # Calculate Year-over-Year Inflation (Assumes monthly input. If daily, change lag to 252)
    Inflation_YoY = (CPIAUCSL / lag(CPIAUCSL, 12) - 1) * 100,
    
    # Calculate Nominal Litterman-Scheinkman Factors
    Level = (FEDFUNDS + DGS2 + DGS10) / 3,
    Slope = DGS10 - FEDFUNDS,
    Curvature = DGS10 - (2 * DGS2) + FEDFUNDS,
    
    # Calculate Real Level (Deflated Baseline)
    Real_Level = Level - Inflation_YoY
  ) %>%
  drop_na()

# --- COINTEGRATION TEST (JOHANSEN) ---
# Verify if nominal yield maturities share a locked, stationary equilibrium
rate_matrix <- macro_enriched %>% select(FEDFUNDS, DGS2, DGS10)
johansen_test <- ca.jo(rate_matrix, type = "trace", ecdet = "none", K = 2)

# --- MULTI-FACTOR REGRESSION SENSITIVITY ---
# Analyze asymmetric asset elasticities against nominal factors and inflation shocks
assets <- c("UAL", "PAL", "BTC", "SPY")
expanded_sensitivity <- data.frame()

for (asset in assets) {
  formula_str <- paste(asset, "~ Level + Slope + Curvature + Inflation_YoY")
  fit <- lm(as.formula(formula_str), data = macro_enriched)
  
  fit_summary <- tidy(fit, conf.int = TRUE) %>%
    mutate(Asset = asset) %>%
    filter(term != "(Intercept)")
  
  expanded_sensitivity <- rbind(expanded_sensitivity, fit_summary)
}

jo_txt   <- paste(capture.output(summary(johansen_test)), collapse = "\n")
sens_txt <- paste(capture.output(print(expanded_sensitivity %>% select(Asset, term, estimate, statistic, p.value))), collapse = "\n")

formatted_output <- sprintf("
=========================================
        COINTEGRATION (PARALLELISM)    
=========================================
%s

=========================================
     ASYMMETRIC SENSITIVITIES (LS + CPI)  
=========================================
%s", 
jo_txt, 
sens_txt
)

cat(formatted_output)

```

```{r}
# Merge VIX back in to color the trajectory by market stress/regime
macro_dynamics <- macro_enriched %>%
  left_join(select(master_df, Date, VIX = VIX.Adjusted), by = "Date") %>%
  drop_na()

# Plot the state-space trajectory
ggplot(macro_dynamics, aes(x = Level, y = Slope)) +
  # Draw the sequential historical path
  geom_path(aes(color = VIX), size = 0.6, alpha = 0.6) +
  # Highlight the starting and ending points of the macro-regime
  geom_point(data = filter(macro_dynamics, Date == min(Date)), color = "blue", size = 4, shape = 17) +
  geom_point(data = filter(macro_dynamics, Date == max(Date)), color = "red", size = 4, shape = 15) +
  # Color scale mapping market fear/regime
  scale_color_gradientn(
    colors = c("#20639B", "#F6D55C", "#D14949"), 
    name = "VIX (Systemic Stress)"
  ) +
  annotate("text", x = min(macro_dynamics$Level), y = max(macro_dynamics$Slope), 
           label = "▲ Blue Triangle: Path Start\n■ Red Square: Path End", 
           hjust = 0, vjust = 1, size = 3, fontface = "italic") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Yield Curve Trajectory in Macro Phase Space",
    subtitle = "Systemic path-dependent evolution: Level (Altitude) vs. Slope (Spread)",
    x = "System Level (Average Interest Rate)",
    y = "System Slope (DGS10 - FEDFUNDS)"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

## Inference

**NCE**: Checks if a single discount rate is adequate for valuing any asset. 
 
 1. Evidence: Multivariate Cointegration and Parallelism Test running a Johansen cointegration test across the interest rate vector $[\text{FEDFUNDS}, \text{DGS2}, \text{DGS10}]$ to verify if the spatial distance between maturities remains strictly locked in a parallel, stationary equilibrium over time ($H_0: \Delta\text{Slope} = 0 \text{ and } \Delta\text{Curvature} = 0$).
 2. Decision Rule:
    - **IF** the rates are cointegrated with stationary spreads **AND** the changes in Slope and Curvature coordinates do not deviate significantly from zero over time:
      - Then NCE discounting is vindicated. The term structure can be safely compressed into a single baseline number representing the global cost of capital without losing valuation accuracy.
    - **IF** the cointegration relationships fail **OR** the spreads display structural non-stationarity and frequent non-parallel shifts:
      - Then the NCE single-scalar model introduces systemic mispricing. A single interest rate variable cannot accurately capture the multi-dimensional macro risk backdrop.
 3. Verdict: Null Hypothesis Rejected. The Johansen trace test rejects the null hypothesis of zero cointegrating vectors ($r = 0$) with a test statistic of $57.73$ (far exceeding the $1\%$ critical value of $37.22$). However, it confirms exactly $r = 1$ cointegrating relationship ($r \le 1$ statistic of $5.34$ is well below the $10\%$ critical value of $15.66$). While the rates share a single long-term stochastic drift, they are not bound in a rigid, parallel structure. Because $r < 2$, the spreads (Slope and Curvature) are free to dynamically warp, twist, and invert over time. Compressing this dynamic system into a single flat $r_f$ scalar ignores these major structural twists and introduces massive pricing distortions.
 
 
**NCE (extended — factor decomposition):** Checks if a firm's or asset's rate sensitivity depends on which part of the curve it's exposed to (e.g., BTC-USD's rate sensitivity behaves differently than UAL's, which carries real duration-like debt exposure). This is still an equilibrium-econometric question - "does this linear factor model have non-zero, asset-specific coefficients" - not yet a claim about feedback or emergent structure.
 
 1. Evidence: 3-Factor Regression Elasticity Test running rolling multivariate linear regressions of asset log returns against the individual, orthogonal coordinates of $\text{Level}_t$, $\text{Slope}_t$, and $\text{Curvature}_t$ under varying inflation (`CPIAUCSL`) backdrops.
 2. Decision Rule:
    - **IF** individual assets display highly asymmetrical, statistically significant elasticities to specific factors (e.g., `BTC-USD` reacting intensely to short-end `FEDFUNDS`/`DGS2` liquidity transformations, while `UAL` is tied strictly to long-end `DGS10` Slope developments):
      - Then the NCE-extended multi-factor view is validated over the single-scalar CAPM baseline. The macro environment still behaves as a (higher-dimensional) equilibrium object, but a single discount rate is the wrong reduction of it - different assets load on different, independently priced factors.
    - **IF** all asset sensitivities are uniform across Level, Slope, and Curvature variables, or if multi-factor sensitivities lack statistical significance:
      - Then the macro environment behaves symmetrically, and separating the term structure into sub-components adds no predictive value over the single-scalar CAPM baseline.
 3. Verdict: Null Hypothesis is NOT Rejected. Empirical asset elasticity plot reveals stark, asymmetric, and highly polarized responses to yield curve transformations.
 
```{r}
# --- 1. PREPARE DATA ---
macro_data <- master_df %>%
  select(
    Date,
    FEDFUNDS, DGS2, DGS10,
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj
  ) %>%
  drop_na() %>%
  mutate(
    Level = (FEDFUNDS + DGS2 + DGS10) / 3,
    Slope = DGS10 - FEDFUNDS,
    Curvature = DGS10 - (2 * DGS2) + FEDFUNDS
  )

# --- 2. STATISTICAL TESTS & AUTOMATED HYPOTHESIS SUMMARY ---

# Test A: Johansen Cointegration Test
rate_matrix <- macro_data %>% select(FEDFUNDS, DGS2, DGS10)
johansen_res <- ca.jo(rate_matrix, type = "trace", ecdet = "none", K = 2)
jo_summary <- summary(johansen_res)

# Dynamically extract Johansen values for automated decision making
r0_stat <- jo_summary@teststat[1]      # Test statistic for r <= 0
r0_crit <- jo_summary@cval[1, "5pct"]  # 5% critical value for r <= 0
r1_stat <- jo_summary@teststat[2]      # Test statistic for r <= 1
r1_crit <- jo_summary@cval[2, "5pct"]  # 5% critical value for r <= 1

if (r0_stat > r0_crit) {
  if (r1_stat > r1_crit) {
    jo_decision <- "REJECT H0 (Accept H1 for high cointegration)"
    jo_meaning  <- "Rates are locked in a strict parallel structure. Spreads are completely stationary."
  } else {
    jo_decision <- "REJECT H0 for r=0, FAIL TO REJECT H0 for r<=1"
    jo_meaning  <- "Exactly 1 cointegrating vector exists. Spreads (Slope and Curvature) are free to dynamically warp and twist over time. Compressing this into a single flat rate introduces systemic pricing distortions."
  }
} else {
  jo_decision <- "FAIL TO REJECT H0 (No Cointegration)"
  jo_meaning  <- "No stable long-term relationship exists between these interest rates. Spreads wander randomly."
}

# Test B: Nested F-Test / ANOVA for each asset
assets <- c("UAL", "PAL", "BTC", "SPY")
inference_test_matrix <- data.frame()

for (asset in assets) {
  fit_full <- lm(as.formula(paste(asset, "~ Level + Slope + Curvature")), data = macro_data)
  fit_reduced <- lm(as.formula(paste(asset, "~ Level")), data = macro_data)
  
  anova_test <- anova(fit_reduced, fit_full)
  f_stat <- anova_test$F[2]
  p_val  <- anova_test$`Pr(>F)`[2]
  
  # Programmatic evaluation of the hypothesis choice and meaning
  hyp_selected <- ifelse(p_val < 0.05, "Alternative (H1) Supported", "Null (H0) Supported")
  meaning_text <- ifelse(p_val < 0.05, 
                         "Curve shape changes (Slope/Curvature) carry unique, asymmetric risk explanatory power. Reduced CAPM models fail.", 
                         "Curve deformations do not matter. The asset responds purely to basic parallel shifts in the interest rate.")
  
  inference_test_matrix <- rbind(inference_test_matrix, data.frame(
    Asset = asset,
    F_Stat = round(f_stat, 2),
    p_Val = round(p_val, 4),
    Decision = hyp_selected,
    Meaning = meaning_text
  ))
}

# --- 3. PRINT POLISHED INFRASTRUCTURE SUMMARY VIA SPRINTF ---

# Flatten the asset test matrix into a text string for display
asset_summary_txt <- paste(
  apply(inference_test_matrix, 1, function(row) {
    sprintf("Asset: %s | F-Stat: %s | p-value: %s\n  Decision: %s\n  Meaning:  %s\n", 
            row['Asset'], row['F_Stat'], row['p_Val'], row['Decision'], row['Meaning'])
  }), 
  collapse = "\n"
)

formatted_output <- sprintf("
================================================================================
HYPOTHESIS SUMMARY REPORT
================================================================================

TEST 1: JOHANSEN COINTEGRATION (Term Structure Stability)
--------------------------------------------------------------------------------
* Test Statistic (r <= 0): %0.2f  (5%% Critical Value: %0.2f)
* Test Statistic (r <= 1): %0.2f  (5%% Critical Value: %0.2f)

* Selected Verdict: %s
* Financial Meaning: 
  %s


TEST 2: NESTED MULTI-FACTOR F-TESTS (Asymmetric Asset Elasticities)
--------------------------------------------------------------------------------
Null Hypothesis (H0): Slope and Curvature coefficients equal zero (Level is enough).
Alternative (H1): Deformations carry unique, independent pricing power.

%s
================================================================================
", 
r0_stat, r0_crit, r1_stat, r1_crit, 
jo_decision, 
jo_meaning, 
asset_summary_txt)

cat(formatted_output)
```

## Synthesis
The Macro Baseline represents our widest analytical lens, mapping the foundational environment that underpins all asset classes. Evaluating Level, Slope, and Curvature alongside inflation shows that the macroeconomy is *not* well described by a single flat cost of capital - but the richer description we've built so far (Litterman-Scheinkman factors, asymmetric elasticities) is still an equilibrium, orthogonal-factor account: a bigger NCE model, not yet a CAS one. If the data reveals non-parallel twists and highly localized asset sensitivities, it confirms that macro risk is structural and curve-geometry-dependent - but *structural* here still means "more dimensions of the same kind of linear object," which is a distinct claim from the CAS thesis that the term structure behaves as a dissipative, path-dependent, feedback-driven system. That distinction, and what would actually be needed to test the latter, is explored in the Footnote below rather than resolved here.

## Footnote:

**What would a genuinely CAS (not just higher-dimensional NCE) account of the yield curve look like?** This is explored conceptually only - it is not implemented in code above, and is flagged as a candidate direction for a future pass rather than a claim this report currently makes.

The Litterman-Scheinkman factors are a *kinematic* description: they tell you the curve's instantaneous shape (its three coordinates at time $t$), the same way a photograph tells you a system's position. They say nothing about *how* the curve got there, whether that process is reversible, or whether energy/information is being dissipated along the way - which is precisely what "complex adaptive" is supposed to mean in this report's framework. A genuine CAS account would need to swap the kinematic question ("what shape is the curve in?") for a thermodynamic/dynamical one ("is the process that produces this shape reversible or does it dissipate, and does it show memory of its own history?"). Two concrete, citable ways to operationalize that:

 1. **Irreversibility / hysteresis.** A rate-hiking cycle and a rate-cutting cycle that pass through the same Level/Slope/Curvature coordinates are not thermodynamically equivalent if the path taken to reach and leave that coordinate differs (e.g., central banks are well documented to cut more cautiously than they hike, and credit spreads that widen during stress don't retrace symmetrically as stress subsides). This is a direct, testable hysteresis-loop question - plot the curve's trajectory through $(L_t, S_t, C_t)$ space and check whether the up-cycle and down-cycle paths overlap or diverge - and it's the same "does the flow contract, expand, or trace out an irreversible loop" question raised in the 06_1 footnote, applied here at the macro scale instead of the single-asset scale.
 2. **Regime shifts as critical phenomena rather than smooth factor rotation.** Under NCE-extended factor rotation, the curve moves continuously through $(L, S, C)$ space. Under a genuine CAS account, transitions like sudden inversions or the depth of an inversion could instead resemble a system approaching a critical point - small, gradual pushes (rate hikes) that eventually trigger a large, discontinuous, self-reinforcing response (a credit event, a liquidity crunch), rather than a proportionate linear reaction. This is the same phenomenon Sornette (2003) models for equity crashes as critical, log-periodic events [17], and it has a general physical mechanism in self-organized criticality (Bak, Tang, & Wiesenfeld, 1987) [18] - systems that are slowly, continuously driven (central bank tightening) but release stress in sudden, scale-free bursts rather than smoothly. On the economics side, this pairs naturally with Minsky's (1992) Financial Instability Hypothesis [19], which makes a structurally similar claim in economic (not physical) language: stability itself breeds the conditions for instability.

Both directions are genuine research extensions rather than something this section currently tests - flagging them here is meant to keep the "NCE vs. CAS" framing honest about which of its own claims are actually earned by the code above, versus which are aspirational for now.
