
Here, we view the entire context of the market and economy.

## Structure

**NCE**: via CAPM, collapses the entire term structure into one risk-free rate — a single number standing in for "the cost of money."
 
 1. Raw Input: Daily DGS10, DGS2, or monthly FEDFUNDS.
 2. Enriched State: A single risk-free scalar ($r_f$) selected based on the closest maturity to the asset's cash flows (e.g., using DGS2 for a 2-year horizon contract).
 3. Compression: The chosen flat scalar applied as a uniform baseline denominator in linear valuation equations.

```{r}
# --- 1. DYNAMIC PARAMETER ESTIMATION FROM DATA ---
daily_returns <- master_df %>%
  select(
    Date,
    SPY = SPY.LogAdj,
    PAL = PAL.LogAdj,
    UAL = UAL.LogAdj,
    BTC = `BTC-USD.LogAdj`
  ) %>%
  drop_na() %>%
  select(-Date)

scale_factor <- 252

# A. Calculate Dynamic Expected Annualized Returns (mu)
mu <- colMeans(daily_returns) * scale_factor

# B. Calculate Dynamic Annualized Volatilities (vol)
vol <- apply(daily_returns, 2, sd) * sqrt(scale_factor)

# C. Calculate Dynamic Correlation Matrix
corr_matrix <- cor(daily_returns)

# D. Calculate Dynamic Covariance Matrix (Sigma)
D <- diag(vol)
Sigma <- D %*% corr_matrix %*% D
names(mu) <- colnames(daily_returns)
colnames(Sigma) <- rownames(Sigma) <- colnames(daily_returns)

# --- 2. SOLVE LONG-ONLY TANGENCY PORTFOLIO (M) VIA QUADRATIC PROGRAMMING ---
rf <- 0.04168  # Standard policy/risk-free rate (4.168%)

# Quadratic Programming Setup: Min 0.5 * y' Sigma y  s.t.  y' (mu - rf) = 1  and  y >= 0
Dmat <- Sigma
dvec <- rep(0, length(mu))
Amat <- cbind(mu - rf, diag(length(mu)))
bvec <- c(1, rep(0, length(mu)))

qp_tangency <- solve.QP(Dmat = Dmat, dvec = dvec, Amat = Amat, bvec = bvec, meq = 1)

# Normalize weights so they sum strictly to 100%
w_tangency <- qp_tangency$solution / sum(qp_tangency$solution)
names(w_tangency) <- names(mu)

# Tangency portfolio metrics
ret_m <- as.numeric(t(w_tangency) %*% mu)
sd_m  <- as.numeric(sqrt(t(w_tangency) %*% Sigma %*% w_tangency))
sharpe_m <- (ret_m - rf) / sd_m

# --- 3. GENERATE LONG-ONLY EFFICIENT FRONTIER POINTS ---
target_returns <- seq(min(mu), max(mu), length.out = 200)
frontier_list <- list()

for (i in seq_along(target_returns)) {
  r_target <- target_returns[i]
  
  # Min 0.5 * w' Sigma w  s.t.  w' mu = r_target, w' 1 = 1, w >= 0
  Amat_ef <- cbind(mu, rep(1, length(mu)), diag(length(mu)))
  bvec_ef <- c(r_target, 1, rep(0, length(mu)))
  
  qp_ef <- tryCatch({
    solve.QP(Dmat = Dmat, dvec = dvec, Amat = Amat_ef, bvec = bvec_ef, meq = 2)
  }, error = function(e) NULL)
  
  if (!is.null(qp_ef)) {
    w_ef <- qp_ef$solution
    sd_ef <- sqrt(as.numeric(t(w_ef) %*% Sigma %*% w_ef))
    frontier_list[[i]] <- tibble(Return = r_target, SD = sd_ef)
  }
}

frontier_points <- bind_rows(frontier_list)

# --- 4. PRINT CONSTRAINED NCE TANGENCY PORTFOLIO OUTPUT ---
cat("======================================================================\n")
cat("          LONG-ONLY NCE TANGENCY PORTFOLIO (M) OPTIMAL OUTPUT          \n")
cat("======================================================================\n")
cat("Risk-Free Rate (rf)             :", scales::percent(rf, accuracy = 0.01), "\n")
cat("Expected Annualized Return (E[R]):", scales::percent(ret_m, accuracy = 0.01), "\n")
cat("Expected Annualized Risk (SD)   :", scales::percent(sd_m, accuracy = 0.01), "\n")
cat("Tangency Sharpe Ratio           :", round(sharpe_m, 4), "\n")
cat("----------------------------------------------------------------------\n")
cat("OPTIMAL REALISTIC ASSET WEIGHTS:\n")

tangency_summary <- data.frame(
  Asset = names(mu),
  Allocation = scales::percent(as.numeric(w_tangency), accuracy = 0.01)
)

print(tangency_summary, row.names = FALSE)
cat("======================================================================\n\n")

# --- 5. RISK DECOMPOSITION FOR UAL ---
ual_ret <- mu["UAL"]
ual_sd  <- vol["UAL"]
sd_a    <- (ual_ret - rf) / sharpe_m

# --- 6. RENDER THE DYNAMIC PLOT ---
ggplot() +
  geom_path(data = frontier_points, aes(x = SD, y = Return), 
            color = "black", linewidth = 1.1) +
  geom_abline(intercept = rf, slope = sharpe_m, 
              color = "blue", linewidth = 1, linetype = "solid") +
  
  # Horizontal and vertical drop lines
  annotate("segment", x = 0, xend = ual_sd, y = ual_ret, yend = ual_ret, 
           color = "grey50", linetype = "dashed") +
  annotate("segment", x = sd_a, xend = sd_a, y = 0, yend = ual_ret, 
           color = "grey50", linetype = "dotted") +
  annotate("segment", x = ual_sd, xend = ual_sd, y = 0, yend = ual_ret, 
           color = "grey50", linetype = "dotted") +
  
  # Risk Brackets at the bottom of the plot
  annotate("segment", x = 0.01, xend = sd_a - 0.01, y = 0.01, yend = 0.01, 
           arrow = arrow(ends = "both", angle = 90, length = unit(.1, "cm")), color = "#D14949", linewidth = 0.8) +
  annotate("text", x = sd_a / 2, y = 0.025, label = "Systematic\nRisk", 
           color = "#D14949", size = 3, fontface = "bold") +
  
  annotate("segment", x = sd_a + 0.01, xend = ual_sd - 0.01, y = 0.01, yend = 0.01, 
           arrow = arrow(ends = "both", angle = 90, length = unit(.1, "cm")), color = "#20639B", linewidth = 0.8) +
  annotate("text", x = (sd_a + ual_sd) / 2, y = 0.025, label = "Diversifiable\n(Idiosyncratic) Risk", 
           color = "#20639B", size = 3, fontface = "bold") +
  
  # Reference Nodes
  annotate("point", x = 0, y = rf, color = "black", size = 4, shape = 18) +
  annotate("text", x = 0.01, y = rf + 0.005, label = "r[f]", parse = TRUE, fontface = "bold", hjust = 0) +
  
  annotate("point", x = sd_m, y = ret_m, color = "darkgreen", size = 4) +
  annotate("text", x = sd_m - 0.015, y = ret_m + 0.01, label = "M (Tangency Portfolio)", 
           color = "darkgreen", fontface = "bold", hjust = 1) +
  
  annotate("point", x = sd_a, y = ual_ret, color = "black", size = 3) +
  annotate("text", x = sd_a + 0.01, y = ual_ret - 0.01, label = "A", fontface = "bold") +
  
  # Asset labels
  geom_point(data = data.frame(SD = vol, Return = mu, Asset = names(mu)), 
             aes(x = SD, y = Return, color = Asset), size = 4) +
  geom_text(data = data.frame(SD = vol, Return = mu, Asset = names(mu)), 
            aes(x = SD, y = Return, label = Asset), vjust = -1.2, fontface = "bold", size = 3.5) +
  
  scale_color_manual(values = c("SPY" = "#3CAEA3", "PAL" = "#F6D55C", "UAL" = "#20639B", "BTC" = "#D14949")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_cartesian(xlim = c(0, max(vol) * 1.2), ylim = c(min(0, min(mu)), max(mu) * 1.2)) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Long-Only Capital Allocation Line & Empirical Efficient Frontier",
    subtitle = "Constrained Portfolio Optimization (No Short-Selling: Weights Between 0% and 100%)",
    x = "Portfolio Total Risk (Standard Deviation)",
    y = "Expected Return (E[R])"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold")
  )
```




**NCE (extended — factor decomposition):** The single-scalar CAPM view above is the *degenerate* case of a richer, but still classical, equilibrium-econometrics model: the Litterman & Scheinkman (1991) three-factor decomposition of the yield curve [4]. This is a linear, orthogonal (PCA-derived) re-coordinatization of the same term structure - it relaxes "one rate" to "three factors," but it is still a static, reversible, equilibrium description with no feedback, heterogeneous agents, or path-dependent mechanism attached. That is why it belongs on the NCE side of this report rather than the CAS side: richer linear algebra is not, by itself, evidence of a complex adaptive system. (See the Footnote at the end of this section for what a genuinely CAS-consistent account of the yield curve would need to look like instead.)

 1. Raw Input: Co-moving rates across the term spectrum (FEDFUNDS, DGS2, and DGS10) matched with macroeconomic inflation context (CPIAUCSL).
 2. Enriched State: A 3-dimensional coordinate state vector mapping the classical Litterman-Scheinkman yield curve factors directly:
   - $\text{Level}_t = \frac{\text{FEDFUNDS}_t + \text{DGS2}_t + \text{DGS10}_t}{3}$ (Overall altitude of interest rates)
   - $\text{Slope}_t = \text{DGS10}_t - \text{FEDFUNDS}_t$ (The long-to-short macro spread)
   - $\text{Curvature}_t = (\text{DGS10}_t - \text{DGS2}_t) - (\text{DGS2}_t - \text{FEDFUNDS}_t) = \text{DGS10}_t - 2\cdot\text{DGS2}_t + \text{FEDFUNDS}_t$ (The structural bend/twist of the medium-term belly relative to the anchors)
 3. Compression: Projecting this 3D coordinate $[L_t, S_t, C_t]$ onto distinct, discrete macro-regime charts (e.g., Bear Flattener, Bull Steeper, or Inverted Belly).

This directly invokes Litterman & Scheinkman's (1991) three-factor principal-component result, which finds that over 95% of term-structure variation is captured by these three orthogonal geometric transformations (Level, Slope, and Curvature) [4].

```{r}
# --- 1. DATA ENRICHMENT ---
macro_enriched <- master_df %>%
  select(Date, FEDFUNDS, DGS2, DGS10, CPIAUCSL) %>%
  # Carry forward monthly CPI to daily observations if dataframe is daily
  mutate(CPIAUCSL = zoo::na.locf(CPIAUCSL, na.rm = FALSE)) %>%
  drop_na() %>%
  mutate(
    
    # Calculate Nominal Factors
    Level = (FEDFUNDS + DGS2 + DGS10) / 3,
    Slope = DGS10 - FEDFUNDS,
    Curvature = DGS10 - (2 * DGS2) + FEDFUNDS,
  ) %>%
  drop_na()

# --- 2. PANEL A: RAW TERM STRUCTURE ---
raw_long <- macro_enriched %>%
  select(Date, FEDFUNDS, DGS2, DGS10) %>%
  pivot_longer(-Date, names_to = "Maturity", values_to = "Rate")

p_raw <- ggplot(raw_long, aes(x = Date, y = Rate, color = Maturity)) +
  geom_line(size = 0.8, alpha = 0.8) +
  scale_color_manual(values = c("FEDFUNDS" = "#20639B", "DGS2" = "#3CAEA3", "DGS10" = "#F6D55C")) +
  theme_minimal(base_size = 11) +
  labs(
    title = "Raw Term Structure Deformations",
    y = "Interest Rate (%)",
    x = NULL
  ) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

# --- 3. PANEL B: LITTERMAN-SCHEINKMAN FACTORS & INFLATION STATE ---
# Expanded selection to include both real level transformations and the inflation rate
factors_long <- macro_enriched %>%
  select(Date, Level, Slope, Curvature) %>%
  pivot_longer(-Date, names_to = "Coordinate", values_to = "Value")

# Enforce a logical ordering for the facet layout
factors_long$Coordinate <- factor(
  factors_long$Coordinate, 
  levels = c("Level", "Slope", "Curvature")
)

p_factors <- ggplot(factors_long, aes(x = Date, y = Value, color = Coordinate)) +
  geom_line(size = 0.8, alpha = 0.8) +
  scale_color_manual(values = c(
    "Level"         = "#D14949", # Nominal baseline
    "Slope"         = "#3F51B5", # Spread
    "Curvature"     = "#4CAF50"  # Twist

  )) +
  facet_wrap(~Coordinate, scales = "free_y", ncol = 1) +
  theme_minimal(base_size = 11) +
  labs(
    title = "Litterman-Scheinkman components",
    subtitle = "",
    y = "Coordinate Value / Rate (%)",
    x = "Date"
  ) +
  theme(
    legend.position = "none", 
    plot.title = element_text(face = "bold"), 
    strip.text = element_text(face = "bold")
  )

# --- 4. RENDER STRUCTURE PLOT ---
# grid.arrange(p_raw, p_factors, ncol = 2)
p_raw
p_factors
```


**CAS**:

```{r}
# --- 1. REGIME FILTERING & DATA PREPARATION ---
# Filter data for High-VIX stress regime (VIX >= 23.9) and set parameters
high_vix_threshold <- 23.9
horizon_T <- 21      # 21-day investment horizon
alpha <- 0.05        # 95% confidence level for CVaR
assets <- c("SPY", "PAL", "UAL", "BTC")

returns_df <- master_df %>%
  select(
    Date,
    SPY = SPY.LogAdj,
    PAL = PAL.LogAdj,
    UAL = UAL.LogAdj,
    BTC = `BTC-USD.LogAdj`,
    VIX = VIX.Adjusted
  ) %>%
  drop_na()

# Isolate full series for Hurst exponent estimation and High-VIX subset for tail risk
full_returns <- returns_df %>% select(all_of(assets))
high_vix_returns <- returns_df %>% filter(VIX >= high_vix_threshold) %>% select(all_of(assets))

# --- 2. ESTIMATE HURST EXPONENTS & FRACTIONAL CVAR ---
# Compute Hurst exponent (H) per asset over full return history
hurst_vector <- sapply(full_returns, function(x) {
  tryCatch(pracma::hurstexp(x, display = FALSE)$Hs, error = function(e) 0.5)
})
names(hurst_vector) <- assets

# Compute empirical daily CVaR (Expected Shortfall at 95%) during High-VIX regime
calc_daily_cvar <- function(returns_matrix, conf_level = 0.05) {
  apply(returns_matrix, 2, function(r) {
    var_cutoff <- quantile(r, probs = conf_level)
    -mean(r[r <= var_cutoff])
  })
}

daily_cvar_vix <- calc_daily_cvar(high_vix_returns, conf_level = alpha)

# --- 3. FRACTIONAL RISK PARITY OPTIMIZATION ENGINE ---
# Function to calculate portfolio Fractional CVaR scaled by weighted Hurst exponent
portfolio_fractional_cvar <- function(w, returns_matrix, hurst_vec, T_days, conf_level = 0.05) {
  port_returns <- as.matrix(returns_matrix) %*% w
  var_cutoff <- quantile(port_returns, probs = conf_level)
  base_cvar <- -mean(port_returns[port_returns <= var_cutoff])
  
  # Scale portfolio risk using weighted fractional scaling exponent (T^H)
  w_hurst <- sum(w * hurst_vec)
  return(base_cvar * (T_days ^ w_hurst))
}

# Objective function: Minimize variance of Marginal Risk Contributions (Equal Risk Contribution)
risk_parity_objective <- function(w, returns_matrix, hurst_vec, T_days, conf_level = 0.05) {
  # Normalize weights inside objective
  w <- w / sum(w)
  n <- length(w)
  eps <- 1e-5
  mrc <- numeric(n)
  
  base_p_cvar <- portfolio_fractional_cvar(w, returns_matrix, hurst_vec, T_days, conf_level)
  
  # Calculate numerical marginal risk contributions (MRC)
  for (i in 1:n) {
    w_plus <- w
    w_plus[i] <- w_plus[i] + eps
    cvar_plus <- portfolio_fractional_cvar(w_plus, returns_matrix, hurst_vec, T_days, conf_level)
    mrc[i] <- (cvar_plus - base_p_cvar) / eps
  }
  
  # Absolute risk contribution per asset
  rc <- w * mrc
  target_rc <- sum(rc) / n
  
  # Sum of squared errors from equal risk contribution
  return(sum((rc - target_rc)^2))
}

# Run bounded optimization (w >= 0)
init_weights <- rep(1 / length(assets), length(assets))

opt_result <- optim(
  par = init_weights,
  fn = risk_parity_objective,
  returns_matrix = high_vix_returns,
  hurst_vec = hurst_vector,
  T_days = horizon_T,
  conf_level = alpha,
  method = "L-BFGS-B",
  lower = rep(0.001, length(assets)),
  upper = rep(1.0, length(assets))
)

# Normalize final optimal weights
w_opt <- opt_result$par / sum(opt_result$par)
names(w_opt) <- assets

# --- 4. RENDER OUTPUT REPORT ---
cat("======================================================================\n")
cat("    FRACTIONAL REGIME-SWITCHING CVAR RISK PARITY PORTFOLIO OUTPUT     \n")
cat("======================================================================\n")
cat("Target Holding Horizon (Days, T):", horizon_T, "\n")
cat("VIX Regime Stress Filter        : VIX >=", high_vix_threshold, "\n")
cat("Tail Risk Confidence Level     :", paste0((1 - alpha) * 100, "% CVaR"), "\n")
cat("----------------------------------------------------------------------\n")

output_summary <- tibble(
  Asset = assets,
  Hurst_Exponent = round(hurst_vector, 3),
  High_VIX_Daily_CVaR = scales::percent(daily_cvar_vix, accuracy = 0.01),
  Optimal_Allocation = scales::percent(w_opt, accuracy = 0.01)
)

print(as.data.frame(output_summary), row.names = FALSE)
cat("======================================================================\n")
```

## Dynamics

**NCE**: Assumes shifts in "the" rate are parallel and exogenous — everything moves together. 
 
 1. Transition (Conservative): The curve transitions uniformly. Short-rate models treat the entire curve as a single string pulled by a central mean-reverting anchor.
 2. Grouped Context of Transitions: Macroeconomic adjustments are uniform. A shift in central bank policy alters the short-term rate, and that change transfers linearly up the maturity spectrum, maintaining rigid, parallel tracking across `FEDFUNDS`, `DGS2`, and `DGS10`.

**NCE (extended — factor decomposition):** Allows the curve to twist (short end rising while the long end holds, or vice versa), which the single-scalar CAPM model above can't represent, but which is still just a richer *linear* decomposition of the same equilibrium object.
 
 1. Transition (Conservative, higher-dimensional): The curve moves via non-parallel but still linear factor rotation. For example, during an aggressive central bank tightening cycle, FEDFUNDS shoots up, DGS2 spikes violently due to medium-term policy stress, while DGS10 lags behind due to long-term economic deceleration.
 2. Grouped Transitions: Systemic rate movements are grouped by their geometric deformations (Litterman & Scheinkman, 1991 [4]). Movements are re-expressed along the curve's Level, Slope, and Curvature axes and act as macroeconomic filters that alter the valuation dynamics of specific assets (e.g., separating capital between pure liquidity proxies like `BTC-USD` and debt-laden operational equities like `UAL`) - but the *mechanism* generating those movements here is still an orthogonal factor rotation, not agent feedback or path dependence.


```{r}
# --- DATA PREPARATION & MACRO FACTOR DECOMPOSITION ---
# Extract assets, interest rates, and inflation context into a unified matrix
macro_enriched <- master_df %>%
  select(
    Date,
    FEDFUNDS, 
    DGS2, 
    DGS10,
    CPIAUCSL,
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj
  ) %>%
  # Carry forward monthly CPI to daily observations to prevent data gaps
  mutate(CPIAUCSL = zoo::na.locf(CPIAUCSL, na.rm = FALSE)) %>%
  drop_na() %>%
  mutate(
    # Calculate Year-over-Year Inflation (Assumes monthly input. If daily, change lag to 252)
    Inflation_YoY = (CPIAUCSL / lag(CPIAUCSL, 12) - 1) * 100,
    
    # Calculate Nominal Litterman-Scheinkman Factors
    Level = (FEDFUNDS + DGS2 + DGS10) / 3,
    Slope = DGS10 - FEDFUNDS,
    Curvature = DGS10 - (2 * DGS2) + FEDFUNDS,
    
    # Calculate Real Level (Deflated Baseline)
    Real_Level = Level - Inflation_YoY
  ) %>%
  drop_na()

# --- COINTEGRATION TEST (JOHANSEN) ---
# Verify if nominal yield maturities share a locked, stationary equilibrium
rate_matrix <- macro_enriched %>% select(FEDFUNDS, DGS2, DGS10)
johansen_test <- ca.jo(rate_matrix, type = "trace", ecdet = "none", K = 2)

# --- MULTI-FACTOR REGRESSION SENSITIVITY ---
# Analyze asymmetric asset elasticities against nominal factors and inflation shocks
assets <- c("UAL", "PAL", "BTC", "SPY")
expanded_sensitivity <- data.frame()

for (asset in assets) {
  formula_str <- paste(asset, "~ Level + Slope + Curvature + Inflation_YoY")
  fit <- lm(as.formula(formula_str), data = macro_enriched)
  
  fit_summary <- tidy(fit, conf.int = TRUE) %>%
    mutate(Asset = asset) %>%
    filter(term != "(Intercept)")
  
  expanded_sensitivity <- rbind(expanded_sensitivity, fit_summary)
}

jo_txt   <- paste(capture.output(summary(johansen_test)), collapse = "\n")
sens_txt <- paste(capture.output(print(expanded_sensitivity %>% select(Asset, term, estimate, statistic, p.value))), collapse = "\n")

formatted_output <- sprintf("
=========================================
        COINTEGRATION (PARALLELISM)    
=========================================
%s

=========================================
     ASYMMETRIC SENSITIVITIES (LS + CPI)  
=========================================
%s", 
jo_txt, 
sens_txt
)

cat(formatted_output)

```

```{r}
# Merge VIX back in to color the trajectory by market stress/regime
macro_dynamics <- macro_enriched %>%
  left_join(select(master_df, Date, VIX = VIX.Adjusted), by = "Date") %>%
  drop_na()

# Plot the state-space trajectory
ggplot(macro_dynamics, aes(x = Level, y = Slope)) +
  # Draw the sequential historical path
  geom_path(aes(color = VIX), size = 0.6, alpha = 0.6) +
  # Highlight the starting and ending points of the macro-regime
  geom_point(data = filter(macro_dynamics, Date == min(Date)), color = "blue", size = 4, shape = 17) +
  geom_point(data = filter(macro_dynamics, Date == max(Date)), color = "red", size = 4, shape = 15) +
  # Color scale mapping market fear/regime
  scale_color_gradientn(
    colors = c("#20639B", "#F6D55C", "#D14949"), 
    name = "VIX (Systemic Stress)"
  ) +
  annotate("text", x = min(macro_dynamics$Level), y = max(macro_dynamics$Slope), 
           label = "▲ Blue Triangle: Path Start\n■ Red Square: Path End", 
           hjust = 0, vjust = 1, size = 3, fontface = "italic") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Yield Curve Trajectory in Macro Phase Space",
    subtitle = "Systemic path-dependent evolution: Level (Altitude) vs. Slope (Spread)",
    x = "System Level (Average Interest Rate)",
    y = "System Slope (DGS10 - FEDFUNDS)"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

## Inference

**NCE**: Checks if a single discount rate is adequate for valuing any asset. 
 
 1. Evidence: Multivariate Cointegration and Parallelism Test running a Johansen cointegration test across the interest rate vector $[\text{FEDFUNDS}, \text{DGS2}, \text{DGS10}]$ to verify if the spatial distance between maturities remains strictly locked in a parallel, stationary equilibrium over time ($H_0: \Delta\text{Slope} = 0 \text{ and } \Delta\text{Curvature} = 0$).
 2. Decision Rule:
    - **IF** the rates are cointegrated with stationary spreads **AND** the changes in Slope and Curvature coordinates do not deviate significantly from zero over time:
      - Then NCE discounting is vindicated. The term structure can be safely compressed into a single baseline number representing the global cost of capital without losing valuation accuracy.
    - **IF** the cointegration relationships fail **OR** the spreads display structural non-stationarity and frequent non-parallel shifts:
      - Then the NCE single-scalar model introduces systemic mispricing. A single interest rate variable cannot accurately capture the multi-dimensional macro risk backdrop.
 3. Verdict: Null Hypothesis Rejected. The Johansen trace test rejects the null hypothesis of zero cointegrating vectors ($r = 0$) with a test statistic of $57.73$ (far exceeding the $1\%$ critical value of $37.22$). However, it confirms exactly $r = 1$ cointegrating relationship ($r \le 1$ statistic of $5.34$ is well below the $10\%$ critical value of $15.66$). While the rates share a single long-term stochastic drift, they are not bound in a rigid, parallel structure. Because $r < 2$, the spreads (Slope and Curvature) are free to dynamically warp, twist, and invert over time. Compressing this dynamic system into a single flat $r_f$ scalar ignores these major structural twists and introduces massive pricing distortions.
 
 
**NCE (extended — factor decomposition):** Checks if a firm's or asset's rate sensitivity depends on which part of the curve it's exposed to (e.g., BTC-USD's rate sensitivity behaves differently than UAL's, which carries real duration-like debt exposure). This is still an equilibrium-econometric question - "does this linear factor model have non-zero, asset-specific coefficients" - not yet a claim about feedback or emergent structure.
 
 1. Evidence: 3-Factor Regression Elasticity Test running rolling multivariate linear regressions of asset log returns against the individual, orthogonal coordinates of $\text{Level}_t$, $\text{Slope}_t$, and $\text{Curvature}_t$ under varying inflation (`CPIAUCSL`) backdrops.
 2. Decision Rule:
    - **IF** individual assets display highly asymmetrical, statistically significant elasticities to specific factors (e.g., `BTC-USD` reacting intensely to short-end `FEDFUNDS`/`DGS2` liquidity transformations, while `UAL` is tied strictly to long-end `DGS10` Slope developments):
      - Then the NCE-extended multi-factor view is validated over the single-scalar CAPM baseline. The macro environment still behaves as a (higher-dimensional) equilibrium object, but a single discount rate is the wrong reduction of it - different assets load on different, independently priced factors.
    - **IF** all asset sensitivities are uniform across Level, Slope, and Curvature variables, or if multi-factor sensitivities lack statistical significance:
      - Then the macro environment behaves symmetrically, and separating the term structure into sub-components adds no predictive value over the single-scalar CAPM baseline.
 3. Verdict: Null Hypothesis is NOT Rejected. Empirical asset elasticity plot reveals stark, asymmetric, and highly polarized responses to yield curve transformations.
 
```{r}
# --- 1. PREPARE DATA ---
macro_data <- master_df %>%
  select(
    Date,
    FEDFUNDS, DGS2, DGS10,
    UAL = UAL.LogAdj, 
    PAL = PAL.LogAdj, 
    BTC = `BTC-USD.LogAdj`, 
    SPY = SPY.LogAdj
  ) %>%
  drop_na() %>%
  mutate(
    Level = (FEDFUNDS + DGS2 + DGS10) / 3,
    Slope = DGS10 - FEDFUNDS,
    Curvature = DGS10 - (2 * DGS2) + FEDFUNDS
  )

# --- 2. STATISTICAL TESTS & AUTOMATED HYPOTHESIS SUMMARY ---

# Test A: Johansen Cointegration Test
rate_matrix <- macro_data %>% select(FEDFUNDS, DGS2, DGS10)
johansen_res <- ca.jo(rate_matrix, type = "trace", ecdet = "none", K = 2)
jo_summary <- summary(johansen_res)

# Dynamically extract Johansen values for automated decision making
r0_stat <- jo_summary@teststat[1]      # Test statistic for r <= 0
r0_crit <- jo_summary@cval[1, "5pct"]  # 5% critical value for r <= 0
r1_stat <- jo_summary@teststat[2]      # Test statistic for r <= 1
r1_crit <- jo_summary@cval[2, "5pct"]  # 5% critical value for r <= 1

if (r0_stat > r0_crit) {
  if (r1_stat > r1_crit) {
    jo_decision <- "REJECT H0 (Accept H1 for high cointegration)"
    jo_meaning  <- "Rates are locked in a strict parallel structure. Spreads are completely stationary."
  } else {
    jo_decision <- "REJECT H0 for r=0, FAIL TO REJECT H0 for r<=1"
    jo_meaning  <- "Exactly 1 cointegrating vector exists. Spreads (Slope and Curvature) are free to dynamically warp and twist over time. Compressing this into a single flat rate introduces systemic pricing distortions."
  }
} else {
  jo_decision <- "FAIL TO REJECT H0 (No Cointegration)"
  jo_meaning  <- "No stable long-term relationship exists between these interest rates. Spreads wander randomly."
}

# Test B: Nested F-Test / ANOVA for each asset
assets <- c("UAL", "PAL", "BTC", "SPY")
inference_test_matrix <- data.frame()

for (asset in assets) {
  fit_full <- lm(as.formula(paste(asset, "~ Level + Slope + Curvature")), data = macro_data)
  fit_reduced <- lm(as.formula(paste(asset, "~ Level")), data = macro_data)
  
  anova_test <- anova(fit_reduced, fit_full)
  f_stat <- anova_test$F[2]
  p_val  <- anova_test$`Pr(>F)`[2]
  
  # Programmatic evaluation of the hypothesis choice and meaning
  hyp_selected <- ifelse(p_val < 0.05, "Alternative (H1) Supported", "Null (H0) Supported")
  meaning_text <- ifelse(p_val < 0.05, 
                         "Curve shape changes (Slope/Curvature) carry unique, asymmetric risk explanatory power. Reduced CAPM models fail.", 
                         "Curve deformations do not matter. The asset responds purely to basic parallel shifts in the interest rate.")
  
  inference_test_matrix <- rbind(inference_test_matrix, data.frame(
    Asset = asset,
    F_Stat = round(f_stat, 2),
    p_Val = round(p_val, 4),
    Decision = hyp_selected,
    Meaning = meaning_text
  ))
}

# --- 3. PRINT POLISHED INFRASTRUCTURE SUMMARY VIA SPRINTF ---

# Flatten the asset test matrix into a text string for display
asset_summary_txt <- paste(
  apply(inference_test_matrix, 1, function(row) {
    sprintf("Asset: %s | F-Stat: %s | p-value: %s\n  Decision: %s\n  Meaning:  %s\n", 
            row['Asset'], row['F_Stat'], row['p_Val'], row['Decision'], row['Meaning'])
  }), 
  collapse = "\n"
)

formatted_output <- sprintf("
================================================================================
HYPOTHESIS SUMMARY REPORT
================================================================================

TEST 1: JOHANSEN COINTEGRATION (Term Structure Stability)
--------------------------------------------------------------------------------
* Test Statistic (r <= 0): %0.2f  (5%% Critical Value: %0.2f)
* Test Statistic (r <= 1): %0.2f  (5%% Critical Value: %0.2f)

* Selected Verdict: %s
* Financial Meaning: 
  %s


TEST 2: NESTED MULTI-FACTOR F-TESTS (Asymmetric Asset Elasticities)
--------------------------------------------------------------------------------
Null Hypothesis (H0): Slope and Curvature coefficients equal zero (Level is enough).
Alternative (H1): Deformations carry unique, independent pricing power.

%s
================================================================================
", 
r0_stat, r0_crit, r1_stat, r1_crit, 
jo_decision, 
jo_meaning, 
asset_summary_txt)

cat(formatted_output)
```

## Synthesis
The Macro Baseline represents our widest analytical lens, mapping the foundational environment that underpins all asset classes. Evaluating Level, Slope, and Curvature alongside inflation shows that the macroeconomy is *not* well described by a single flat cost of capital - but the richer description we've built so far (Litterman-Scheinkman factors, asymmetric elasticities) is still an equilibrium, orthogonal-factor account: a bigger NCE model, not yet a CAS one. If the data reveals non-parallel twists and highly localized asset sensitivities, it confirms that macro risk is structural and curve-geometry-dependent - but *structural* here still means "more dimensions of the same kind of linear object," which is a distinct claim from the CAS thesis that the term structure behaves as a dissipative, path-dependent, feedback-driven system. That distinction, and what would actually be needed to test the latter, is explored in the Footnote below rather than resolved here.

