# Financial Investigation

In this section, we model and investigate the relationship of `UAL`, `PAL`, `BTC-USD`, and `SPY` to the wider economy. The analysis proceeds outward in scale: Price Movement -> Risk and Correlation -> Macro Baseline.

Each subsection is a wider lens than the one before it, zooming out from an asset's daily behavior to the macroeconomic backdrop that conditions it. Concretely, each section asks a progressively broader question:

 1. **Asset Movement** - how does *one* asset behave in isolation?
 2. **Risk and Correlation** - how do assets behave *relative to one another*?
 3. **Macro Baseline** - what *shared backdrop* (rates, inflation, the dollar's cost of capital) is quietly setting the terms for all of the above?

This ordering is deliberate, not just organizational. We can't sensibly ask "how does UAL's correlation with SPY behave under stress" before establishing what "return" even means for UAL itself; and we can't meaningfully ask "what macro regime are we in" until regime-dependent behavior has already surfaced empirically in the correlation data. Each section's findings become the next section's motivation: unexplained memory in Price Movement motivates the correlation-breakdown tests in Risk and Correlation, and unexplained co-movement across *assets* in Risk and Correlation motivates the search for a common macro driver in Macro Baseline.

This progression also tracks data frequency. Price movement is fast, granular, and asset-specific (daily closes); correlation structure lives at the weekly-to-monthly cadence where regimes become visible; and the macro baseline (FEDFUNDS, DGS10, CPIAUCSL) is slow-moving, low-frequency, and shared across every asset in the dataset.

Within each subsection, we apply a consistent four-part structure:

 1. **Structure** - what is the raw state (e.g., log-adjusted returns) to be analyzed, what is the enriched state derived from it, and how is that enriched state compressed into something tractable?
 2. **Dynamics** - what governs the *transition* between states over time? Is that transition conservative (energy/information-preserving, reversible) or dissipative (lossy, path-dependent)? What surrounding context lets us group transitions into regimes?
 3. **Inference** - what statistical evidence discriminates between competing accounts of the dynamics, what decision rule turns that evidence into a verdict, and what is the verdict?
 4. **Synthesis** - given Structure, Dynamics, and Inference, and given both theoretical paradigms below, what can we actually say about this asset/relationship/regime?

For each of the four parts, we deliberately hold up **two competing theoretical paradigms** as lenses, since they lead to different structural assumptions, different transition models, and ultimately different risk conclusions from the *same* underlying data:

 1. **Neoclassical Economics (NCE)** - the classical finance paradigm built on the Efficient Market Hypothesis , which holds that market prices fully reflect all available information (Fama, 1970 [2]), and on the mean-variance/CAPM tradition (Markowitz, 1952 [6]; Sharpe, 1964 [7]). Under NCE, the market is treated as a linear, rational, mechanical system: price innovations are independent, identically distributed, and Gaussian, and the system naturally tends toward a stable equilibrium in which risk scales predictably (e.g., via the square-root-of-time rule).
 2. **Complex Adaptive Systems (CAS)** - a paradigm drawn from *econophysics*, the application of statistical-physics tools to financial systems (Mantegna & Stanley, 1999 [10]). Rather than a sterile, equilibrium-seeking machine, CAS treats the market as an open, dissipative system: heterogeneous agents, feedback loops, memory, and friction continually push the system away from equilibrium (Bouchaud, Farmer, & Lillo, 2009 [1]; Lux & Marchesi, 1999 [11]). This tradition traces back to Mandelbrot's (1963) observation that speculative price changes are non-Gaussian and heavy-tailed [5], and is summarized empirically in Cont's (2001) catalogue of "stylized facts" of asset returns [12] - volatility clustering, heavy tails, and long-range dependence that the NCE baseline cannot generate.

Each Structure/Dynamics/Inference/Synthesis step below therefore reports both an NCE account and a CAS account side by side: Objective, Code (SQL/dplyr), Output, and Financial Interpretation for each.

> **Note (WIP):** the paradigm split isn't meant to declare a "winner" outright — CAS is not simply "NCE plus more math." The two paradigms make genuinely different claims about *why* markets behave the way they do (equilibrium-seeking rational agents vs. adaptive heterogeneous agents generating emergent structure), and our Inference steps are designed to let the data adjudicate which claim better describes `UAL`, `PAL`, `BTC-USD`, and `SPY` over 2020-2025, and whether the answer differs by asset.

# Price Movement

```{r exec-price-movement, echo=TRUE, child='06_1_asset_movement.rmd'}
```

Having established what "return" means for each asset in isolation - and whether that return process shows evidence of memory, fat tails, or regime-dependent scaling - we now widen the lens from single-asset behavior to *joint* behavior. If Price Movement finds that an asset's own variance does not scale the way NCE predicts, the natural next question is whether that same non-equilibrium behavior shows up in how assets move *together*, particularly under stress.

# Risk and Correlation

```{r exec-risk-correlation, echo=TRUE, child='06_2_risk_and_correlation.rmd'}
```

With cross-asset relationships characterized - including whether diversification benefits hold up or collapse during high-VIX regimes - we widen the lens once more, from asset-to-asset structure to the macroeconomic backdrop shared by all four assets. If correlations concentrate during stress, the next question is what macro conditions (rate levels, curve shape, inflation regime) are driving that stress in the first place, and whether a single discount rate is even an adequate stand-in for "the cost of money" across `UAL`, `PAL`, and `BTC-USD` simultaneously.

# Macro Baseline

```{r exec-macro-baseline, echo=TRUE, child='06_3_macro_baseline.rmd'}
```

# Conclusion

*(TBD - synthesize the Price Movement, Risk and Correlation, and Macro Baseline verdicts into a single statement of which paradigm - NCE, CAS, or a regime-dependent mix of both - best characterizes `UAL`, `PAL`, `BTC-USD`, and `SPY` over the sample period, and carry that conclusion into the Section 07 investment allocation.)*

