# Financial Investigation

In this section, we aim to model and investigate the relationship of these assets to the wide economy. The order of the analysis: Price Movement -> Risk and Correlation -> Macro Baseline.

Each subsection is a wider lens that the previous and is meant from zooming in the economy in daily life to zooming out to the wider picture. Here, different sections ask fundamental questions:

 1. **Price Movement** - ask a question of one one asset at a time.
 2. **Risk and Correlation** - ask about the relationship between assets.
 3. **Macro Baseline** - ask about the backdrop of all those assets.

This is to bring a meaningful scope from the previous subsection. We can't sensibly ask "how does UAL's correlation with SPY behave under stress" before you've established what "return" even means for UAL iself; meanwhile, you can't meaningfully frame "what regime are we in" until you've already seen regime-dependent behavior show up in the correlation data.

This is also going by date frequency as price movement is all about fast-moving and granular data while the Macro Baseline is slowing-moving and very contextual.

Moreover, for each subsection, we aim to analyze in a more structured manner and that is by:

 1. **Structure** - what is the raw state (e.g. log-adjusted return assets) to be analyzed, what is the enriched state, how do we compress this further?
 2. **Dynamics** - defines the transition of each state, viewing either as conservative or dissaptive system, and what context from the transitions can be grouped together?
 3. **Inference** - gather evidence w/ statistics from inferred behavior, create the decision rule, and create a verdict from them. 
 4. **Synthesis** - from the three, and the two theoretical paradigms of each, conclude what can be said.

Finally, for each we are going to use two theoretical paradigms to see how lenses view the market:

 1. **Neoclassical Economics (NCE)** - Classic financial concepts that treats the financial system as a linear, rational, mechanical engine that naturally tends toward a stable equilibrium.
 2. **Complex Adaptive Systems (CAS)** - This comes from the field of Econophysics. It views markets not as sterile machines, but as physical ecosystems where feedback loops, memory, friction, and shifting participant behavior constantly warp the environment out of equilibrium. 

Of course, we will state Objective, Code (SQL/dplyr), Output, Financial Interpretation. 

# Price Movement

```{r exec-macro-baseline, echo=TRUE, child='06.1_asset_movement.rmd'}
```

# Risk and Correlation

```{r exec-macro-baseline, echo=TRUE, child='06.2_risk_and_correlation.rmd'}
```


# Macro Baseline

```{r exec-macro-baseline, echo=TRUE, child='06.3_macro_baseline.rmd'}
```

# Conclusion

# References:

- [1] Bouchaud, J. P., Farmer, J. D., & Lillo, F. (2009). "How Markets Slowly Digest Large Trading Orders." Handbook of Financial Markets: Dynamics and Evolution, 57–160.
- [2] Fama, E. F. (1970). "Efficient Capital Markets: A Review of Theory and Empirical Work." The Journal of Finance, 25(2), 383–417.
- [3] Grinold, R. C. (1989). "The Fundamental Law of Active Management." The Journal of Portfolio Management, 15(3), 30–38.
- [4] Litterman, R., & Scheinkman, J. (1991). "Common Factors Affecting Bond Returns." The Journal of Fixed Income, 1(1), 54–61.
- [5] Mandelbrot, B. (1963). "The Variation of Certain Speculative Prices." The Journal of Business, 36(4), 394–419.
- [6] Markowitz, H. (1952). "Portfolio Selection." The Journal of Finance, 7(1), 77–91.
- [7] Sharpe, W. F. (1964). "Capital Asset Prices: A Theory of Market Equilibrium under Conditions of Risk." The Journal of Finance, 19(3), 425–442.
- [8] jetbundle. (2026). the black book of quantitative finance. https://www.youtube.com/watch?v=AKIkhuR7HYY
- [9] jetbundle. (2026). 6 laws behind modern quantitative finance. https://www.youtube.com/watch?v=cvL_uWtMTAo

