# Financial Investigation

In this section, we aim to model and investigate the relationship of these assets to the wide economy. The order of the analysis: Price Movement -> Risk and Correlation -> Macro Baseline.

Each subsection is a wider lens that the previous and is meant from zooming in the economy in daily life to zooming out to the wider picture. Here, different sections ask fundamental questions:

 1. **Price Movement** - ask a question of one one asset at a time.
 2. **Risk and Correlation** - ask about the relationship between assets.
 3. **Macro Baseline** - ask about the backdrop of all those assets.

This is to bring a meaningful scope from the previous subsection. We can't sensibly ask "how does UAL's correlation with SPY behave under stress" before you've established what "return" even means for UAL iself; meanwhile, you can't meaningfully frame "what regime are we in" until you've already seen regime-dependent behavior show up in the correlation data.

This is also going by date frequency as price movement is all about fast-moving and granular data while the Macro Baseline is slowing-moving and very contextual.

Moreover, for each subsection, we aim to analyze in a more structured manner and that is by:

 1. **Structure** - what is the raw state (e.g. log-adjusted return assets) to be analyzed, what is the enriched state, how do we compress this further?
 2. **Dynamics** - defines the transition of each state, viewing either as conservative or dissaptive system, and what context from the transitions can be grouped together?
 3. **Inference** - gather evidence w/ statistics from inferred behavior, create the decision rule, and create a verdict from them. 
 4. **Synthesis** - from the three, and the two theoretical paradigms of each, conclude what can be said.

Finally, for each we are going to use two theoretical paradigms to see how lenses view the market:

 1. **Neoclassical Economics (NCE)** - Classic financial concepts that treats the financial system as a linear, rational, mechanical engine that naturally tends toward a stable equilibrium.
 2. **Complex Adaptive Systems (CAS)** - It views markets not as sterile machines, but as biological ecosystems where feedback loops, friction, and shifting participant behavior constantly warp the environment out of equilibrium.

Of course, we will state Objective, Code (SQL/dplyr), Output, Financial Interpretation. 

# Price Movement

Here, we view how each asset moves in the economies they are in.

## Structure

**NCE**: These treat asset as a singular scalar, closing price or log return of it, assumed to be the object of interest.

 1. Raw Input: Adjusted Closing Price on `UAL`, `PAL`, `BTC-USD`, and `SPY`.
 2. Enriched State: Log Return.
 3. Compression: Rolling Window of mean, variance, skewness, and kurtosis.

Here, we are using Fama's (1970) Efficient Market Hypothesis (EMH) which states that prices fully reflect all available information [2]. Because information arrives as unpredictable, exogenous shocks, price updates are independent and identically distributed ($i.i.d.$). This reduces the asset's structural state to a trivial, memoryless coordinate point on a flat Euclidean space.

**Objective**:

**Code**:
```{r price-structure-df-nce}

price_structure_df <- master_df %>%
    select(Date, UAL.LogAdj, PAL.LogAdj, `BTC-USD.LogAdj`, SPY.LogAdj) %>%
    mutate(
    # UAL Rolling Moments
    UAL_roll_mean = rollapplyr(UAL.LogAdj, 21, mean, fill = NA),
    UAL_roll_var  = rollapplyr(UAL.LogAdj, 21, var, fill = NA),
    UAL_roll_skew = rollapplyr(UAL.LogAdj, 21, skewness, fill = NA),
    UAL_roll_kurt = rollapplyr(UAL.LogAdj, 21, kurtosis, fill = NA),
    
    # Repeat the identical rolling pattern for PAL
    PAL_roll_mean = rollapplyr(PAL.LogAdj, 21, mean, fill = NA),
    PAL_roll_var  = rollapplyr(PAL.LogAdj, 21, var, fill = NA),
    PAL_roll_skew = rollapplyr(PAL.LogAdj, 21, skewness, fill = NA),
    PAL_roll_kurt = rollapplyr(PAL.LogAdj, 21, kurtosis, fill = NA),
    
    # Repeat for BTC-USD
    BTC_roll_mean = rollapplyr(`BTC-USD.LogAdj`, 21, mean, fill = NA),
    BTC_roll_var  = rollapplyr(`BTC-USD.LogAdj`, 21, var, fill = NA),
    BTC_roll_skew = rollapplyr(`BTC-USD.LogAdj`, 21, skewness, fill = NA),
    BTC_roll_kurt = rollapplyr(`BTC-USD.LogAdj`, 21, kurtosis, fill = NA),
    
    # Repeat for SPY
    SPY_roll_mean = rollapplyr(SPY.LogAdj, 21, mean, fill = NA),
    SPY_roll_var  = rollapplyr(SPY.LogAdj, 21, var, fill = NA),
    SPY_roll_skew = rollapplyr(SPY.LogAdj, 21, skewness, fill = NA),
    SPY_roll_kurt = rollapplyr(SPY.LogAdj, 21, kurtosis, fill = NA)
  ) %>%
  # Remove rows that don't have enough history for a full 21-day window
  filter(!is.na(UAL_roll_mean))

print(price_structure_df)

plot_df <- price_structure_df %>%
  select(Date, PAL_roll_mean, PAL_roll_var, PAL_roll_skew, PAL_roll_kurt) %>%
  pivot_longer(
    cols = -Date, 
    names_to = "Metric", 
    values_to = "Value"
  )

p_structure <- ggplot(plot_df, aes(x = Date, y = Value, color = Metric)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~Metric, scales = "free_y", ncol = 1) +
  theme_minimal() +
  labs(title = "NCE Structural State: Rolling Moments (21-Day)",
       y = "Moment Value")
  
ggplotly(p_structure)
```

**Output**:

**Financial Interpretation**:

**CAS**: Treats the shape of the return distribution itself is the object of interest.
 
 1. Raw Input: Adjusted Closing Price on `UAL`, `PAL`, `BTC-USD`, and `SPY`.
 2. Enriched State: Jet Bundle Lift of Low Return, with its Velocity, and Curvature.
 3. Compression: Rolling Window, of 21 days (Days in a month for financial trading), of its mean, variance, skewness, and kurtosis.



This follows from Mandelbrot's work [5], which states that financial returns are non-Gaussian and follow a stable Paretian distribution where variance is infinite or non-convergent over local windows. By lifting the returns into a Jet Bundle (position, velocity, curvature), the system recognizes that two assets sharing identical log returns can be on entirely different paths depending on their structural velocity (momentum) and curvature (volatility acceleration).


**Objective**:

**Code**:

**Output**:

**Financial Interpretation**:

```{r price-structure-df-cas}
# 1. Identify all active log return columns in your master dataframe
logadj_cols <- names(master_df)[str_detect(names(master_df), "\\.LogAdj$")]

# 2. Compute the Jet Bundle Lift (Position, Velocity, Curvature)
jet_bundle_df <- master_df %>%
  select(Date, all_of(logadj_cols)) %>%
  mutate(
    # --- STEP A: CALCULATE VELOCITY (First Derivative: x_t - x_{t-1}) ---
    # This measures daily return momentum / directional shifts
    across(
      all_of(logadj_cols),
      ~ .x - lag(.x),
      .names = "{str_remove(.col, '\\\\.LogAdj')}.Velocity"
    ),
    
    # --- STEP B: CALCULATE CURVATURE (Second Derivative: dx_t - dx_{t-1}) ---
    # This measures volatility acceleration or momentum decay
    across(
      all_of(logadj_cols),
      ~ (.x - lag(.x)) - (lag(.x) - lag(.x, 2)),
      .names = "{str_remove(.col, '\\\\.LogAdj')}.Curvature"
    )
  )

# 3. Rename the original log return columns to explicitly serve as "Position" Coordinates
jet_bundle_df <- jet_bundle_df %>%
  rename_with(~ str_replace(.x, "\\.LogAdj$", ".Position"), all_of(logadj_cols))

print(jet_bundle_df)

# Select a target asset to compress (e.g., UAL)
target_prefix <- "UAL"

compressed_cas_df <- jet_bundle_df %>%
  select(Date, starts_with(target_prefix)) %>%
  mutate(
    # Compress Position Space
    UAL_pos_mean = rollapplyr(UAL.Position, 21, mean, fill = NA),
    UAL_pos_var  = rollapplyr(UAL.Position, 21, var, fill = NA),
    UAL_pos_skew = rollapplyr(UAL.Position, 21, skewness, fill = NA),
    
    # Compress Velocity Space (Momentum Stability)
    UAL_vel_mean = rollapplyr(UAL.Velocity, 21, mean, fill = NA),
    UAL_vel_var  = rollapplyr(UAL.Velocity, 21, var, fill = NA),
    UAL_vel_skew = rollapplyr(UAL.Velocity, 21, skewness, fill = NA),
    
    # Compress Curvature Space (Volatility Acceleration/Shocks)
    UAL_cur_mean = rollapplyr(UAL.Curvature, 21, mean, fill = NA),
    UAL_cur_var  = rollapplyr(UAL.Curvature, 21, var, fill = NA),
    UAL_cur_skew = rollapplyr(UAL.Curvature, 21, skewness, fill = NA)
  ) %>%
  # Drop early dates that do not possess a full 21-day historical window
  filter(!is.na(UAL_pos_mean))

# 1. Pivot all metrics into a long format together
cas_plot_long <- compressed_cas_df %>%
  select(Date, all_of(list_cas)) %>%
  pivot_longer(
    cols = -Date,
    names_to = "Full_Metric",
    values_to = "Value"
  ) %>%
  # Break out the metadata to create smart facet labels
  mutate(
    Space = case_when(
      str_detect(Full_Metric, "_pos_") ~ "Position Space (Yield Baseline)",
      str_detect(Full_Metric, "_vel_") ~ "Velocity Space (Momentum/Flow)",
      str_detect(Full_Metric, "_cur_") ~ "Curvature Space (Volatility Acceleration)"
    ),
    Stat = case_when(
      str_detect(Full_Metric, "_mean$") ~ "21d Rolling Mean",
      str_detect(Full_Metric, "_var$") ~ "21d Rolling Variance",
      str_detect(Full_Metric, "_skew$") ~ "21d Rolling Skewness"
    )
  )

# 2. Loop through each Space tier to create 3 consolidated interactive summaries
for (current_space in unique(cas_plot_long$Space)) {
  
  space_data <- cas_plot_long %>% filter(Space == current_space)
  
  p_space <- ggplot(space_data, aes(x = Date, y = Value, color = Stat)) +
    geom_line(alpha = 0.7) +
    facet_wrap(~Stat, scales = "free_y", ncol = 1) +
    theme_minimal() +
    labs(
      title = paste("CAS Structural State Space:", current_space),
      x = "Date",
      y = "Value"
    ) +
    theme(legend.position = "none")
  
  # Explicitly print the interactive widget inside the loop
  print(ggplotly(p_space))
}
```


## Dynamics

**NCE**: Returns are draws from a fixed Gaussian, so risk scales predictably with thr square root of time.
 
 1. Transition (Conservative): Modeled via Black-Scholes Diffusion where price evolution is a continuous random walk driven by standard Brownian motion ($dW_t$).
 2. Grouped Context of Transitions: Portfolio risk across multiple time horizons ($T$) aggregates linearly according to the standard square-root-of-time scaling rule: $\sigma_T = \sigma_1 \times \sqrt{T}$. The temporal progression contains zero memory or feedback loops.
 
 
```{r rolling-dynamic-plot}
# --- DYNAMICS STEP ---
# 1. Compute empirical variance for each horizon (uncorrected by T)
empirical_summary <- price_dynamics_df %>%
  summarise(
    `1-Day`  = var(PAL.LogAdj, na.rm = TRUE),
    `5-Day`  = var(PAL_5d_return, na.rm = TRUE),
    `21-Day` = var(PAL_21d_return, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Horizon", values_to = "Empirical_Variance")

# 2. Extract the base 1-day baseline variance (our variance unit scale sigma^2)
base_var <- empirical_summary %>% 
  filter(Horizon == "1-Day") %>% 
  pull(Empirical_Variance)

# 3. Inject the NCE Black-Scholes Linear Grouped Progression Scale
scaling_df <- empirical_summary %>%
  mutate(
    # Map Horizon strings to their integer T parameters
    T_days = case_when(
      Horizon == "1-Day"  ~ 1,
      Horizon == "5-Day"  ~ 5,
      Horizon == "21-Day" ~ 21
    ),
    # NCE Definition: Variance scales linearly with time (sigma_1^2 * T)
    NCE_Theoretical_Variance = base_var * T_days
  ) %>%
  # Pivot to long format so ggplot can create separate color layers
  pivot_longer(
    cols = c(Empirical_Variance, NCE_Theoretical_Variance), 
    names_to = "Type", 
    values_to = "VarianceValue"
  ) %>%
  mutate(
    Type = case_when(
      Type == "Empirical_Variance"         ~ "Empirical (Actual Data)",
      Type == "NCE_Theoretical_Variance"   ~ "NCE: Black-Scholes Theoretical Line"
    )
  )

# 4. Generate the Visualization highlighting the grouping context deviation
p_dynamics <- ggplot(scaling_df, aes(x = factor(T_days), y = VarianceValue, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("steelblue", "darkred")) +
  theme_minimal() +
  labs(
    title = "PAL Dynamics: Empirical vs. Black-Scholes Linear Risk Scaling",
    subtitle = "NCE requires empirical columns to exactly match theoretical anchors",
    x = "Time Horizon (Days)",
    y = "Total Return Variance",
    fill = "Framework Layer"
  )

ggplotly(p_dynamics)
```
 
**CAS**: Variance scaling across horizons deviates from that straight line — evidence of memory, clustering, or regime persistence rather than pure noise.
 
 1. Transition (Diffusion): Modeled as a continuous Flow on a Manifold splitting into a Symplectic core (conservative supply/demand matching) and Contact geometry (dissipative, thermodynamic features such as liquidation cascades and market-maker risk-premium bleed).
 2. Grouped Context of Transitions: Follows a multi-scale variance scaling rule: $\text{Var}(r_{t+T}) \propto T^{2H}$, where the Hurst Exponent ($H \neq 0.5$) captures temporal memory. Transitions group into persistent regimes where high-volatility events cluster endogenously (Mandelbrot 1963).

```{r price-movement-cas-dynamics}

cas_dynamics_df <- jet_bundle_df %>%
  select(Date, PAL.Position, PAL.Velocity, PAL.Curvature) %>%
  mutate(
    # Position multi-horizon paths
    PAL_pos_5d  = rollapplyr(PAL.Position, 5, sum, fill = NA),
    PAL_pos_21d = rollapplyr(PAL.Position, 21, sum, fill = NA),
    
    # Velocity multi-horizon paths
    PAL_vel_5d  = rollapplyr(PAL.Velocity, 5, sum, fill = NA),
    PAL_vel_21d = rollapplyr(PAL.Velocity, 21, sum, fill = NA),
    
    # Curvature multi-horizon paths
    PAL_cur_5d  = rollapplyr(PAL.Curvature, 5, sum, fill = NA),
    PAL_cur_21d = rollapplyr(PAL.Curvature, 21, sum, fill = NA)
  )

# --- CONSOLIDATE AND GENERATE SCALING MATRIX ---
# Gather empirical variances across every coordinate space tier
cas_empirical <- cas_dynamics_df %>%
  summarise(
    # Position Horizons
    pos_1d  = var(PAL.Position, na.rm = TRUE),
    pos_5d  = var(PAL_pos_5d, na.rm = TRUE),
    pos_21d = var(PAL_pos_21d, na.rm = TRUE),
    
    # Velocity Horizons
    vel_1d  = var(PAL.Velocity, na.rm = TRUE),
    vel_5d  = var(PAL_vel_5d, na.rm = TRUE),
    vel_21d = var(PAL_vel_21d, na.rm = TRUE),
    
    # Curvature Horizons
    cur_1d  = var(PAL.Curvature, na.rm = TRUE),
    cur_5d  = var(PAL_cur_5d, na.rm = TRUE),
    cur_21d = var(PAL_cur_21d, na.rm = TRUE)
  ) %>%
  pivot_longer(everything(), names_to = "Key", values_to = "Empirical_Variance") %>%
  separate(Key, into = c("Space", "Horizon"), sep = "_")

# Map integer T scales and baseline parameters to compute NCE theoretical projections
cas_scaling_df <- cas_empirical %>%
  mutate(
    T_days = case_when(
      Horizon == "1d"  ~ 1,
      Horizon == "5d"  ~ 5,
      Horizon == "21d" ~ 21
    ),
    Space_Label = case_when(
      Space == "pos" ~ "Position Space (Returns)",
      Space == "vel" ~ "Velocity Space (Momentum)",
      Space == "cur" ~ "Curvature Space (Acceleration)"
    )
  ) %>%
  # Dynamically fetch the 1-Day baseline variance for each respective space tier
  group_by(Space) %>%
  mutate(
    NCE_Theoretical_Variance = Empirical_Variance[Horizon == "1d"] * T_days
  ) %>%
  ungroup() %>%
  # Reshape for multi-layered side-by-side column mapping
  pivot_longer(
    cols = c(Empirical_Variance, NCE_Theoretical_Variance),
    names_to = "Framework_Type",
    values_to = "VarianceValue"
  ) %>%
  mutate(
    Framework_Type = case_when(
      Framework_Type == "Empirical_Variance"       ~ "Empirical (Actual CAS Dynamics)",
      Framework_Type == "NCE_Theoretical_Variance" ~ "NCE Benchmark (Linear Scaling)"
    )
  )

# --- PLOT CAS DYNAMICS SEPARATELY BY SPACE TIER ---
for (current_space in unique(cas_scaling_df$Space_Label)) {
  
  space_dynamics_data <- cas_scaling_df %>% filter(Space_Label == current_space)
  
  p_cas_dynamics <- ggplot(space_dynamics_data, aes(x = factor(T_days), y = VarianceValue, fill = Framework_Type)) +
    geom_col(position = "dodge", alpha = 0.8) +
    scale_fill_manual(values = c("darkblue", "darkgrey")) +
    theme_minimal() +
    labs(
      title = paste("CAS Dynamics Scaling:", current_space),
      subtitle = "Significant differences between actual and benchmark indicate memory structures",
      x = "Time Horizon (Days)",
      y = "Total Coordinate Variance",
      fill = "Model Hypothesis"
    )
  
  print(ggplotly(p_cas_dynamics))
}
```

```{r}
# 1. Isolate the base 1-Day empirical variance for each Space tier to prevent recycling errors
base_variance_map <- cas_scaling_df %>%
  filter(Framework_Type == "Empirical (Actual CAS Dynamics)", Horizon == "1d") %>%
  select(Space, Base_Var = VarianceValue)

# 2. Join the base variance map and the estimated Hurst exponents back to the scaling dataset
plotting_matrix_df <- cas_scaling_df %>%
  left_join(base_variance_map, by = "Space") %>%
  left_join(
    cas_true_dynamics_df %>% select(Space, Hurst_Est) %>% unique(),
    by = "Space"
  )

# 3. Create the distinct CAS Fractional scaling curves using your exact formula: Var_1d * T^(2H)
cas_lines_df <- plotting_matrix_df %>%
  filter(Framework_Type == "NCE Benchmark (Linear Scaling)") %>%
  mutate(
    Framework_Type = "CAS Fractional Target (T^{2H})",
    # Override NCE linear scaling with Mandelbrot's power law profile
    VarianceValue = Base_Var * (T_days^(2 * Hurst_Est))
  )

# 4. Bind the new fractional curve back into your master visual comparison matrix
plotting_matrix_df <- bind_rows(cas_scaling_df, cas_lines_df) %>%
  left_join(
    unique(cas_true_dynamics_df[, c("Space_Label", "Hurst_Est")]), 
    by = "Space_Label", 
    suffix = c("", ".y")
  ) %>%
  # Fill missing Hurst_Est text descriptors generated by the bind_rows step
  group_by(Space) %>%
  fill(Hurst_Est, Space_Label, .direction = "updown") %>%
  ungroup()

# --- PLOT THE THREE-WAY PARADIGM COMPARISON ---
for (current_space in unique(plotting_matrix_df$Space_Label)) {
  
  space_data <- plotting_matrix_df %>% filter(Space_Label == current_space)
  h_val <- round(unique(space_data$Hurst_Est), 3)
  
  p_cas_true <- ggplot(space_data, aes(x = factor(T_days), y = VarianceValue, fill = Framework_Type)) +
    geom_col(position = "dodge", alpha = 0.85) +
    scale_fill_manual(values = c("darkblue", "darkorange", "darkgrey")) +
    theme_minimal() +
    labs(
      title = paste("True CAS Dynamics Scaling:", current_space),
      subtitle = paste0("Empirical Memory Profile vs. Linear Benchmarks (Estimated Hurst H = ", h_val, ")"),
      x = "Time Horizon (Days)",
      y = "Total Coordinate Variance",
      fill = "Model Class"
    )
  
  print(ggplotly(p_cas_true))
}
```

## Inference

**NCE**: Checks if short-horizon volatility is sufficient to price any longer-horizon risk. 
 
 1. Gathering Evidence: Goodness of Fit test (Jarque-Bera) for normality, and the Lo-MacKinlay Variance Ratio Test ($VR(k)$) to evaluate variance scaling across multiple horizons ($k$).
 2. Decision Rule: 
    - **IF** Jarque-Bera fails to reject the null hypothesis of normality ($p > 0.05$) **AND** $VR(k) \approx 1$ across all designated horizons ($k$):
      - Then we accept the NCE framework. Daily volatility is sufficient to price longer-horizon risk, confirming a flat, memoryless Euclidean price space.
    - **IF** Jarque-Bera rejects normality ($p \le 0.05$) **OR** $VR(k)$ significantly drifts away from $1$ as $k$ increases:
      - Then the NCE assumption fails. Short-horizon volatility calculations systematically misprice or understate long-horizon structural risk.
 3. Verdict: `TBD`

```{r inference-}

# --- INFERENCE STEP ---
# Pick an asset to test dynamically (or loop through them all)
target_asset <- "PAL"

# Gather vectors from your completed dynamics/structure frames
daily_returns  <- na.omit(price_dynamics_df[[paste0(target_asset, ".LogAdj")]])
ret_5d         <- na.omit(price_dynamics_df[[paste0(target_asset, "_5d_return")]])
ret_21d        <- na.omit(price_dynamics_df[[paste0(target_asset, "_21d_return")]])

# Execute NCE Significance Tests
jb_test   <- jarque.bera.test(daily_returns)
p_val_jb  <- jb_test$p.value

# Form the Variance Ratios to check against 1
vr_5d  <- var(ret_5d) / (5 * var(daily_returns))
vr_21d <- var(ret_21d) / (21 * var(daily_returns))

# Execute CAS Significance Tests
# Note:Hurst Exponents via Rescaled Range, and Hill Estimators for heavy tails.
hurst_val <- pracma::hurstexp(daily_returns, display = FALSE)$Hs
# tail_index <- custom_hill_estimator(daily_returns) 

# 4. Print Outputs to clear the TBD fields
cat("--- INFERENCE REPORT FOR:", target_asset, "---\n")
cat("Jarque-Bera p-value: ", p_val_jb, "\n")
cat("Variance Ratio (5-Day Target = 1): ", vr_5d, "\n")
cat("Variance Ratio (21-Day Target = 1): ", vr_21d, "\n")
cat("Hurst Exponent (Target = 0.5): ", hurst_val, "\n")
```


**CAS**: Check if that the tail risk is systematically underpriced by that approach, because rare, large moves happen more often than a Gaussian model would predict.
 
 1. Evidence:
   - Hurst Exponent ($H$) via Rescaled Range Analysis to evaluate long-range dependency and clustering ($H \neq 0.5$).
   - Hill’s Estimator for Tail Index ($\alpha$) to compute the power-law decay exponent of the return distribution's extreme tails.
 2. Decision Rule:
    - **IF** the estimated Hurst exponent deviates significantly from equilibrium ($H > 0.5$ for persistence or $H < 0.5$ for anti-persistence) **AND** Hill's tail index indicates heavy tails ($\alpha < 3$, where variance or kurtosis begins to fail standard convergence limits):
      - Then we validate the CAS paradigm. The asset demonstrates structural memory and fat tails, proving that return distributions cannot be compressed into a memoryless scalar without ignoring catastrophic systemic tail risk.
    - **IF** $H \approx 0.5$ **AND** $\alpha \ge 3$ (approaching Gaussian tail decay thresholds):
      - Then the asset behaves like a standard equilibrium system. Deviations are transient noise, and the traditional random walk approximation remains robust.
 3. Verdict: `TBD`
 
```{r cas-inference-engine, echo=TRUE, message=FALSE, warning=FALSE}
# ==============================================================================
# CAS INFERENCE ENGINE: EVIDENCE, DECISION RULE, & PROGRAMMATIC VERDICT
# ==============================================================================

# 1. Isolate target asset log returns (e.g., PAL) and strip NA values
returns_vector <- master_df %>%
  filter(!is.na(PAL.LogAdj)) %>%
  pull(PAL.LogAdj)

# ------------------------------------------------------------------------------
# STEP 1: GATHERING EVIDENCE (STATISTICAL COMPUTATION)
# ------------------------------------------------------------------------------

### A. Hurst Exponent (H) via Log-Variance Multi-Scale Regression
# We compute variance across horizons 1 to 21 to capture the exact memory decay profile
horizons <- 1:21
horizon_variances <- sapply(horizons, function(k) {
  # Generate k-day rolling compounded paths
  k_day_returns <- zoo::rollapplyr(returns_vector, k, sum, fill = NA)
  var(k_day_returns, na.rm = TRUE)
})

# Log-linear fit: ln(Var) = ln(C) + 2H*ln(T)
log_T <- log(horizons)
log_Var <- log(horizon_variances)
fit_hurst <- lm(log_Var ~ log_T)
estimated_H <- as.numeric(coef(fit_hurst)[2] / 2)


### B. Hill's Estimator for Tail Index (alpha)
# We isolate the upper extreme tail (top 5% positive tail shocks) to extract the power law decay
tail_threshold <- 0.05 
sorted_returns <- sort(returns_vector, decreasing = TRUE)
n_tail <- floor(length(sorted_returns) * tail_threshold)

# Extract extreme threshold anchor
x_threshold <- sorted_returns[n_tail + 1]
tail_shocks <- sorted_returns[1:n_tail]

# Hill Estimator Formula: 1 / [ (1/k) * Sum( ln(x_i / x_{k+1}) ) ]
hill_alpha <- 1 / (mean(log(tail_shocks / x_threshold)))

# Print Computed Evidence Summary Table
cat("===========================================\n")
cat("      CAS EVIDENCE GATHERING MATRIX        \n")
cat("===========================================\n")
cat("Target Asset Component: PAL\n")
cat("Estimated Hurst Exponent (H) : ", round(estimated_H, 4), "\n")
cat("Hill's Tail Decay Index (α)  : ", round(hill_alpha, 4), "\n")
cat("-------------------------------------------\n\n")


# ------------------------------------------------------------------------------
# STEP 2 & 3: APPLICATION OF DECISION RULE & VERDICT SPECIFICATION
# ------------------------------------------------------------------------------

# Establish strict structural thresholds based on your framework rules
hurst_lower_bound <- 0.45
hurst_upper_bound <- 0.55
alpha_fat_tail_threshold <- 3.0

if ((estimated_H > hurst_upper_bound || estimated_H < hurst_lower_bound) && hill_alpha < alpha_fat_tail_threshold) {
  
  # --- BRANCH 1: VALIDATE COMPLEX ADAPTIVE SYSTEMS PARADIGM ---
  cas_verdict <- "VALIDATED (REJECT NEOCLASSICAL EQUILIBRIUM)"
  
  cas_interpretation <- paste0(
    "VERDICT ANALYSIS: The CAS paradigm is empirically confirmed.\n",
    "Reasoning: The asset demonstrates strong anomalous scaling (H = ", round(estimated_H, 3), 
    if_else(estimated_H < 0.5, " indicating anti-persistent mean-reversion", " indicating persistent trend-reinforcement"), 
    ") coupled with extreme fat-tailed decay (α = ", round(hill_alpha, 3), ").\n",
    "Systemic Impact: Because α < 3, the traditional Gaussian variance boundaries are violated.\n",
    "Compressing this asset into standard local variance metrics or flat i.i.d assumptions will systematically\n",
    "underprice options structures and risk profiles, rendering standard Black-Scholes risk models blind\n",
    "to endogenous, structural liquidation cascades."
  )
  
} else {
  
  # --- BRANCH 2: VALIDATE NEOCLASSICAL EQUILIBRIUM ---
  cas_verdict <- "REJECTED (NEOCLASSICAL RANDOM WALK ROBUST)"
  
  cas_interpretation <- paste0(
    "VERDICT ANALYSIS: The Neoclassical Equilibrium framework holds robustly.\n",
    "Reasoning: The asset's scaling footprint closely approximates memoryless Brownian motion (H = ", round(estimated_H, 3), 
    ") and the distribution's tail profile safely matches or exceeds normal exponential decay structures (α = ", round(hill_alpha, 3), ").\n",
    "Systemic Impact: Shocks are transient, exogenous, and uncorrelated. Traditional Value-at-Risk (VaR)\n",
    "and square-root-of-time scaling mechanics provide a dependable framework for structural risk budgeting."
  )
}

# Output programmatically derived Verdict and Detailed Summary
cat("======================================================================\n")
cat("FINAL PROGRAMMATIC INFERENCE VERDICT:", cas_verdict, "\n")
cat("======================================================================\n")
cat(cas_interpretation, "\n")
```
 
## Synthesis
The Price Movement analysis serves as our most granular lens, examining individual asset vectors in isolation. If the empirical verdicts yield deviations from normality and square-root scaling, it proves that "price" is not a simple, self-contained coordinate point. Instead, assets contain intrinsic path dependency and memory, prompting us to step outward into our next lens: how these unique asset states interact with one another under varying regimes.

# Risk and Correlation

Here, we view how the relationships of the assets moves in the economies they are in.

## Structure

**NCE**: Single, time-invariant covariance matrix — one number per asset pair, good for all seasons. 
 
 1. **Raw Input**: Co-moving daily log returns matrix for UAL, PAL, BTC-USD, and SPY.
 2. **Enriched State**: A static sample covariance matrix ($\Sigma$) calculated across the entire dataset.
 3. **Compressio**n: A single cross-asset correlation scalar ($\rho_{i,j}$) per asset pair.
 
 Here, we utilize Markowitz’s Modern Portfolio Theory (MPT) and the foundations of Sharpe’s Capital Asset Pricing Model (CAPM). Under this architecture, the asset space is a static Euclidean vector space where asset relationships are treated as fixed, linear geometries. Risk is fully quantified by the variance-covariance matrix ($\Sigma$), assuming structural errors or shifts are merely random, zero-mean Gaussian noise around a permanent equilibrium.

**CAS**: CAS's state is that same matrix conditioned on regime — a High-VIX matrix and a Low-VIX matrix, treated as genuinely different objects, not noise around one "true" matrix.
 
 1. Raw Input: Daily log returns matrix for UAL, PAL, BTC-USD, SPY, and the daily conditioning regime marker VIX.
 2. Enriched State: Conditional, rolling covariance matrices $\Sigma_t \vert{} \text{VIX}_t$, explicitly split into separate geometric charts.
 3. Compression: Partitioned states mapped into distinct manifolds: Low-VIX Chart ($\text{VIX} \le 25\text{th percentile}$) and High-VIX Chart ($\text{VIX} \ge 75\text{th percentile}$).

This builds on Bouchaud’s Law of Endogenous Correlation Breakdown, which proves that cross-asset dependencies are non-stationary, state-dependent, and governed by systemic feedback loops. Rather than viewing correlation shifts as statistical noise, the system recognizes them as distinct geometric structures. High-volatility matrices represent entirely separate topological charts on the market manifold that cannot be smoothly mapped using a single, global Euclidean average.

## Dynamics

**NCE**: Assumes correlations are stationary — the relationship between UAL and SPY today is the same relationship in a crash.
 
 1. Transition (Conservative): Portfolio variance transitions over time via a deterministic, linear law: $\sigma^2_p = w^T \Sigma w$. The weights $w$ change, but the inner product structure governed by $\Sigma$ is treated as a conservation law.
 2. Grouped Context of Transitions: Multi-asset cross-dynamics are modeled as permanently coupled geometric Brownian motions. Co-movements scale symmetrically across all economic cycles, treating structural changes as impossible.

**CAS**: Correlations are regime-dependent and tend to converge upward precisely during stress (diversification "disappears" when you need it most).
 
 1. Transition (Diffusion): Systemic risk behaves like a thermodynamic fluid following Contact Geometry / Non-equilibrium Dissipation. In calm regimes, the flow is Symplectic (information-conserving, asset pairs preserve their unique orthogonal vectors). When a threshold is crossed (e.g., a high-VIX shock), margin calls and liquidations force a massive energy transfer (entropy bleed).
 2. Grouped Context of Transitions: Asset transitions are grouped into distinct market volatility regimes. Under high-volatility environments (conditioned by a high `VIX` threshold), individual asset behaviors lose their unique degrees of freedom, and correlations across completely different asset classes (e.g., `UAL`, `PAL`, and `BTC-USD`) shift simultaneously. 

## Inference

**NCE**: The Neoclassical paradigm infers that because asset relations are governed by a stationary covariance matrix, a diversified portfolio’s risk reduction remains stable and can be reliably trusted for long-term strategic allocation and value-at-risk (VaR) planning.
 
 1. Gathering Evidence: Rolling Correlation Homogeneity Test (such as Jennrich’s test for covariance matrix equality or a rolling distance evaluation) comparing the structural covariance matrix of a Low-VIX sub-sample against a High-VIX sub-sample ($H_0: \Sigma_{\text{Low}} = \Sigma_{\text{High}}$).
 2. Decision Rule:
    - **IF** the test fails to reject the null hypothesis of covariance homogeneity ($p > 0.05$):
      - Then NCE is verified. The asset covariance structure is stationary over time, and a single historical covariance matrix is sufficient to build dependable, permanently diversified portfolios.
    - **IF** the test strongly rejects covariance homogeneity ($p \le 0.05$):
      - Then the NCE assumption of a stable portfolio shield is broken. Covariance is non-stationary, meaning that a single time-invariant matrix will severely miscalculate risk during shifts in environment.
 3. Verdict: `TBD`
 
**CAS**: The Complex Adaptive Systems paradigm infers that because joint asset transitions are governed by regime-dependent contact geometry, a portfolio’s effective diversification benefit will systematically shrink exactly during the deep market drawdowns an investor cares about avoiding.
 
 1. Gathering Evidence: Asymmetric Correlation Co-Movement Test measuring the mean difference in cross-asset correlations ($\Delta \rho_{i,j}$) and the distance (Frobenius Norm) between $\Sigma_{\text{Low}}$ and $\Sigma_{\text{High}}$ during distinct `VIX` regimes.
 2. Decision Rule:
    - **IF** the distance metric between the matrices is statistically significant **AND** the average cross-asset correlation coefficient increases dramatically in the High-VIX matrix ($\bar{\rho}_{\text{High}} > \bar{\rho}_{\text{Low}}$):
      - Then the CAS paradigm is validated. Cross-asset dependencies are endogenously regime-dependent, meaning that traditional asset diversification systematically breaks down and disappears precisely when market stress peaks.
    - **IF** the distance metric is negligible **OR** correlations stay stable/decrease during high-volatility spikes:
      - Then the asset ecosystem follows a simpler linear geometry. Correlation shifts are temporary statistical artifacts rather than systemic phase changes.
 3. Verdict: `TBD`
 

## Synthesis

By widening our lens from isolated prices to joint relationships, we can test whether cross-asset correlations behave like fixed, immutable laws (NCE) or dynamic structures that warp based on market stress (CAS). If the tests show that correlations converge toward $1$ during high-VIX environments, it proves that diversification is highly unstable. This motivates our final zoom-out: examining the underlying macroeconomic yield curve that anchors these assets.

# Macro Baseline

Here, we view the entire context of the market and economy.

## Structure

**NCE**: via CAPM, collapses the entire term structure into one risk-free rate — a single number standing in for "the cost of money."
 
 1. Raw Input: Daily DGS10, DGS2, or monthly FEDFUNDS.
 2. Enriched State: A single risk-free scalar ($r_f$) selected based on the closest maturity to the asset's cash flows (e.g., using DGS2 for a 2-year horizon contract).
 3. Compression: The chosen flat scalar applied as a uniform baseline denominator in linear valuation equations.

**CAS**: via the Level/Slope/Curve Decomposition, treats the yield curve as a small multi-dimensional object, where the short end and long end can move independently.
 
 1. Raw Input: Co-moving rates across the term spectrum (FEDFUNDS, DGS2, and DGS10) matched with macroeconomic inflation context (CPIAUCSL).
 2. Enriched State: A 3-dimensional coordinate state vector mapping the classical Litterman-Scheinkman yield curve factors directly:
   - $\text{Level}_t = \frac{\text{FEDFUNDS}_t + \text{DGS2}_t + \text{DGS10}_t}{3}$ (Overall altitude of interest rates)
   - $\text{Slope}_t = \text{DGS10}_t - \text{FEDFUNDS}_t$ (The long-to-short macro spread)
   - $\text{Curvature}_t = (\text{DGS10}_t - \text{DGS2}_t) - (\text{DGS2}_t - \text{FEDFUNDS}_t) = \text{DGS10}_t - 2\cdot\text{DGS2}_t + \text{FEDFUNDS}_t$ (The structural bend/twist of the medium-term belly relative to the anchors)
 3. Compression: Projecting this 3D coordinate $[L_t, S_t, C_t]$ onto distinct, discrete macro-regime charts (e.g., Bear Flattener, Bull Steeper, or Inverted Belly).

This directly invokes the Litterman-Scheinkman 3-Factor Principal Component Law, which states that over 95% of all term structure variations are completely captured by these three orthogonal geometric transformations (Level, Slope, and Curvature). 

## Dynamics

**NCE**: Assumes shifts in "the" rate are parallel and exogenous — everything moves together. 
 
 1. Transition (Conservative): The curve transitions uniformly. Short-rate models treat the entire curve as a single string pulled by a central mean-reverting anchor.
 2. Grouped Context of Transitions: Macroeconomic adjustments are uniform. A shift in central bank policy alters the short-term rate, and that change transfers linearly up the maturity spectrum, maintaining rigid, parallel tracking across `FEDFUNDS`, `DGS2`, and `DGS10`.

**CAS**: Allows the curve to twist (short end rising while the long end holds, or vice versa) which a single-number model can't represent at all. 
 
 1. Transition (Diffusion): The curve moves via continuous Non-Parallel Transport and Curvature Shifts. For example, during an aggressive central bank tightening cycle, FEDFUNDS shoots up, DGS2 spikes violently due to medium-term policy stress, while DGS10 lags behind due to long-term economic deceleration.
 2. Grouped Transitions: Systemic rate movements are grouped by their geometric deformations (Litterman & Scheinkman 1991). Changes are path-dependent, where shifts in the curve's Slope or Curvature act as macroeconomic filters that alter the valuation dynamics of specific assets (e.g., separating capital between pure liquidity proxies like `BTC-USD` and debt-laden operational equities like `UAL`).

## Inference

**NCE**: Checks if a single discount rate is adequate for valuing any asset. 
 
 1. Evidence: Multivariate Cointegration and Parallelism Test running a Johansen cointegration test across the interest rate vector $[\text{FEDFUNDS}, \text{DGS2}, \text{DGS10}]$ to verify if the spatial distance between maturities remains strictly locked in a parallel, stationary equilibrium over time ($H_0: \Delta\text{Slope} = 0 \text{ and } \Delta\text{Curvature} = 0$).
 2. Decision Rule:
    - **IF** the rates are cointegrated with stationary spreads **AND** the changes in Slope and Curvature coordinates do not deviate significantly from zero over time:
      - Then NCE discounting is vindicated. The term structure can be safely compressed into a single baseline number representing the global cost of capital without losing valuation accuracy.
    - **IF** the cointegration relationships fail **OR** the spreads display structural non-stationarity and frequent non-parallel shifts:
      - Then the NCE single-scalar model introduces systemic mispricing. A single interest rate variable cannot accurately capture the multi-dimensional macro risk backdrop.
 3. Verdict: `TBD`
 
 
**CAS**: Checks if a firm's or asset's rate sensitivity depends on which part of the curve it's exposed to (e.g., BTC-USD's rate sensitivity behaves differently than UAL's, which carries real duration-like debt exposure). 
 
 1. Evidence: 3-Factor Regression Elasticity Test running rolling multivariate linear regressions of asset log returns against the individual, orthogonal coordinates of $\text{Level}_t$, $\text{Slope}_t$, and $\text{Curvature}_t$ under varying inflation (`CPIAUCSL`) backdrops.
 2. Decision Rule:
    - **IF** individual assets display highly asymmetrical, statistically significant elasticities to specific factors (e.g., `BTC-USD` reacting intensely to short-end `FEDFUNDS`/`DGS2` liquidity transformations, while `UAL` is tied strictly to long-end `DGS10` Slope developments):
      - Then the CAS multi-factor view is fully validated. The macro environment is a multi-dimensional geometry where curve deformations (like a steep 2-year inversion) cause localized, non-linear shifts in risk and value across different asset ecosystems.
    - **IF** all asset sensitivities are uniform across Level, Slope, and Curvature variables, or if multi-factor sensitivities lack statistical significance:
      - Then the macro environment behaves symmetrically, and separating the term structure into sub-components adds no predictive value.
 3. Verdict: `TBD`
 

## Synthesis
The Macro Baseline represents our widest analytical lens, mapping the foundational environment that underpins all asset classes. By evaluating Level, Slope, and Curvature factors alongside inflation, we can infer whether the macroeconomy behaves as a flat baseline cost of capital (NCE) or a dynamic multi-dimensional field (CAS). If the data reveals non-parallel twists and highly localized asset sensitivities, it confirms that macro risk is highly structural and dependent on curve geometry.

# Conclusion

# References:

- [1] Bouchaud, J. P., Farmer, J. D., & Lillo, F. (2009). "How Markets Slowly Digest Large Trading Orders." Handbook of Financial Markets: Dynamics and Evolution, 57–160.
- [2] Fama, E. F. (1970). "Efficient Capital Markets: A Review of Theory and Empirical Work." The Journal of Finance, 25(2), 383–417.
- [3] Grinold, R. C. (1989). "The Fundamental Law of Active Management." The Journal of Portfolio Management, 15(3), 30–38.
- [4] Litterman, R., & Scheinkman, J. (1991). "Common Factors Affecting Bond Returns." The Journal of Fixed Income, 1(1), 54–61.
- [5] Mandelbrot, B. (1963). "The Variation of Certain Speculative Prices." The Journal of Business, 36(4), 394–419.
- [6] Markowitz, H. (1952). "Portfolio Selection." The Journal of Finance, 7(1), 77–91.
- [7] Sharpe, W. F. (1964). "Capital Asset Prices: A Theory of Market Equilibrium under Conditions of Risk." The Journal of Finance, 19(3), 425–442.
- [8] jetbundle. (2026). the black book of quantitative finance. https://www.youtube.com/watch?v=AKIkhuR7HYY
- [9] jetbundle. (2026). 6 laws behind modern quantitative finance. https://www.youtube.com/watch?v=cvL_uWtMTAo

