# SQL Based Financial Investigation

## Framework

Before executing the database queries, its best to define the theoretical framework driving them. This section is split into two sub frameworks: 

 - **The Neoclassical Economics (NCE) Paradigm** - classic financial concepts that treats the financial system as a linear, rational, mechanical engine that naturally tends toward a stable equilibrium.
 - **The Complex Adaptive Systems (CAS) Paradigm** - it views markets not as sterile machines, but as biological ecosystems where feedback loops, friction, and shifting participant behavior constantly warp the environment out of equilibrium.

Here are financial mechanics for the SQL Based Investigations
| Financial Mechanic  | NCE Framework                         | CAS Framework                         |
| :-----------------  | :------------------------------------ | :------------------------------------ |
| Price Movement      | Random Walk / Normal Distribution     | Anomalous Diffusion & Fat Tails       |
| Risk & Correlation  | Static Modern Portfolio Theory (MPT)  | Volatility Regimes & Convergence      |
| Macro Baseline      | Single-Point Risk-Free Rate (CAPM)    | Yield Curve Dimensional Topology      |


## SQL Investigations

Side note, every SQL Investigation has fixed pattern set by the instructor:
Objective, Code, Output, Financial Investigation. This will be shorten to OCOFI


Outline:
### Financial Analysis: 
Models how the economy works based on the two paradigms.

#### Characterizing Price Movements

**Main point**: Modeling how financial instruments move is essential to understanding how they move.

##### NCE: Random Walk and Normal Distribution

**Def. Random Walk**: Asset price movements are modeled as Geometric Brownian motion where log-returns are independent and identically distributed normal random variables - a random walk. 

Reference: https://www.investopedia.com/terms/r/randomwalktheory.asp

**Log-Adjacent-Returns**:
```{r visualize-random-walks-log-adj,  fig.height = 15}
long_returns <- master_df %>%
  select(Date, ends_with(".LogAdj")) %>%
  pivot_longer(cols = -Date, names_to="Financial_Instrument", values_to = "Log_Return")
  
random_walks <- ggplot(
  long_returns,
  aes(
    x = Date, 
    y = Log_Return, 
    color = Financial_Instrument)
  ) +
  geom_line(
    alpha = 0.6, 
    show.legend = TRUE
  ) +
  facet_wrap(
    ~ Financial_Instrument, 
    scales = "free_y",
    ncol = 1
  ) +
  labs(
    title = "Historical Log Returns Across All Instruments",
    subtitle = "Faceted view with independent y-axes",
    x = "Date",
    y = "Log Return"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold", size = 14)
  )

ggplotly(random_walks)
```


**Def. Normal Distribution**: 

**Normal distribution**:
```{r normal-distribution}
long_returns <- master_df %>%
  select(Date, ends_with(".LogAdj")) %>%
  pivot_longer(cols = -Date, names_to="Financial_Instrument", values_to = "Log_Return")

normal_dist <- long_returns %>%
  group_by("Financial_Instrument") %>%
  reframe(
    Log_Return = seq(min(Log_Return), max(Log_Return), length.out = 200),
    Density = dnorm(Log_Return, mean = mean(Log_Return), sd = sd(Log_Return))
  )

normal_distribution <- ggplot() +
  geom_histogram(
    data=long_returns,
    aes(
      x = Log_Return,
      y = after_stat(density)
    ),
    bins = 200,
    fill = "gray40",
    color = "white",
    alpha = 0.5
  ) + 
  geom_density(
    data = long_returns, 
    aes(
      x = Log_Return, 
      color = "Empirical Density"
    ), 
    size = 0.8
  ) + 
  geom_line(data = normal_dist, aes(x = Log_Return, y = Density, color = "Theoretical Normal"), 
            size = 0.9, linetype = "dashed") +
  # Separation by Instrument
  facet_wrap(~ Financial_Instrument, scales = "free", ncol = 2) +
  scale_color_manual(values = c("Empirical Density" = "#1f77b4", "Theoretical Normal" = "#d62728")) +
  labs(
    title = "Empirical Log Returns vs. Theoretical Normal Distributions",
    subtitle = "Assessing Fat Tails vs. Gaussian Normality across assets",
    x = "Log Return",
    y = "Density",
    color = "Curve Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10)
  )

overall_sd <- sd(long_returns$Log_Return, na.rm = TRUE)
overall_mean <- mean(long_returns$Log_Return, na.rm = TRUE)

normal_distribution_zoomed <- normal_distribution +
  coord_cartesian(
    xlim = c(overall_mean - 1.5 * overall_sd, overall_mean + 1.5 * overall_sd)
  )

ggplotly(normal_distribution_zoomed)
```

```{r left-risk-tail}
left_tail <- ggplot() +
  geom_histogram(data = log_returns, aes(x = Log_Return, y = after_stat(density)), 
                 bins = 100, fill = "gray40", color = "white", alpha = 0.5) +
  facet_wrap(~ Financial_Instrument, scales = "free", ncol = 2) +
  
  # ZOOMING IN: Focus precisely on the extreme left tail (e.g., -20% to -5% returns)
  coord_cartesian(xlim = c(-0.25, -0.05)) + 
  
  labs(
    title = "Extreme Negative Tail Risk Details",
    subtitle = "Zoomed coordinate mapping without discarding outlier rows",
    x = "Log Return",
    y = "Density"
  ) +
  theme_minimal()

ggplotly(left_tail)
```

```{r right-risk-tail}
right_tail <- ggplot() +
  geom_histogram(data = log_returns, aes(x = Log_Return, y = after_stat(density)), 
                 bins = 100, fill = "gray40", color = "white", alpha = 0.5) +
  facet_wrap(~ Financial_Instrument, scales = "free", ncol = 2) +
  
  # ZOOMING IN: Focus precisely on the extreme left tail (e.g., -20% to -5% returns)
  coord_cartesian(xlim = c(0.05, 0.25)) + 
  
  labs(
    title = "Extreme Negative Tail Risk Details",
    subtitle = "Zoomed coordinate mapping without discarding outlier rows",
    x = "Log Return",
    y = "Density"
  ) +
  theme_minimal()

ggplotly(right_tail)
```

##### CAS: Anomalous Diffusion & Fat Tails  

**Def. Fat Tails** 

**Fat Tails**:
```{r fat-tail-verification}

# Using your long-format log returns
long_log_returns <- master_df %>%
  select(Date, ends_with(".LogAdj")) %>%
  pivot_longer(cols = -Date, names_to = "Financial_Instrument", values_to = "Log_Return")


# Generate baseline normal curves for comparison
normal_bounds <- long_log_returns %>%
  group_by(Financial_Instrument) %>%
  reframe(
    Log_Return = seq(min(Log_Return), max(Log_Return), length.out = 300),
    Density = dnorm(Log_Return, mean = mean(Log_Return), sd = sd(Log_Return))
  )

# Plot with log-scaled density to expose the Fat Tails
cas_plot <- ggplot() +
  geom_density(data = long_log_returns, aes(x = Log_Return, color =Financial_Instrument), size = 0.5, alpha=0.6) +
  geom_line(data = normal_bounds, aes(x = Log_Return, y = Density, color = "Theoretical NCE (Normal)"), linetype = "dashed", size = 0.5) +
  facet_wrap(~ Financial_Instrument, scales = "free", ncol = 2) +
  labs(
    title = "Fat Tail Verification (Log-Scale Density)",
    subtitle = "Straight lines/bowing in the outer wings reveal power-law fat tails",
    x = "Log Return",
    y = "Log(Density)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))

ggplotly(cas_plot)
```

**Def. Anamalous Diffusion**:

**Anamalous Diffusion**:
```{r anamalous-diffusion-visualization}
# Define the time horizons in days to evaluate variance scaling
horizons <- c(1, 5, 10, 20, 40, 60, 80, 120)

# Reshape master_df into long format and drop NA rows
long_logs <- master_df %>%
  select(Date, ends_with(".LogAdj")) %>%
  pivot_longer(cols = -Date, names_to = "Instrument", values_to = "Log_Return") %>%
  filter(!is.na(Log_Return))

# Calculate Empirical Variance across all horizons for each asset
empirical_scaling <- map_dfr(horizons, function(h) {
  long_logs %>%
    group_by(Instrument) %>%
    arrange(Date) %>%
    # Calculate rolling h-day log returns using a rolling sum
    mutate(H_Return = rollsum(Log_Return, k = h, fill = NA, align = "right")) %>%
    filter(!is.na(H_Return)) %>%
    summarise(Variance = var(H_Return), .groups = "drop") %>%
    mutate(Horizon = h, Type = "Empirical CAS")
})

# Calculate Theoretical NCE Variance Scaling baseline (Linear scaling: Var_h = h * Var_1)
theoretical_scaling <- empirical_scaling %>%
  filter(Horizon == 1) %>%
  select(Instrument, Base_Var = Variance) %>%
  cross_join(data.frame(Horizon = horizons)) %>%
  mutate(
    Variance = Horizon * Base_Var,
    Type = "Theoretical NCE (Linear)"
  ) %>%
  select(Instrument, Variance, Horizon, Type)

# Combine both data sets for a direct comparison
all_assets_scaling <- bind_rows(empirical_scaling, theoretical_scaling)

# Visualization
anamalous_diffusion <- ggplot(
  all_assets_scaling, 
  aes(
    x = Horizon, 
    y = Variance, 
    color = Type, 
    linetype = Type
  )) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Instrument, scales = "free_y", ncol = 2) +
  
  # Log-Log Scale: Essential to spot power laws and Hurst exponent slopes
  scale_x_log10(breaks = horizons) +
  scale_y_log10() +
  
  scale_color_manual(values = c("Empirical CAS" = "#00b4d8", "Theoretical NCE (Linear)" = "#7209b7")) +
  labs(
    title = "Anomalous Diffusion Matrix: Variance Scaling Across All Assets",
    subtitle = "Divergence from the purple line reveals market memory & systemic friction",
    x = "Log(Time Horizon in Days)",
    y = "Log(Variance of Compound Returns)",
    color = "Framework Paradigm"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold", size = 14)
  )

ggplotly(anamalous_diffusion)
```

##### Comparison and Synthesis

#### Risk & Covariance: 
**Main point**
##### NCE: Static Modern Portfolio Theory (MPT)
**Def. Modern Portfolio Theory**:

```{r }
mpt_assets <- master_df %>%
  select(UAL.LogAdj, PAL.LogAdj, `BTC-USD.LogAdj`, SPY.LogAdj) %>%
  drop_na()

expected_returns <- colMeans(mpt_assets) * 100 # Expressed as %
print("--- Expected Daily Returns (%) ---")
print(round(expected_returns, 4))

covariance_matrix <- cov(mpt_assets)
print("--- Static Covariance Matrix ---")
print(round(covariance_matrix, 6))

correlation_matrix <- cor(mpt_assets)
```

##### CAS
      - Def. Volatitlity Regimes
      - Def. Convergence
      - OCOFI
##### Comparison and Synthesis
    
    
   - Macro Baseline:
    - Main point
    - NCE:
      - Def. CAPM
      - OCOFI
    - CAS:
      - Def. Yield Curve Dimensional Topology
      - OCOFI
    - Comparison and Synthesis
   - Market Efficiency / Information Movement:
    - Main point
    - NCE:
      - Def. Perfect Information / Zero Friction
      - OCOFI
    - CAS:
      - Def. Relativistic Arbitrage
      - OCOFI
    - Comparison and Synthesis
   - Portfolio Sizing:
    - Main point
    - NCE:
      - Def. Mean-Variance Optimization
      - OCOFI
    - CAS:
      - Fundamental Law of Active Management
      - OCOFI
    - Comparison and Synthesis
 - Portfolio Investment Specific:
  - Mainpoint: Recommend an allocation of USD 1,000,000 among the selected U.S. asset, Philippine asset, cryptocurrency, and cash. Justify your recommendation using evidence from the analysis.
  - Points from the previous section
  - Sections of SQL code that provide additional insight.


## References:
[1] Bouchaud, J. P., Farmer, J. D., & Lillo, F. (2009). "How Markets Slowly Digest Large Trading Orders." Handbook of Financial Markets: Dynamics and Evolution, 57–160.
[2] Fama, E. F. (1970). "Efficient Capital Markets: A Review of Theory and Empirical Work." The Journal of Finance, 25(2), 383–417.
[3] Grinold, R. C. (1989). "The Fundamental Law of Active Management." The Journal of Portfolio Management, 15(3), 30–38.
[4] Litterman, R., & Scheinkman, J. (1991). "Common Factors Affecting Bond Returns." The Journal of Fixed Income, 1(1), 54–61.
[5] Mandelbrot, B. (1963). "The Variation of Certain Speculative Prices." The Journal of Business, 36(4), 394–419.
[6] Markowitz, H. (1952). "Portfolio Selection." The Journal of Finance, 7(1), 77–91.
[7] Sharpe, W. F. (1964). "Capital Asset Prices: A Theory of Market Equilibrium under Conditions of Risk." The Journal of Finance, 19(3), 425–442.
[8] jetbundle. (2026). the black book of quantitative finance. https://www.youtube.com/watch?v=AKIkhuR7HYY
[9] jetbundle. (2026). 6 laws behind modern quantitative finance. https://www.youtube.com/watch?v=cvL_uWtMTAo


