---
title: "Main"
author: "Roan Campo, Denzel Macayan, Bea Uy, Miguel Uy"
date: "2026-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-environment-dependencies}
required_packages <- c(
  "tidyr", "dplyr", "tibble", "purrr", "moments", "quantmod", "reticulate", "httr", "jsonlite", "xts", "readr","stringr", "ggplot2", "quantmod"
)

missing_packages <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]

if (length(missing_packages) > 0) {
  cat("Installing missing dependencies:", paste(missing_packages, collapse = ", "), "\n")
  install.packages(missing_packages)
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r set-date}
start_date <- as.Date("2019-12-01")
end_date <- as.Date("2025-12-31")
```

```{r fetch-pse-data}
pal_raw <- read_csv("pal_stock.csv", show_col_types = FALSE) 

pal_df <- pal_raw %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y")) %>%
  
  filter(!is.na(Date)) %>%
  
  arrange(Date) %>%
  
  mutate(
    `Vol.` = case_when(
      str_detect(`Vol.`, "K") ~ as.numeric(str_remove(`Vol.`, "K")) * 1000,
      str_detect(`Vol.`, "M") ~ as.numeric(str_remove(`Vol.`, "M")) * 1000000,
      `Vol.` == "-" ~ 0,
      TRUE ~ as.numeric(`Vol.`)
    )
  ) %>%
  select(Date, Open, High, Low, Price, `Vol.`) %>% 
  rename(Close = Price, Volume = `Vol.`) %>%
  
  mutate(AdjClose = Close) 

pal_xts <- xts(pal_df[, -1], order.by = pal_df$Date)

colnames(pal_xts) <- c("PAL.Open", "PAL.High", "PAL.Low", "PAL.Close", "PAL.Volume", "PAL.Adjusted")

PAL <- pal_xts["2019-12/2025-12"]
```

```{r load-yahoo}
yahoo_tickers <- c(
  US_Equity = "UAL",     # US Equity: United Airlines
  Bitcoin = "BTC-USD",   # Bitcoin in USD
  SPY = "SPY",           # SPDR S&P 500 ETF Trust
  VIX = "^VIX"           # CBOE Volatility Index
)

getSymbols(yahoo_tickers, src = "yahoo", from = start_date, to = end_date)
```


```{r load-fred}
fred_tickers <- c(
  "DGS10",
  "FEDFUNDS",
  "CPIAUCSL",
  "DCOILWTICO"
)

getSymbols(fred_tickers, src = "FRED")

DGS10 <- window(DGS10, start = start_date, end = end_date)
FEDFUNDS <- window(FEDFUNDS, start = start_date, end = end_date)
CPIAUCSL <- window(CPIAUCSL, start = start_date, end = end_date)
DCOILWTICO <- window(DCOILWTICO, start = start_date, end = end_date)
```
```{r init-obs}
instruments <- c("UAL", "PAL", "BTC-USD", "SPY", "VIX", "DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO")
init_obs <- function(ticker_list, sample_size = 3){
   for (sym in ticker_list) {
    data_obj <- get(sym)
    cat("INSTRUMENT:", sym, "| Total Observations:", nrow(data_obj), "\n")
    
    print(head(data_obj))
    cat("\n")
    
    na_indices <- which(rowSums(is.na(data_obj)) > 0)
    missing_count <- length(na_indices)
    
    cat("Total NAs:", missing_count, "\n\n")
    if (missing_count > 0) {
      for (idx in head(na_indices, sample_size)) {
        start_idx <- max(1, idx - 1)
        end_idx <- min(nrow(data_obj), idx + 1)
        print(data_obj[start_idx:end_idx, ])
        cat("------\n")
      }
      cat("\n")
    }
    
    dup_indices <- which(duplicated(zoo::index(data_obj)))
    dup_count <- length(dup_indices)
    cat("Total Duplicates:", dup_count, "\n\n")
    
    if (dup_count > 0) {
      for (idx in head(dup_indices, sample_size)) {
        start_idx <- max(1, idx - 1)
        end_idx <- min(nrow(data_obj), idx + 1)
        
        print(data_obj[start_idx:end_idx, ])
        cat("------\n")
      }
      cat("\n")
    }
  } 
}
init_obs(instruments)
```

```{r fix-local-na}
for (sym in instruments) {
  temp_data <- get(sym)
  temp_data <- na.locf(temp_data)
  assign(sym, temp_data)
}
```

```{r check-local-na-fix}
init_obs(instruments)
```

```{r convert-to-df}
df_list <- map(instruments, function(sym) {
  data_obj <- get(sym)
  
  # If it has an Adjusted column, extract it. Otherwise, use the raw data object.
  if (has.Ad(data_obj)) {
    extracted_data <- Ad(data_obj)
  } else {
    extracted_data <- data_obj
  }
  
  clean_df <- as.data.frame(extracted_data) %>% 
    rownames_to_column(var = "Date") %>% 
    mutate(Date = as.Date(Date)) %>%
    as_tibble()
  
  return(clean_df)
})
names(df_list) <- instruments

master_df <- df_list %>%
  compact() %>%
  reduce(full_join, by = "Date") %>%
  arrange(Date) %>%
  fill(-Date, .direction = "down")
```

This is to detect PAL's trading halt window. PAL was suspended by PSE due to audit rule violation.
```{r bea-pal-halt-detect}
pal_native <- df_list[["PAL"]] %>% 
  arrange(Date)

# find which dates have a gap greater than 10 -> will consider this as halt
pal_gap_idx <- which(as.numeric(diff(pal_native$Date)) > 10)

pal_halt_windows <- tibble(
  halt_start = pal_native$Date[pal_gap_idx] + 1,
  halt_end   = pal_native$Date[pal_gap_idx + 1] - 1
)

cat("Detected PAL trading halt window(s):\n")
print(pal_halt_windows)
cat("(Expected: 2021-06-17 to 2022-03-27 -- PSE audit-opinion suspension)\n")

in_pal_halt <- function(dates) {
  purrr::map_lgl(dates, function(d) any(d >= pal_halt_windows$halt_start & d <= pal_halt_windows$halt_end))
}
```

```{r log-returns}
master_df <- master_df %>%
  mutate(across(
    -Date, 
    ~ c(NA, diff(log(.x)))
  )) %>%
  rename_with(~ str_replace(.x, "\\.Adjusted$", ".LogAdj"), -Date)
```

```{r explore-master-df}
explore_instruments <- function(instruments, master_df) {
  for (sym in instruments) {
    cat("INSTRUMENT:", sym, "\n")
    
    instrument_df <- master_df %>% 
      select(Date, starts_with(sym))
    
    print(head(instrument_df), n = 3)
    print(tail(instrument_df), n = 3)
    cat("\n")
    
    cat("--- Missing Values per Column ---\n")
    instrument_df %>% 
      summarise(across(everything(), ~sum(is.na(.)))) %>% 
      print(width = Inf)
    cat("\n")
    
    missing_rows <- instrument_df %>%
      filter(if_any(starts_with(sym), is.na))
    
    total_missing <- nrow(missing_rows)
    cat("Total Rows with NAs:", total_missing, "\n")
    
    if (total_missing > 0) {
      cat("--- Top 10 Missing Rows ---\n")
      print(missing_rows, n = 10)
    }
    
    dup_count <- sum(duplicated(instrument_df$Date))
  }
  
  dplyr::glimpse(master_df)
}
explore_instruments(instruments, master_df)
```

```{r fix-missing-master-df}
master_df <- master_df %>%
  fill(-Date, .direction = "down") %>%
  filter(Date >= as.Date("2020-01-01"))

explore_instruments(instruments, master_df)
```

```{r function-descriptive-statistics}
analyze_instrument <- function(df, col_name) {
  inst_data <- df %>%
    select(Date, Value = all_of(col_name)) %>%
    filter(!is.na(Value))
  
  summary_stats <- tibble(
    Instrument = col_name,
    Mean = mean(inst_data$Value),
    Median = median(inst_data$Value),
    Variance = var(inst_data$Value),
    Std_Dev = sd(inst_data$Value),
    Min = min(inst_data$Value),
    Max = max(inst_data$Value),
    Range = Max - Min,
    Skewness = skewness(inst_data$Value), 
    Kurtosis = kurtosis(inst_data$Value)
  ) %>%
    mutate(across(where(is.numeric), ~round(., 4)))
  
  cat("\n======================================================\n")
  cat("### Daily Summary Statistics:", col_name, "\n")
  cat("======================================================\n")
  print(summary_stats)
  cat("\n")
  
  line_plot <- ggplot(inst_data, aes(x = Date, y = Value)) +
    geom_line(color = "darkblue", alpha = 0.8) +
    theme_minimal() +
    labs(
      title = paste("Daily Timeline of", col_name, "(2020 - 2025)"),
      x = "Date",
      y = "Value / Log Return"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold")
    )
  
  dist_plot <- ggplot(inst_data, aes(x = Value)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "steelblue", color = "white", alpha = 0.7) +
    geom_density(color = "darkred", linewidth = 1) +
    theme_minimal() +
    labs(
      title = paste("Daily Distribution of", col_name, "(2020 - 2025)"),
      x = "Value / Log Return",
      y = "Density"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold")
    )
  
  print(line_plot)
  print(dist_plot)
}

target_cols <- c(
  "UAL.LogAdj", 
  "PAL.LogAdj", 
  "BTC-USD.LogAdj", 
  "SPY.LogAdj", 
  "VIX.LogAdj", 
  "DGS10", 
  "FEDFUNDS", 
  "CPIAUCSL",
  "DCOILWTICO"
)

for (col in target_cols) {
  analyze_instrument(master_df, col)
}
```
```{r name-df-list}
# df_list was built in instrument order in `convert-to-df` (pre full_join, pre fill).
# Name it so individual per-instrument tibbles can be pulled out by ticker for the
# join demonstrations below, instead of re-splitting the already-joined master_df.
names(df_list) <- instruments
```

```{r bea-leftjoin1-pre-obs fuel-pass-through}
cat("UAL:", nrow(df_list[["UAL"]]), "rows |",
    format(min(df_list[["UAL"]]$Date)), "to", format(max(df_list[["UAL"]]$Date)), "\n")
cat("PAL:", nrow(df_list[["PAL"]]), "rows |",
    format(min(df_list[["PAL"]]$Date)), "to", format(max(df_list[["PAL"]]$Date)), "\n")
cat("DCOILWTICO:", nrow(df_list[["DCOILWTICO"]]), "rows |",
    format(min(df_list[["DCOILWTICO"]]$Date)), "to", format(max(df_list[["DCOILWTICO"]]$Date)), "\n\n")
cat("Note: PSE (PAL) and NYMEX (WTI) calendars differ from NYSE (UAL).\n")
cat("A left_join on UAL's calendar will introduce NAs on days PAL/WTI didn't print,\n")
cat("which we carry forward via fill() rather than dropping UAL rows.\n")
```
The difference in rows could be explained by the following:
- UAL is based on NYSE weekday calendar, which is closed on weekends and US market holidays. 
- PAL is based on PSE weekday calendar, which is closed on weekends and PH holidays.
- PAL had a 285-day trading suspension imposed by the PSE between (2021-06-17 to 2022-03-28),
  as its financial statements had violated a rule. This was during COVID-19 travel restrictions.
- DCOILWTICO is from FRED instead of a stock market trading calendar. 
```{r bea-leftjoin1}
fuel_leftjoin_df <- df_list[["UAL"]] %>%
  select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  left_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  left_join(df_list[["DCOILWTICO"]] %>% select(Date, WTI = DCOILWTICO), by = "Date") %>%
  arrange(Date) %>%
  fill(PAL_Adjusted, WTI, .direction = "down") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted))),
    WTI_ret = c(NA, diff(log(WTI)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(WTI_ret))

cat("Rows kept (UAL trading calendar):", nrow(fuel_leftjoin_df), "\n\n")
print(fuel_leftjoin_df, n = 6)

ual_fuel_model <- lm(UAL_ret ~ WTI_ret, data = fuel_leftjoin_df)
pal_fuel_model <- lm(PAL_ret ~ WTI_ret, data = fuel_leftjoin_df)
print(summary(ual_fuel_model)$coefficients)
print(summary(pal_fuel_model)$coefficients)
```

```{r bea-leftjoin1-post-obs}
ual_oil_beta <- coef(ual_fuel_model)[["WTI_ret"]]
pal_oil_beta <- coef(pal_fuel_model)[["WTI_ret"]]
ual_oil_p <- summary(ual_fuel_model)$coefficients["WTI_ret", "Pr(>|t|)"]
pal_oil_p <- summary(pal_fuel_model)$coefficients["WTI_ret", "Pr(>|t|)"]

cat("Rows retained on UAL's calendar:", nrow(fuel_leftjoin_df), "\n")
cat("UAL oil-return coefficient:", round(ual_oil_beta, 4), "| p-value:", round(ual_oil_p, 4), "\n")
cat("PAL oil-return coefficient:", round(pal_oil_beta, 4), "| p-value:", round(pal_oil_p, 4), "\n")
```

- `fuel_leftjoin_df` - PAL's price forward-filled across every gap (including weekends, holidays, and the trading suspension)
- `fuel_leftjoin_df_masked` - weekends and holidays are forward-filled but the trading suspension is 
```{r bea-pal-masked-companion}

# dataframe where the trading halt (285-day suspension) are masked as NA
# this is built from the fuel_leftjoin_df which is forward-filled
fuel_leftjoin_df_masked <- fuel_leftjoin_df %>%
  mutate(PAL_ret = if_else(in_pal_halt(Date), NA_real_, PAL_ret)) %>%
  filter(!is.na(PAL_ret))

pal_fuel_model_masked <- lm(PAL_ret ~ WTI_ret, data = fuel_leftjoin_df_masked)
pal_oil_beta_masked <- coef(pal_fuel_model_masked)[["WTI_ret"]]
pal_oil_p_masked <- summary(pal_fuel_model_masked)$coefficients["WTI_ret", "Pr(>|t|)"]

cat("Naive (forward-filled) rows:", nrow(fuel_leftjoin_df), "| PAL oil-beta:", round(pal_oil_beta, 4), "\n")
cat("Masked (halt-excluded) rows:", nrow(fuel_leftjoin_df_masked), "| PAL oil-beta:", round(pal_oil_beta_masked, 4), "\n")
```

```{r bea-innerjoin-pre-obs BTC/airline correlation}
cat("BTC-USD:", nrow(df_list[["BTC-USD"]]), "rows | UAL:", nrow(df_list[["UAL"]]),
    "rows | PAL:", nrow(df_list[["PAL"]]), "rows\n")
cat("inner_join keeps only the intersection -- every BTC weekend/holiday move is\n")
cat("excluded by construction; loss is quantified explicitly below.\n")
```

```{r bea-innerjoin}
btc_innerjoin_df <- df_list[["BTC-USD"]] %>%
  select(Date, BTC_Adjusted = `BTC-USD.Adjusted`) %>%
  inner_join(df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`), by = "Date") %>%
  inner_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  arrange(Date) %>%
  mutate(
    BTC_ret = c(NA, diff(log(BTC_Adjusted))),
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(BTC_ret), !is.na(UAL_ret), !is.na(PAL_ret))

n_btc_total <- nrow(df_list[["BTC-USD"]])
n_common <- nrow(btc_innerjoin_df)
cor_matrix <- cor(btc_innerjoin_df %>% select(BTC_ret, UAL_ret, PAL_ret), use = "complete.obs")
print(round(cor_matrix, 4))
```

```{r bea-innerjoin-post-obs}
btc_ual_cor <- cor_matrix["BTC_ret", "UAL_ret"]
btc_pal_cor <- cor_matrix["BTC_ret", "PAL_ret"]
ual_pal_cor <- cor_matrix["UAL_ret", "PAL_ret"]

cat("Common trading days retained:", n_common, "of", n_btc_total, "BTC calendar days\n")
cat("BTC-UAL correlation:", round(btc_ual_cor, 4), "\n")
cat("BTC-PAL correlation:", round(btc_pal_cor, 4), "\n")
cat("UAL-PAL correlation:", round(ual_pal_cor, 4), "\n")
```

```{r bea-antijoin-pre-obs}
cat("PAL:", nrow(df_list[["PAL"]]), "rows | UAL:", nrow(df_list[["UAL"]]), "rows\n")
cat("anti_join surfaces every UAL trading day with no matching PAL date --\n")
cat("both routine PSE holidays and the 285-day suspension. Splitting the two\n")
cat("apart tests whether the suspension or routine calendar drift dominates.\n")
```

```{r bea-antijoin pal-ual-gap}
pal_gap_days <- df_list[["UAL"]] %>% select(Date) %>%
  anti_join(df_list[["PAL"]] %>% select(Date), by = "Date")

holiday_gap_days <- pal_gap_days %>% filter(!in_pal_halt(Date))
halted_gap_days  <- pal_gap_days %>% filter(in_pal_halt(Date))

cat("Total UAL trading days with no PAL match:", nrow(pal_gap_days), "\n")
cat("Routine PSE-holiday-only days:", nrow(holiday_gap_days), "\n")
cat("Suspension-only days:", nrow(halted_gap_days), "\n\n")
print(holiday_gap_days, n = 10)
```

```{r bea-antijoin-post-obs}
cat("Of", nrow(pal_gap_days), "total PAL gap days,", nrow(halted_gap_days),
    sprintf("(%.0f%%)", 100 * nrow(halted_gap_days) / nrow(pal_gap_days)),
    "are attributable to the single 285-day suspension --\n")
cat("routine PSE-holiday mismatch accounts for only", nrow(holiday_gap_days), "days.\n")
```

# SQL-Based Financial Investigation

## SQL Investigation 1: Does Forward-Filling PAL's Suspension Bias Its Return Statistics?

**Objective:** Quantifies the how forward-filling PAL's 285-day suspension affects its measured return distribution, compared to excluding (masking) the halt window. Reuses the two parallel PAL series already built above (`fuel_leftjoin_df` naive, `fuel_leftjoin_df_masked` corrected).

```{r bea-sql1}
pal_stats_comparison <- bind_rows(
  fuel_leftjoin_df        %>% mutate(version = "Forward-filled (naive)") %>% select(version, PAL_ret),
  fuel_leftjoin_df_masked %>% mutate(version = "Halt-masked (corrected)") %>% select(version, PAL_ret)
) %>%
  group_by(version) %>%
  summarise(
    N = n(),
    Zero_Return_Pct = round(100 * mean(PAL_ret == 0), 1),
    Mean = round(mean(PAL_ret), 6),
    Median = round(median(PAL_ret), 6),
    Variance = round(var(PAL_ret), 6),
    SD = round(sd(PAL_ret), 4),
    Min = round(min(PAL_ret), 4),
    Max = round(max(PAL_ret), 4),
    Range = round(max(PAL_ret) - min(PAL_ret), 4),
    Skewness = round(skewness(PAL_ret), 3),
    Kurtosis = round(kurtosis(PAL_ret), 3),
    .groups = "drop"
  )

print(pal_stats_comparison, width = Inf)
```


## SQL Investigation 2: Which Airline Is More Exposed to Oil -- UAL or PAL?

**Objective:** Compare United Airlines' and Philippine Airlines' sensitivity to WTI crude oil returns, using the corrected (halt-masked) PAL series, to determine which airline's stock carries more fuel-cost risk.

```{r bea-sql2}
oil_direction_summary <- fuel_leftjoin_df_masked %>%
  mutate(oil_direction = if_else(WTI_ret > 0, "Oil Up", "Oil Down")) %>%
  group_by(oil_direction) %>%
  summarise(
    n_days = n(),
    mean_UAL_ret = round(mean(UAL_ret), 5),
    mean_PAL_ret = round(mean(PAL_ret), 5),
    .groups = "drop"
  ) %>%
  arrange(oil_direction)

print(oil_direction_summary)

cat("\nFull-sample regression betas (UAL from bea-leftjoin1; PAL from the masked companion):\n")
cat("UAL oil-beta:", round(ual_oil_beta, 4), "| p =", round(ual_oil_p, 4), "\n")
cat("PAL oil-beta:", round(pal_oil_beta_masked, 4), "| p =", round(pal_oil_p_masked, 4), "\n")
```


## SQL Investigation 3: The Ukraine Invasion Oil Shock -- A Natural Experiment in Market Friction

**Objective:** Test whether UAL's oil-sensitivity shifts during the acute Feb 24-Mar 8, 2022 shock (WTI spiked from ~$92.77 to a $123.64 peak, +33%) relative to its full-sample beta, and highlight that PAL has zero data for this exact window.

```{r bea-sql3}
ukraine_shock_df_all_ual <- fuel_leftjoin_df %>%
  filter(between(Date, as.Date("2022-02-24"), as.Date("2022-03-08")))
ukraine_shock_df_pal <- fuel_leftjoin_df_masked %>%
  filter(between(Date, as.Date("2022-02-24"), as.Date("2022-03-08")))

cat("UAL trading days in Ukraine shock window:", nrow(ukraine_shock_df_all_ual), "\n")
cat("Of which PAL has real (non-halted) data:", nrow(ukraine_shock_df_pal), "\n\n")

shock_vs_full <- tibble(
  Period = c("Full sample", "Ukraine shock (2022-02-24 to 2022-03-08)"),
  N = c(nrow(fuel_leftjoin_df), nrow(ukraine_shock_df_all_ual)),
  UAL_oil_beta = c(
    round(cov(fuel_leftjoin_df$UAL_ret, fuel_leftjoin_df$WTI_ret, use = "complete.obs") / var(fuel_leftjoin_df$WTI_ret, na.rm = TRUE), 4),
    round(cov(ukraine_shock_df_all_ual$UAL_ret, ukraine_shock_df_all_ual$WTI_ret, use = "complete.obs") / var(ukraine_shock_df_all_ual$WTI_ret, na.rm = TRUE), 4)
  ),
  Mean_UAL_ret = c(round(mean(fuel_leftjoin_df$UAL_ret), 5), round(mean(ukraine_shock_df_all_ual$UAL_ret), 5)),
  Mean_WTI_ret = c(round(mean(fuel_leftjoin_df$WTI_ret), 5), round(mean(ukraine_shock_df_all_ual$WTI_ret), 5))
)
print(shock_vs_full, width = Inf)
```


## SQL Investigation 4: Is UAL's Oil-Beta a Stable Relationship, or Just the Ukraine Shock?

**Objective:** Split the sample into pre-shock, shock-era, and recovery sub-periods and recompute UAL's and PAL's oil-beta in each, to test whether Investigation 2's relationship is a structural feature or an artifact of one event. UAL is grouped from the naive `fuel_leftjoin_df` (it has no gaps to mask -- restricting it to PAL's available days would silently exclude the actual shock dates, since they fall inside PAL's halt window); PAL is grouped from `fuel_leftjoin_df_masked`. The two are joined back together by period for a side-by-side table.

```{r bea-sql4}
period_label <- function(d) case_when(
  d < as.Date("2022-01-01") ~ "1. Pre-shock (2020-2021)",
  d < as.Date("2023-01-01") ~ "2. Shock era (2022)",
  TRUE ~ "3. Recovery (2023-2025)"
)

ual_period_beta <- fuel_leftjoin_df %>%
  mutate(period = period_label(Date)) %>%
  group_by(period) %>%
  summarise(
    n_days_ual = n(),
    UAL_beta = round(cov(UAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

pal_period_beta <- fuel_leftjoin_df_masked %>%
  mutate(period = period_label(Date)) %>%
  group_by(period) %>%
  summarise(
    n_days_pal = n(),
    PAL_beta = round(cov(PAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

period_beta <- ual_period_beta %>%
  left_join(pal_period_beta, by = "period") %>%
  arrange(period)

print(period_beta, width = Inf)
```


## SQL Investigation 5: `full_join` vs. `inner_join` -- Quantifying the Correlation Deflation

**Objective:** Directly measure how much a naive `full_join` + forward-fill deflates the UAL-PAL return correlation compared to a strict `inner_join`, closing the loop back to Investigation 1's methodology point and justifying `inner_join`'s use in `bea-innerjoin`.

```{r bea-sql5}
ual_pal_fulljoin_df <- df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  full_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  arrange(Date) %>%
  fill(UAL_Adjusted, PAL_Adjusted, .direction = "down") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(PAL_ret))

ual_pal_innerjoin_df <- df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  inner_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  arrange(Date) %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(PAL_ret))

naive_cor  <- cor(ual_pal_fulljoin_df$UAL_ret, ual_pal_fulljoin_df$PAL_ret, use = "complete.obs")
strict_cor <- cor(ual_pal_innerjoin_df$UAL_ret, ual_pal_innerjoin_df$PAL_ret, use = "complete.obs")

cat("Naive full_join + fill() correlation:", round(naive_cor, 4), "| N =", nrow(ual_pal_fulljoin_df), "\n")
cat("Strict inner_join correlation:", round(strict_cor, 4), "| N =", nrow(ual_pal_innerjoin_df), "\n")
```

