---
title: "Main"
author: "Roan Campo, Denzel Macayan, Bea Uy, Miguel Uy"
date: "2026-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-environment-dependencies}
required_packages <- c(
  "tidyr", "dplyr", "tibble", "purrr", "moments", "quantmod", "reticulate", "httr", "jsonlite", "xts", "readr","stringr", "ggplot2", "quantmod", "ggrepel","zoo", "lubridate"
)

missing_packages <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]

if (length(missing_packages) > 0) {
  cat("Installing missing dependencies:", paste(missing_packages, collapse = ", "), "\n")
  install.packages(missing_packages)
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r set-date}
start_date <- as.Date("2019-12-01")
end_date <- as.Date("2025-12-31")
```

```{r fetch-pse-data}
pal_raw <- read_csv("pal_stock.csv", show_col_types = FALSE) 

pal_df <- pal_raw %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y")) %>%
  
  filter(!is.na(Date)) %>%
  
  arrange(Date) %>%
  
  mutate(
    `Vol.` = case_when(
      str_detect(`Vol.`, "K") ~ as.numeric(str_remove(`Vol.`, "K")) * 1000,
      str_detect(`Vol.`, "M") ~ as.numeric(str_remove(`Vol.`, "M")) * 1000000,
      `Vol.` == "-" ~ 0,
      TRUE ~ as.numeric(`Vol.`)
    )
  ) %>%
  select(Date, Open, High, Low, Price, `Vol.`) %>% 
  rename(Close = Price, Volume = `Vol.`) %>%
  
  mutate(AdjClose = Close) 

pal_xts <- xts(pal_df[, -1], order.by = pal_df$Date)

colnames(pal_xts) <- c("PAL.Open", "PAL.High", "PAL.Low", "PAL.Close", "PAL.Volume", "PAL.Adjusted")

PAL <- pal_xts["2019-12/2025-12"]
```

```{r load-yahoo}
yahoo_tickers <- c(
  US_Equity = "UAL",     # US Equity: United Airlines
  Bitcoin = "BTC-USD",   # Bitcoin in USD
  SPY = "SPY",           # SPDR S&P 500 ETF Trust
  VIX = "^VIX"           # CBOE Volatility Index
)

getSymbols(yahoo_tickers, src = "yahoo", from = start_date, to = end_date)
```


```{r load-fred}
fred_tickers <- c(
  "DGS10",
  "FEDFUNDS",
  "CPIAUCSL",
  "DCOILWTICO"
)

getSymbols(fred_tickers, src = "FRED")

DGS10 <- window(DGS10, start = start_date, end = end_date)
FEDFUNDS <- window(FEDFUNDS, start = start_date, end = end_date)
CPIAUCSL <- window(CPIAUCSL, start = start_date, end = end_date)
DCOILWTICO <- window(DCOILWTICO, start = start_date, end = end_date)
```
```{r init-obs}
instruments <- c("UAL", "PAL", "BTC-USD", "SPY", "VIX", "DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO")
init_obs <- function(ticker_list, sample_size = 3){
   for (sym in ticker_list) {
    data_obj <- get(sym)
    cat("INSTRUMENT:", sym, "| Total Observations:", nrow(data_obj), "\n")
    
    print(head(data_obj))
    cat("\n")
    
    na_indices <- which(rowSums(is.na(data_obj)) > 0)
    missing_count <- length(na_indices)
    
    cat("Total NAs:", missing_count, "\n\n")
    if (missing_count > 0) {
      for (idx in head(na_indices, sample_size)) {
        start_idx <- max(1, idx - 1)
        end_idx <- min(nrow(data_obj), idx + 1)
        print(data_obj[start_idx:end_idx, ])
        cat("------\n")
      }
      cat("\n")
    }
    
    dup_indices <- which(duplicated(zoo::index(data_obj)))
    dup_count <- length(dup_indices)
    cat("Total Duplicates:", dup_count, "\n\n")
    
    if (dup_count > 0) {
      for (idx in head(dup_indices, sample_size)) {
        start_idx <- max(1, idx - 1)
        end_idx <- min(nrow(data_obj), idx + 1)
        
        print(data_obj[start_idx:end_idx, ])
        cat("------\n")
      }
      cat("\n")
    }
  } 
}
init_obs(instruments)
```

```{r fix-local-na}
for (sym in instruments) {
  temp_data <- get(sym)
  temp_data <- na.locf(temp_data)
  assign(sym, temp_data)
}
```

```{r check-local-na-fix}
init_obs(instruments)
```

```{r convert-to-df}
df_list <- map(instruments, function(sym) {
  data_obj <- get(sym)
  
  # If it has an Adjusted column, extract it. Otherwise, use the raw data object.
  if (has.Ad(data_obj)) {
    extracted_data <- Ad(data_obj)
  } else {
    extracted_data <- data_obj
  }
  
  clean_df <- as.data.frame(extracted_data) %>% 
    rownames_to_column(var = "Date") %>% 
    mutate(Date = as.Date(Date)) %>%
    as_tibble()
  
  return(clean_df)
})
names(df_list) <- instruments

master_df <- df_list %>%
  compact() %>%
  reduce(full_join, by = "Date") %>%
  arrange(Date) %>%
  fill(-Date, .direction = "down")
```

This is to detect PAL's trading halt window. PAL was suspended by PSE due to audit rule violation.
```{r bea-pal-halt-detect}
pal_native <- df_list[["PAL"]] %>% 
  arrange(Date)

# find which dates have a gap greater than 10 -> will consider this as halt
pal_gap_idx <- which(as.numeric(diff(pal_native$Date)) > 10)

pal_halt_windows <- tibble(
  halt_start = pal_native$Date[pal_gap_idx] + 1,
  halt_end   = pal_native$Date[pal_gap_idx + 1] - 1
)

cat("Detected PAL trading halt window(s):\n")
print(pal_halt_windows)
cat("(Expected: 2021-06-17 to 2022-03-27 -- PSE audit-opinion suspension)\n")

in_pal_halt <- function(dates) {
  purrr::map_lgl(dates, function(d) any(d >= pal_halt_windows$halt_start & d <= pal_halt_windows$halt_end))
}
```

```{r bea-master-df-levels-snapshot}
# Snapshot before log-returns converts every column to a return, so the raw levels survive for the values time series.
master_df_levels <- master_df
```

```{r log-returns}
master_df <- master_df %>%
  mutate(across(
    -Date, 
    ~ c(NA, diff(log(.x)))
  )) %>%
  rename_with(~ str_replace(.x, "\\.Adjusted$", ".LogAdj"), -Date)
```

```{r explore-master-df}
explore_instruments <- function(instruments, master_df) {
  for (sym in instruments) {
    cat("INSTRUMENT:", sym, "\n")
    
    instrument_df <- master_df %>% 
      select(Date, starts_with(sym))
    
    print(head(instrument_df), n = 3)
    print(tail(instrument_df), n = 3)
    cat("\n")
    
    cat("--- Missing Values per Column ---\n")
    instrument_df %>% 
      summarise(across(everything(), ~sum(is.na(.)))) %>% 
      print(width = Inf)
    cat("\n")
    
    missing_rows <- instrument_df %>%
      filter(if_any(starts_with(sym), is.na))
    
    total_missing <- nrow(missing_rows)
    cat("Total Rows with NAs:", total_missing, "\n")
    
    if (total_missing > 0) {
      cat("--- Top 10 Missing Rows ---\n")
      print(missing_rows, n = 10)
    }
    
    dup_count <- sum(duplicated(instrument_df$Date))
  }
  
  dplyr::glimpse(master_df)
}
explore_instruments(instruments, master_df)
```

```{r fix-missing-master-df}
master_df <- master_df %>%
  fill(-Date, .direction = "down") %>%
  filter(Date >= as.Date("2020-01-01"))

explore_instruments(instruments, master_df)
```

```{r bea-pal-masked-column}
# another column is created `PAL_masked.LogAdj` where its halted dates are masked to be NA
# this is based on PAL.LogAdj which all missing dates were forward filled,
# they will exist alongside each other, useful for more analysis later.
master_df <- master_df %>%
  mutate(PAL_masked.LogAdj = if_else(in_pal_halt(Date), NA_real_, PAL.LogAdj))

cat("PAL.LogAdj non-NA:", sum(!is.na(master_df$PAL.LogAdj)),
    "| PAL_masked.LogAdj non-NA:", sum(!is.na(master_df$PAL_masked.LogAdj)), "\n")
```
```{r}
# Helper function to map return columns back to level columns
level_col_name <- function(ret_col) {
  if (ret_col == "PAL_masked.LogAdj") return(NA_character_)
  if (str_detect(ret_col, "\\.LogAdj$")) return(str_replace(ret_col, "\\.LogAdj$", ".Adjusted"))
  return(ret_col)
}

target_cols <- c(
  "UAL.LogAdj",
  "PAL.LogAdj",
  "PAL_masked.LogAdj",
  "BTC-USD.LogAdj",
  "SPY.LogAdj",
  "VIX.LogAdj",
  "DGS10",
  "FEDFUNDS",
  "CPIAUCSL",
  "DCOILWTICO"
)

# Print Daily Summary Statistics
for (col in target_cols) {
  inst_data <- master_df %>%
    select(Date, Value = all_of(col)) %>%
    filter(!is.na(Value))

  summary_stats <- tibble(
    Instrument = col,
    Mean = mean(inst_data$Value),
    Median = median(inst_data$Value),
    Variance = var(inst_data$Value),
    Std_Dev = sd(inst_data$Value),
    Min = min(inst_data$Value),
    Max = max(inst_data$Value),
    Range = Max - Min,
    Skewness = skewness(inst_data$Value),
    Kurtosis = kurtosis(inst_data$Value)
  ) %>%
    mutate(across(where(is.numeric), ~round(., 4)))

  cat("\n======================================================\n")
  cat("### Daily Summary Statistics:", col, "\n")
  cat("======================================================\n")
  print(summary_stats)
  cat("\n")
}
```
```{r}
options(scipen = 999) # Prevent scientific notation on axes

target_cols <- c(
  "UAL.LogAdj", "PAL.LogAdj", "PAL_masked.LogAdj", "BTC-USD.LogAdj",
  "SPY.LogAdj", "VIX.LogAdj", "DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO"
)

# 1. Compile all statistics into a single dataframe
all_stats_list <- list()

for (col in target_cols) {
  inst_data <- master_df %>%
    select(Date, Value = all_of(col)) %>%
    filter(!is.na(Value))

  # Calculate stats for this column and store in the list
  all_stats_list[[col]] <- tibble(
    Instrument = col,
    Mean = mean(inst_data$Value, na.rm = TRUE),
    Median = median(inst_data$Value, na.rm = TRUE),
    Variance = var(inst_data$Value, na.rm = TRUE),
    Std_Dev = sd(inst_data$Value, na.rm = TRUE),
    Min = min(inst_data$Value, na.rm = TRUE),
    Max = max(inst_data$Value, na.rm = TRUE),
    Range = Max - Min,
    Skewness = skewness(inst_data$Value, na.rm = TRUE),
    Kurtosis = kurtosis(inst_data$Value, na.rm = TRUE)
  )
}

# Bind into one master table
all_stats_df <- bind_rows(all_stats_list)

# ====================================================================
# 2. STANDARD BAR CHARTS (Mean, Median, Var, SD, Skew, Kurtosis)
# ====================================================================

# Exclude Min, Max, and Range from the basic loop
standard_stats <- setdiff(names(all_stats_df), c("Instrument", "Min", "Max", "Range"))

for (stat in standard_stats) {
  
  p <- ggplot(all_stats_df, aes(x = reorder(Instrument, .data[[stat]]), y = .data[[stat]], fill = Instrument)) +
    geom_col(alpha = 0.8, color = "black", show.legend = FALSE) +
    # Add exact values at the end of each bar
    geom_text(aes(label = sprintf("%.4f", .data[[stat]]), 
                  y = ifelse(.data[[stat]] < 0, .data[[stat]] - max(abs(.data[[stat]]))*0.05, .data[[stat]] + max(abs(.data[[stat]]))*0.05)),
              size = 3.5, fontface = "bold", color = "black") +
    coord_flip() +
    theme_minimal() +
    labs(
      title = paste("Comparison of", stat, "(2020 - 2025)"),
      x = "Instrument",
      y = stat
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.15, 0.15))) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold"),
      axis.text.y = element_text(face = "bold")
    )
  
  print(p)
}

# ====================================================================
# 3. DIVERGING BAR CHART (Min, Max, Range)
# ====================================================================

# Create a new label combining Instrument and Range, then reorder it by Range
diverging_data <- all_stats_df %>%
  mutate(
    # Append the range to the instrument name for the axis label
    Instrument_Label = paste0(Instrument, " (Range: ", sprintf("%.4f", Range), ")"),
    Instrument_Label = reorder(Instrument_Label, Range)
  )

# Determine the absolute max value to ensure symmetrical Y-axis expansion
max_val <- max(abs(c(diverging_data$Min, diverging_data$Max))) * 1.25

p_min_max <- ggplot(diverging_data, aes(x = Instrument_Label)) +
  # Negative bars for Minimums (Tomato Red)
  geom_col(aes(y = Min), fill = "tomato", alpha = 0.8, color = "black") +
  # Positive bars for Maximums (Steel Blue)
  geom_col(aes(y = Max), fill = "steelblue", alpha = 0.8, color = "black") +
  
  # Text labels for exact Min values
  geom_text(aes(y = Min, label = sprintf("%.4f", Min)), 
            hjust = ifelse(diverging_data$Min < 0, 1.1, -0.1), # Shift depending on pos/neg
            size = 3.5, fontface = "bold", color = "black") +
            
  # Text labels for exact Max values
  geom_text(aes(y = Max, label = sprintf("%.4f", Max)), 
            hjust = ifelse(diverging_data$Max > 0, -0.1, 1.1), 
            size = 3.5, fontface = "bold", color = "black") +
             
  # Set symmetrical limits so zero stays exactly in the middle
  coord_flip(ylim = c(-max_val, max_val)) +
  theme_minimal() +
  labs(
    title = "Diverging Min & Max Values (2020 - 2025)",
    subtitle = "Sorted by Total Range. Min = Red, Max = Blue",
    x = "Instrument",
    y = "Value"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold")
  )

print(p_min_max)
```

```{r}
for (col in target_cols) {
  lvl_col <- level_col_name(col)
  
  if (!is.na(lvl_col) && lvl_col %in% names(master_df_levels)) {
    values_data <- master_df_levels %>% 
      select(Date, Value = all_of(lvl_col)) %>% 
      filter(!is.na(Value)) %>%
      arrange(Date) %>%
      # Calculate 30-day trailing moving average
      mutate(MA_30 = rollmean(Value, k = 30, fill = NA, align = "right"))
      
    values_plot <- ggplot(values_data, aes(x = Date)) +
      # Faint raw daily line
      geom_line(aes(y = Value), color = "darkgreen", alpha = 0.3, linewidth = 0.5) +
      # Bold 30-day Moving Average line
      geom_line(aes(y = MA_30), color = "darkgreen", alpha = 1, linewidth = 1.2) +
      theme_minimal() +
      labs(
        title = paste("Value Time Series:", lvl_col, "(2020 - 2025)"),
        subtitle = "30-Day Moving Average (Bold) vs Daily Value (Faint)",
        x = "Date",
        y = "Value"
      ) +
      theme(
        plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(face = "bold")
      )
      
    print(values_plot)
  }
}
```

```{r}
macro_cols <- c("DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO","VIX.LogAdj")
investable_cols <- target_cols[!target_cols %in% macro_cols]

for (col in investable_cols) {
  inst_data <- master_df %>%
    select(Date, Value = all_of(col)) %>%
    filter(!is.na(Value)) %>%
    arrange(Date) %>%
    # Calculate 30-day trailing moving average
    mutate(MA_30 = rollmean(Value, k = 30, fill = NA, align = "right"))

  return_plot <- ggplot(inst_data, aes(x = Date)) +
    # Faint raw daily return line
    geom_line(aes(y = Value), color = "darkblue", alpha = 0.3, linewidth = 0.5) +
    # Bold 30-day Moving Average line
    geom_line(aes(y = MA_30), color = "darkblue", alpha = 1, linewidth = 1.2) +
    theme_minimal() +
    labs(
      title = paste("Daily Timeline of", col, "(2020 - 2025)"),
      subtitle = "30-Day Moving Average (Bold) vs Daily Return (Faint)",
      x = "Date",
      y = "Log Return"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold")
    )

  print(return_plot)
}
```
```{r,fig.width=12}
options(scipen = 999) 

macro_cols <- c("DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO", "VIX.LogAdj", "PAL_masked.LogAdj")
investable_cols <- target_cols[!target_cols %in% macro_cols]

# 1. Combine all investable columns into a single long dataframe and calculate MA
long_data <- master_df %>%
  select(Date, all_of(investable_cols)) %>%
  pivot_longer(cols = all_of(investable_cols), names_to = "Asset", values_to = "Return") %>%
  filter(!is.na(Return)) %>%
  arrange(Asset, Date) %>%
  group_by(Asset) %>%
  # ADDED: rollapply with FUN = mean and partial = TRUE
  mutate(MA_90 = rollapply(Return, width = 90, FUN = mean, fill = NA, align = "right", partial = TRUE)) %>%
  ungroup()

# 2. Find the Leader (Max MA_90) for each Date
leader_data <- long_data %>%
  filter(!is.na(MA_90)) %>%
  group_by(Date) %>%
  # Keep the single highest MA for the day
  slice_max(order_by = MA_90, n = 1, with_ties = FALSE) %>%
  select(Date, Leader = Asset, Max_MA = MA_90) %>%
  ungroup()

# 3. Determine reasonable zoom limits based on the Moving Averages
min_ma <- min(long_data$MA_90, na.rm = TRUE)
max_ma <- max(long_data$MA_90, na.rm = TRUE)
# Add a 15% visual buffer above and below the moving averages
zoom_lower <- min_ma - abs(min_ma * 0.15)
zoom_upper <- max_ma + abs(max_ma * 0.15)

# 4. Create the Combined Plot
combined_plot <- ggplot() +
  # Leader Fill: Draw a colored rectangle for the leader from the bottom up to the MA line
  geom_rect(
    data = leader_data,
    aes(xmin = Date - 0.5, xmax = Date + 0.5, ymin = zoom_lower, ymax = Max_MA, fill = Leader),
    alpha = 0.25
  ) +
  # 90-Day Moving Average lines (bold)
  geom_line(
    data = long_data, 
    aes(x = Date, y = MA_90, color = Asset), 
    alpha = 1, linewidth = 1
  ) +
  theme_minimal() +
  # Zoom in to the moving average range
  coord_cartesian(ylim = c(zoom_lower, zoom_upper)) +
  labs(
    title = "Combined Asset 90-Day Moving Averages (2020 - 2025)",
    subtitle = "Shaded area highlights the asset with the highest 90-Day Moving Average. Early periods use <90 days.",
    x = "Date",
    y = "90-Day MA (Log Return)",
    color = "Asset (Lines)",
    fill = "Leader (Shaded Area)"
  ) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

print(combined_plot)
```

```{r}
options(scipen = 999) 

macro_cols <- c("DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO", "VIX.LogAdj", "PAL.LogAdj")
investable_cols <- target_cols[!target_cols %in% macro_cols]

quarterly_rank_data <- master_df %>%
  select(Date, all_of(investable_cols)) %>%
  pivot_longer(cols = all_of(investable_cols), names_to = "Asset", values_to = "Return") %>%
  filter(!is.na(Return)) %>%
  mutate(Quarter_Start = yq(paste0(year(Date), "-Q", quarter(Date)))) %>%
  group_by(Asset, Quarter_Start) %>%
  summarise(Quarterly_Return = sum(Return, na.rm = TRUE), .groups = "drop") %>%
  group_by(Quarter_Start) %>%
  mutate(Rank = rank(-Quarterly_Return, ties.method = "first")) %>%
  ungroup()

rank_plot <- ggplot(quarterly_rank_data, aes(x = Quarter_Start, y = Rank, color = Asset, group = Asset)) +
  geom_line(linewidth = 1.5, alpha = 0.8) +
  geom_point(size = 5) +
  # Place the rank number inside the point for readability
  geom_text(aes(label = Rank), color = "white", size = 3, fontface = "bold") +
  # Reverse the Y-axis so Rank 1 is at the top
  scale_y_reverse(breaks = 1:length(investable_cols)) +
  # Format X-axis to show Year-Quarter intervals clearly
  scale_x_date(date_breaks = "3 months", labels = function(x) paste0(year(x), "-Q", quarter(x))) +
  theme_minimal() +
  labs(
    title = "Quarterly Asset Performance Rankings",
    subtitle = "Ranked by cumulative log return per quarter (1 = Best)",
    x = "Quarter",
    y = "Rank",
    color = "Asset"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor.y = element_blank(), # Removes the decimal grid lines (e.g., 1.5, 2.5)
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

print(rank_plot)
```

```{r, fig.width=12}
options(scipen = 999) 

macro_cols <- c("DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO", "VIX.LogAdj", "PAL_masked.LogAdj")
investable_cols <- target_cols[!target_cols %in% macro_cols]

# 1. Combine all investable columns and calculate Rolling Risk (Standard Deviation)
long_data <- master_df %>%
  select(Date, all_of(investable_cols)) %>%
  pivot_longer(cols = all_of(investable_cols), names_to = "Asset", values_to = "Return") %>%
  filter(!is.na(Return)) %>%
  arrange(Asset, Date) %>%
  group_by(Asset) %>%
  # Use rollapply with sd (standard deviation) to measure volatility/risk
  mutate(Rolling_Risk = rollapply(Return, width = 90, FUN = sd, fill = NA, align = "right", partial = TRUE)) %>%
  ungroup()

# 2. Find the Leader (Highest Risk) for each Date
leader_data <- long_data %>%
  filter(!is.na(Rolling_Risk)) %>%
  group_by(Date) %>%
  # Keep the single highest risk asset for the day
  slice_max(order_by = Rolling_Risk, n = 1, with_ties = FALSE) %>%
  select(Date, Leader = Asset, Max_Risk = Rolling_Risk) %>%
  ungroup()

# 3. Determine reasonable zoom limits based on the Rolling Risk
min_risk <- min(long_data$Rolling_Risk, na.rm = TRUE)
max_risk <- max(long_data$Rolling_Risk, na.rm = TRUE)
# Add a 15% visual buffer, but ensure the bottom doesn't drop below 0 (risk is strictly positive)
zoom_lower <- max(0, min_risk - abs(min_risk * 0.15))
zoom_upper <- max_risk + abs(max_risk * 0.15)

# 4. Create the Combined Plot
combined_plot <- ggplot() +
  # Leader Fill: Draw a colored rectangle for the most volatile asset
  geom_rect(
    data = leader_data,
    aes(xmin = Date - 0.5, xmax = Date + 0.5, ymin = zoom_lower, ymax = Max_Risk, fill = Leader),
    alpha = 0.25
  ) +
  # 90-Day Rolling Risk lines (bold)
  geom_line(
    data = long_data, 
    aes(x = Date, y = Rolling_Risk, color = Asset), 
    alpha = 1, linewidth = 1
  ) +
  theme_minimal() +
  # Zoom in to the risk range
  coord_cartesian(ylim = c(zoom_lower, zoom_upper)) +
  labs(
    title = "Combined Asset Rolling Risk (90-Day Standard Deviation)",
    subtitle = "Shaded area under the curve highlights the most volatile asset",
    x = "Date",
    y = "90-Day Rolling SD (Risk)",
    color = "Asset (Lines)",
    fill = "Highest Risk (Shaded Area)"
  ) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

print(combined_plot)
```

```{r}
options(scipen = 999) 

macro_cols <- c("DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO", "VIX.LogAdj", "PAL.LogAdj")
investable_cols <- target_cols[!target_cols %in% macro_cols]

quarterly_risk_rank_data <- master_df %>%
  select(Date, all_of(investable_cols)) %>%
  pivot_longer(cols = all_of(investable_cols), names_to = "Asset", values_to = "Return") %>%
  filter(!is.na(Return)) %>%
  mutate(Quarter_Start = yq(paste0(year(Date), "-Q", quarter(Date)))) %>%
  group_by(Asset, Quarter_Start) %>%
  # Calculate risk (standard deviation of daily returns) for the quarter
  summarise(Quarterly_Risk = sd(Return, na.rm = TRUE), .groups = "drop") %>%
  group_by(Quarter_Start) %>%
  # Rank assets for each quarter (ascending order: lowest SD gets Rank 1)
  mutate(Rank = rank(Quarterly_Risk, ties.method = "first")) %>%
  ungroup()

risk_rank_plot <- ggplot(quarterly_risk_rank_data, aes(x = Quarter_Start, y = Rank, color = Asset, group = Asset)) +
  geom_line(linewidth = 1.5, alpha = 0.8) +
  geom_point(size = 5) +
  # Place the rank number inside the point for readability
  geom_text(aes(label = Rank), color = "white", size = 3, fontface = "bold") +
  # Reverse the Y-axis so Rank 1 (Least Risk) is at the top
  scale_y_reverse(breaks = 1:length(investable_cols)) +
  # Format X-axis to show Year-Quarter intervals clearly
  scale_x_date(date_breaks = "3 months", labels = function(x) paste0(year(x), "-Q", quarter(x))) +
  theme_minimal() +
  labs(
    title = "Quarterly Asset Risk Rankings",
    subtitle = "Ranked by standard deviation of daily log returns per quarter (1 = Least Risk)",
    x = "Quarter",
    y = "Rank",
    color = "Asset"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor.y = element_blank(), # Removes the decimal grid lines
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

print(risk_rank_plot)
```

```{r distribution}
for (col in target_cols) {
  inst_data <- master_df %>%
    select(Date, Value = all_of(col)) %>%
    filter(!is.na(Value)) %>%
    # Standardize the data to Z-scores: (Value - Mean) / SD
    mutate(Standardized_Value = as.numeric(scale(Value)))

  dist_plot <- ggplot(inst_data, aes(x = Standardized_Value)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "steelblue", color = "white", alpha = 0.7) +
    geom_density(color = "darkred", linewidth = 1) +
    theme_minimal() +
    labs(
      title = paste("Standardized Distribution of", col, "(2020 - 2025)"),
      subtitle = "Z-Scores: Data centered at 0 with a Standard Deviation of 1",
      x = "Standardized Value (Z-Score)",
      y = "Density"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold")
    )

  print(dist_plot)
}

long_dist_data <- master_df %>%
  # Select only the target columns
  select(all_of(target_cols)) %>%
  # Pivot to long format so we can group by Asset
  pivot_longer(cols = everything(), names_to = "Asset", values_to = "Value") %>%
  filter(!is.na(Value)) %>%
  group_by(Asset) %>%
  # Standardize the data to Z-scores: (Value - Mean) / SD
  mutate(Standardized_Value = as.numeric(scale(Value))) %>%
  ungroup()

# 2. Plot all standardized distributions in one graph
combined_dist_plot <- ggplot(long_dist_data, aes(x = Standardized_Value, fill = Asset, color = Asset)) +
  # Using density instead of histogram so overlapping distributions remain readable
  geom_density(alpha = 0.25, linewidth = 0.7) +
  theme_minimal() +
  # Constrain the X-axis to +/- 5 standard deviations. 
  # This prevents massive one-off outliers (like crypto crashes) from squashing the visual bell curves
  coord_cartesian(xlim = c(-5, 5)) +
  labs(
    title = "Combined Standardized Distributions (2020 - 2025)",
    subtitle = "Z-Scores: Data centered at 0 with a Standard Deviation of 1",
    x = "Standardized Value (Z-Score)",
    y = "Density",
    fill = "Instrument",
    color = "Instrument"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

print(combined_dist_plot)
```

```{r boxplot-comparison}

options(scipen = 999) 

for (col in target_cols) {
  # 1. Prepare data
  if (col %in% c("CPIAUCSL", "FEDFUNDS")) {
    box_data <- df_list[[col]] %>%
      arrange(Date) %>%
      transmute(Date, Value = c(NA, diff(log(.data[[col]])))) %>%
      filter(!is.na(Value))
    plot_label <- paste("Box Plot:", col, "(2020 - 2025)")
  } else {
    box_data <- master_df %>% select(Date, Value = all_of(col)) %>% filter(!is.na(Value))
    plot_label <- paste("Box Plot:", col, "(2020 - 2025)")
  }

  # 2. Calculate summary statistics
  stats <- box_data %>%
    summarise(
      Min    = min(Value, na.rm = TRUE),
      Q1     = quantile(Value, 0.25, na.rm = TRUE),
      Median = median(Value, na.rm = TRUE),
      Mean   = mean(Value, na.rm = TRUE),
      Q3     = quantile(Value, 0.75, na.rm = TRUE),
      Max    = max(Value, na.rm = TRUE)
    )
  
  iqr <- stats$Q3 - stats$Q1
  
  # 3. Create a label dataframe for ggrepel to process all at once
  # Using sprintf ensures strictly formatted 4-decimal floats instead of scientific notation
  label_data <- data.frame(
    x_pos = factor(""), 
    y_pos = c(stats$Max, stats$Q3, stats$Median, stats$Mean, stats$Q1, stats$Min),
    label_text = c(
      paste("Max:", sprintf("%.4f", stats$Max)),
      paste("Q3:", sprintf("%.4f", stats$Q3)),
      paste("Med:", sprintf("%.4f", stats$Median)),
      paste("Mean:", sprintf("%.4f", stats$Mean)),
      paste("Q1:", sprintf("%.4f", stats$Q1)),
      paste("Min:", sprintf("%.4f", stats$Min))
    )
  )

  # 4. Generate the plot
  box_plot <- ggplot(box_data, aes(x = factor(""), y = Value)) +
    geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.color = "darkred", outlier.alpha = 0.4) +
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "gold", color = "black") +
    coord_cartesian(ylim = c(stats$Q1 - 3 * iqr, stats$Q3 + 3 * iqr)) +
    
    # Use ggrepel to place labels. 
    # nudge_x pushes them to the right; direction = "y" keeps them vertically stacked
    geom_text_repel(
      data = label_data,
      aes(x = x_pos, y = y_pos, label = label_text),
      nudge_x = 0.45,
      direction = "y",
      hjust = 0,
      segment.color = "darkgray", # Draws a faint line if the label has to move far
      segment.size = 0.5,
      size = 3.5,
      fontface = c("plain", "plain", "bold", "bold", "plain", "plain"),
      color = c("darkgray", "black", "black", "darkgoldenrod4", "black", "darkgray")
    ) +
    
    theme_minimal() +
    labs(
      title = plot_label,
      subtitle = "Black line = median, gold diamond = mean",
      x = NULL,
      y = "Log Return"
    ) +
    # Force the Y-axis to also avoid scientific notation
    scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
    # Expand the right side of the plot to make room for repelled labels
    scale_x_discrete(expand = expansion(add = c(0.5, 1.5))) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold"),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    )

  print(box_plot)
}
```
Forward-filling PAL's halted trading dates introduces extra zero-return days, which compresses the interquartile range: IQR narrows from 0.0173 in the halt-masked version to 0.0130 in the naive forward-filled version (a ~25% reduction), while the median stays at 0 in both.

```{r name-df-list}
# df_list was built in instrument order in `convert-to-df` (pre full_join, pre fill).
# Name it so individual per-instrument tibbles can be pulled out by ticker for the
# join demonstrations below, instead of re-splitting the already-joined master_df.
names(df_list) <- instruments
```

```{r bea-leftjoin1-pre-obs fuel-pass-through}
cat("UAL:", nrow(df_list[["UAL"]]), "rows |",
    format(min(df_list[["UAL"]]$Date)), "to", format(max(df_list[["UAL"]]$Date)), "\n")
cat("PAL:", nrow(df_list[["PAL"]]), "rows |",
    format(min(df_list[["PAL"]]$Date)), "to", format(max(df_list[["PAL"]]$Date)), "\n")
cat("DCOILWTICO:", nrow(df_list[["DCOILWTICO"]]), "rows |",
    format(min(df_list[["DCOILWTICO"]]$Date)), "to", format(max(df_list[["DCOILWTICO"]]$Date)), "\n\n")
cat("Note: PSE (PAL) and NYMEX (WTI) calendars differ from NYSE (UAL).\n")
cat("A left_join on UAL's calendar will introduce NAs on days PAL/WTI didn't print,\n")
cat("which we carry forward via fill() rather than dropping UAL rows.\n")
```
The difference in rows could be explained by the following:
- UAL is based on NYSE weekday calendar, which is closed on weekends and US market holidays. 
- PAL is based on PSE weekday calendar, which is closed on weekends and PH holidays.
- PAL had a 285-day trading suspension imposed by the PSE between (2021-06-17 to 2022-03-28),
  as its financial statements had violated a rule. This was during COVID-19 travel restrictions.
- DCOILWTICO is from FRED instead of a stock market trading calendar. 
```{r bea-leftjoin1}
fuel_leftjoin_df <- df_list[["UAL"]] %>%
  select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  left_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  left_join(df_list[["DCOILWTICO"]] %>% select(Date, WTI = DCOILWTICO), by = "Date") %>%
  arrange(Date) %>%
  fill(PAL_Adjusted, WTI, .direction = "down") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted))),
    WTI_ret = c(NA, diff(log(WTI)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(WTI_ret))

cat("Rows kept (UAL trading calendar):", nrow(fuel_leftjoin_df), "\n\n")
print(fuel_leftjoin_df, n = 6)

ual_fuel_model <- lm(UAL_ret ~ WTI_ret, data = fuel_leftjoin_df)
pal_fuel_model <- lm(PAL_ret ~ WTI_ret, data = fuel_leftjoin_df)
print(summary(ual_fuel_model)$coefficients)
print(summary(pal_fuel_model)$coefficients)
```

```{r bea-leftjoin1-post-obs}
ual_oil_beta <- coef(ual_fuel_model)[["WTI_ret"]]
pal_oil_beta <- coef(pal_fuel_model)[["WTI_ret"]]
ual_oil_p <- summary(ual_fuel_model)$coefficients["WTI_ret", "Pr(>|t|)"]
pal_oil_p <- summary(pal_fuel_model)$coefficients["WTI_ret", "Pr(>|t|)"]

cat("Rows retained on UAL's calendar:", nrow(fuel_leftjoin_df), "\n")
cat("UAL oil-return coefficient:", round(ual_oil_beta, 4), "| p-value:", round(ual_oil_p, 4), "\n")
cat("PAL oil-return coefficient:", round(pal_oil_beta, 4), "| p-value:", round(pal_oil_p, 4), "\n")
```

- `fuel_leftjoin_df` - PAL's price forward-filled across every gap (including weekends, holidays, and the trading suspension)
- `fuel_leftjoin_df_masked` - weekends and holidays are forward-filled but the trading suspension is 
```{r bea-pal-masked-companion}

# dataframe where the trading halt (285-day suspension) are masked as NA
# this is built from the fuel_leftjoin_df which is forward-filled
fuel_leftjoin_df_masked <- fuel_leftjoin_df %>%
  mutate(PAL_ret = if_else(in_pal_halt(Date), NA_real_, PAL_ret)) %>%
  filter(!is.na(PAL_ret))

pal_fuel_model_masked <- lm(PAL_ret ~ WTI_ret, data = fuel_leftjoin_df_masked)
pal_oil_beta_masked <- coef(pal_fuel_model_masked)[["WTI_ret"]]
pal_oil_p_masked <- summary(pal_fuel_model_masked)$coefficients["WTI_ret", "Pr(>|t|)"]

cat("Naive (forward-filled) rows:", nrow(fuel_leftjoin_df), "| PAL oil-beta:", round(pal_oil_beta, 4), "\n")
cat("Masked (halt-excluded) rows:", nrow(fuel_leftjoin_df_masked), "| PAL oil-beta:", round(pal_oil_beta_masked, 4), "\n")
```

```{r bea-innerjoin-pre-obs BTC/airline correlation}
cat("BTC-USD:", nrow(df_list[["BTC-USD"]]), "rows | UAL:", nrow(df_list[["UAL"]]),
    "rows | PAL:", nrow(df_list[["PAL"]]), "rows\n")
cat("inner_join keeps only the intersection -- every BTC weekend/holiday move is\n")
cat("excluded by construction; loss is quantified explicitly below.\n")
```

```{r bea-innerjoin}
btc_innerjoin_df <- df_list[["BTC-USD"]] %>%
  select(Date, BTC_Adjusted = `BTC-USD.Adjusted`) %>%
  inner_join(df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`), by = "Date") %>%
  inner_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  arrange(Date) %>%
  mutate(
    BTC_ret = c(NA, diff(log(BTC_Adjusted))),
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(BTC_ret), !is.na(UAL_ret), !is.na(PAL_ret))

n_btc_total <- nrow(df_list[["BTC-USD"]])
n_common <- nrow(btc_innerjoin_df)
cor_matrix <- cor(btc_innerjoin_df %>% select(BTC_ret, UAL_ret, PAL_ret), use = "complete.obs")
print(round(cor_matrix, 4))
```

```{r bea-innerjoin-post-obs}
btc_ual_cor <- cor_matrix["BTC_ret", "UAL_ret"]
btc_pal_cor <- cor_matrix["BTC_ret", "PAL_ret"]
ual_pal_cor <- cor_matrix["UAL_ret", "PAL_ret"]

cat("Common trading days retained:", n_common, "of", n_btc_total, "BTC calendar days\n")
cat("BTC-UAL correlation:", round(btc_ual_cor, 4), "\n")
cat("BTC-PAL correlation:", round(btc_pal_cor, 4), "\n")
cat("UAL-PAL correlation:", round(ual_pal_cor, 4), "\n")
```

```{r bea-antijoin-pre-obs}
cat("PAL:", nrow(df_list[["PAL"]]), "rows | UAL:", nrow(df_list[["UAL"]]), "rows\n")
cat("anti_join surfaces every UAL trading day with no matching PAL date --\n")
cat("both routine PSE holidays and the 285-day suspension. Splitting the two\n")
cat("apart tests whether the suspension or routine calendar drift dominates.\n")
```

```{r bea-antijoin pal-ual-gap}
pal_gap_days <- df_list[["UAL"]] %>% select(Date) %>%
  anti_join(df_list[["PAL"]] %>% select(Date), by = "Date")

holiday_gap_days <- pal_gap_days %>% filter(!in_pal_halt(Date))
halted_gap_days  <- pal_gap_days %>% filter(in_pal_halt(Date))

cat("Total UAL trading days with no PAL match:", nrow(pal_gap_days), "\n")
cat("Routine PSE-holiday-only days:", nrow(holiday_gap_days), "\n")
cat("Suspension-only days:", nrow(halted_gap_days), "\n\n")
print(holiday_gap_days, n = 10)
```

```{r bea-antijoin-post-obs}
cat("Of", nrow(pal_gap_days), "total PAL gap days, which are dates that are in UAL but are not in the PAL dataset,", nrow(halted_gap_days),
    sprintf("(%.0f%%)", 100 * nrow(halted_gap_days) / nrow(pal_gap_days)),
    "are attributable to the 285-day suspension. Note that the 285-day suspension is the total days between 2021-06-17 to 2022-03-27, which includes weekends, but because PAL and UAL does not include weekends, which is why only 196 is attributable to the suspension. \n")
cat("Routine PSE-holiday mismatch accounts for only", nrow(holiday_gap_days), "days, which accounts for holidays starting from December 2019 to Decemer 2025. These are dates that PAL does not account for, but is included in the UAL dataset.\n")
```

# SQL-Based Financial Investigation

## SQL Investigation 1: Does Forward-Filling PAL's Suspension Bias Its Return Statistics?

**Objective:** Determine whether treating PAL's 285-day trading suspension as if its price simply stayed unchanged distorts its measured returns and volatility, compared to properly excluding that period from the calculation.

```{r bea-sql1}
pal_stats_comparison <- bind_rows(
  fuel_leftjoin_df        %>% mutate(version = "Forward-filled (naive)") %>% select(version, PAL_ret),
  fuel_leftjoin_df_masked %>% mutate(version = "Halt-masked (corrected)") %>% select(version, PAL_ret)
) %>%
  group_by(version) %>%
  summarise(
    N = n(),
    Zero_Return_Pct = round(100 * mean(PAL_ret == 0), 1),
    Mean = round(mean(PAL_ret), 6),
    Median = round(median(PAL_ret), 6),
    Variance = round(var(PAL_ret), 6),
    SD = round(sd(PAL_ret), 4),
    Min = round(min(PAL_ret), 4),
    Max = round(max(PAL_ret), 4),
    Range = round(max(PAL_ret) - min(PAL_ret), 4),
    Skewness = round(skewness(PAL_ret), 3),
    Kurtosis = round(kurtosis(PAL_ret), 3),
    .groups = "drop"
  )

print(pal_stats_comparison, width = Inf)
```

**Financial Interpretation:** Forward-filling raises PAL's share of zero-return days from 25.8% to 35.5% (about 1.4x), this pulls its standard deviation down (0.0252 to 0.0235) and inflating kurtosis (13.1 to 15.1). The halt gets recorded as "196 calm trading days" compared to "no market existed." 

In practical terms, anyone using the naive series to size a PAL position or estimate its downside risk would be working from an artificially calm risk profile: a standard deviation of 0.0235 instead of the true 0.0252, and a kurtosis reading of 15.1 instead of 13.1 that overstates how rare PAL's extreme moves really are. Both numbers move the same direction because forward-filling those 196 days manufactures calm on exactly the days when the real series was actually more volatile and more prone to extreme moves. For anyone sizing a position or setting a stop-loss off the naive numbers, that translates into a genuine underestimate of risk.

## SQL Investigation 2: Which Airline Is More Exposed to Oil: UAL or PAL?

**Objective:** Determine which airline, United or Philippine Airlines, carries more fuel-cost risk by comparing their sensitivity to crude oil price movements.

```{r bea-sql2}
oil_direction_summary <- fuel_leftjoin_df_masked %>%
  mutate(oil_direction = if_else(WTI_ret > 0, "Oil Up", "Oil Down")) %>%
  group_by(oil_direction) %>%
  summarise(
    n_days = n(),
    mean_UAL_ret = round(mean(UAL_ret), 5),
    mean_PAL_ret = round(mean(PAL_ret), 5),
    .groups = "drop"
  ) %>%
  arrange(oil_direction)

print(oil_direction_summary)

cat("\nFull-sample regression betas (UAL from bea-leftjoin1; PAL from the masked companion):\n")
cat("UAL oil-beta:", round(ual_oil_beta, 4), "| p =", round(ual_oil_p, 4), "\n")
cat("PAL oil-beta:", round(pal_oil_beta_masked, 4), "| p =", round(pal_oil_p_masked, 4), "\n")
```

**Financial Interpretation:** Interestingly, although UAL's beta is *larger* in magnitude (+0.104) but *positively* signed, signifying that it moves with oil, consistent with oil and airline stocks both being pro-cyclical/demand-driven in a simple two-variable regression with no macro controls. PAL's beta is smaller (-0.078) but *negatively* signed, implying that the "rising fuel cost hurts an airline". So UAL shows the stronger statistical co-movement, but PAL shows the theoretically correct cost-channel relationship. 

## SQL Investigation 3: The Ukraine Invasion Oil Shock

**Objective:** Examine how United Airlines' sensitivity to oil prices behaved during the acute 2022 Russia-Ukraine oil shock (February 24-March 8), and assess the impact of Philippine Airlines being unable to trade during this same window.

```{r bea-sql3}
ukraine_shock_df_all_ual <- fuel_leftjoin_df %>%
  filter(between(Date, as.Date("2022-02-24"), as.Date("2022-03-08")))
ukraine_shock_df_pal <- fuel_leftjoin_df_masked %>%
  filter(between(Date, as.Date("2022-02-24"), as.Date("2022-03-08")))

cat("UAL trading days in Ukraine shock window:", nrow(ukraine_shock_df_all_ual), "\n")
cat("Of which PAL has real (non-halted) data:", nrow(ukraine_shock_df_pal), "\n\n")

shock_vs_full <- tibble(
  Period = c("Full sample", "Ukraine shock (2022-02-24 to 2022-03-08)"),
  N = c(nrow(fuel_leftjoin_df), nrow(ukraine_shock_df_all_ual)),
  UAL_oil_beta = c(
    round(cov(fuel_leftjoin_df$UAL_ret, fuel_leftjoin_df$WTI_ret, use = "complete.obs") / var(fuel_leftjoin_df$WTI_ret, na.rm = TRUE), 4),
    round(cov(ukraine_shock_df_all_ual$UAL_ret, ukraine_shock_df_all_ual$WTI_ret, use = "complete.obs") / var(ukraine_shock_df_all_ual$WTI_ret, na.rm = TRUE), 4)
  ),
  Mean_UAL_ret = c(round(mean(fuel_leftjoin_df$UAL_ret), 5), round(mean(ukraine_shock_df_all_ual$UAL_ret), 5)),
  Mean_WTI_ret = c(round(mean(fuel_leftjoin_df$WTI_ret), 5), round(mean(ukraine_shock_df_all_ual$WTI_ret), 5))
)
print(shock_vs_full, width = Inf)
```

**Financial Interpretation:** PAL has zero trading days in this window, its suspension (2021-06-17 to 2022-03-27) covers the entire shock. UAL investors could reprice the stock in real time against the oil spike; PAL investors could not trade at any price, favorable or not. That absence is itself the finding: a real-world illustration of market friction/information asymmetry, not just a data-cleaning footnote. UAL's own beta is also striking here: it *flips sign*, from +0.104 full-sample to -0.417 during the shock, alongside a -3.4% average daily return over those 9 days. That's consistent with the invasion being a broad risk-off/war-fear event that hit airline stocks directly (falling, alongside rising oil), temporarily overwhelming the smaller demand-channel co-movement seen in the full-sample average. (With only 9 trading days behind it, this specific beta shouldn't be treated as a precise, tradeable number, but the direction and size of the swing support the broader point: a real oil shock can temporarily overwhelm an airline's typical demand-channel behavior.)

## SQL Investigation 4: Is UAL's Oil-Beta a Stable Relationship, or Just the Ukraine Shock?

**Objective:** Determine whether the fuel-cost relationship identified for United and Philippine Airlines is a stable, ongoing pattern or simply a result of the 2022 oil shock, by comparing pre-shock, shock-era, and recovery periods.

```{r bea-sql4}
period_label <- function(d) case_when(
  d < as.Date("2022-01-01") ~ "1. Pre-shock (2020-2021)",
  d < as.Date("2023-01-01") ~ "2. Shock era (2022)",
  TRUE ~ "3. Recovery (2023-2025)"
)

ual_period_beta <- fuel_leftjoin_df %>%
  mutate(period = period_label(Date)) %>%
  group_by(period) %>%
  summarise(
    n_days_ual = n(),
    UAL_beta = round(cov(UAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

pal_period_beta <- fuel_leftjoin_df_masked %>%
  mutate(period = period_label(Date)) %>%
  group_by(period) %>%
  summarise(
    n_days_pal = n(),
    PAL_beta = round(cov(PAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

period_beta <- ual_period_beta %>%
  left_join(pal_period_beta, by = "period") %>%
  arrange(period)

print(period_beta, width = Inf)
```

**Financial Interpretation:** UAL's oil-beta is not a stable relationship. It swings from +0.178 pre-shock to -0.155 during the shock era to -0.020 in the recovery period, meaning Investigation 2's full-sample beta (+0.104) is really a blended average across three very different regimes, not something an investor could reliably trade on as a fixed relationship. PAL's beta, by contrast, stays negative throughout (-0.091, -0.083, -0.020): the fuel-cost story holds up more consistently for PAL than the demand-channel story does for UAL, though PAL's shock-era reading rests on a noticeably smaller sample (193 trading days) since the suspension cuts directly into that period. Practically, this means UAL's oil exposure is regime-dependent and needs to be reassessed as market conditions shift, whereas PAL's fuel-cost relationship is the steadier of the two.

## SQL Investigation 5: How Much Does PAL's Stagnant Period Deflate the UAL-PAL Correlation?

**Objective:** Measure how much the correlation between United and Philippine Airlines' stock returns changes depending on how the two datasets are combined.

```{r bea-sql5}
ual_pal_fulljoin_df <- df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  full_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  arrange(Date) %>%
  fill(UAL_Adjusted, PAL_Adjusted, .direction = "down") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(PAL_ret))

ual_pal_innerjoin_df <- df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  inner_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  arrange(Date) %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(PAL_ret))

naive_cor  <- cor(ual_pal_fulljoin_df$UAL_ret, ual_pal_fulljoin_df$PAL_ret, use = "complete.obs")
strict_cor <- cor(ual_pal_innerjoin_df$UAL_ret, ual_pal_innerjoin_df$PAL_ret, use = "complete.obs")

cat("Naive full_join + fill() correlation:", round(naive_cor, 4), "| N =", nrow(ual_pal_fulljoin_df), "\n")
cat("Strict inner_join correlation:", round(strict_cor, 4), "| N =", nrow(ual_pal_innerjoin_df), "\n")
```

**Financial Interpretation:** The naive version reports a UAL-PAL correlation of just 0.0355, built on 1539 rows, many of them repeated/fabricated values from the forward-filled suspension rather than real paired price movements. Using only the 1206 days both stocks actually traded, the true correlation is 0.0759, more than double the naive estimate. That gap matters for portfolio decisions: relying on the naive number would lead you to conclude UAL and PAL offer more diversification benefit together than they actually do, understating how much their returns really move together. The true relationship is still fairly weak in absolute terms (two airlines on opposite sides of the world with very different local demand drivers), but it's meaningfully stronger than the naive join suggests. This is the same deflation mechanism as Investigation 1, just showing up in a pairwise statistic instead of a single asset's own volatility.

## SQL Investigation 6: Do UAL and PAL Behave Differently in High-Inflation Periods?

**Objective:** Determine whether United and Philippine Airlines perform differently during periods of high inflation versus periods of low or moderate inflation.

```{r bea-sql6}
cpi_daily <- tibble(Date = seq(min(df_list[["CPIAUCSL"]]$Date), max(fuel_leftjoin_df$Date), by = "day")) %>%
  left_join(
    df_list[["CPIAUCSL"]] %>% arrange(Date) %>% mutate(CPI_YoY = CPIAUCSL / lag(CPIAUCSL, 12) - 1) %>% select(Date, CPI_YoY),
    by = "Date"
  ) %>%
  arrange(Date) %>%
  fill(CPI_YoY, .direction = "down")

label_inflation <- function(df) {
  df %>%
    left_join(cpi_daily, by = "Date") %>%
    filter(!is.na(CPI_YoY)) %>%
    mutate(inflation_regime = if_else(CPI_YoY > 0.04, "High inflation (>4% YoY)", "Low/moderate inflation (<=4% YoY)"))
}

ual_inflation_df <- fuel_leftjoin_df %>% label_inflation()
pal_inflation_df <- fuel_leftjoin_df_masked %>% label_inflation()

ual_inflation_summary <- bind_rows(
  ual_inflation_df %>% mutate(inflation_regime = "Overall (all regimes)"),
  ual_inflation_df
) %>%
  group_by(inflation_regime) %>%
  summarise(n_days_ual = n(), mean_UAL_ret = round(mean(UAL_ret), 5), sd_UAL_ret = round(sd(UAL_ret), 5), .groups = "drop")

pal_inflation_summary <- bind_rows(
  pal_inflation_df %>% mutate(inflation_regime = "Overall (all regimes)"),
  pal_inflation_df
) %>%
  group_by(inflation_regime) %>%
  summarise(n_days_pal = n(), mean_PAL_ret = round(mean(PAL_ret), 5), sd_PAL_ret = round(sd(PAL_ret), 5), .groups = "drop")

inflation_summary <- ual_inflation_summary %>%
  left_join(pal_inflation_summary, by = "inflation_regime") %>%
  arrange(inflation_regime)

print(inflation_summary, width = Inf)
```

**Financial Interpretation:** UAL's average daily return flips from +0.15% when inflation is low or moderate to -0.04% once inflation climbs above 4%. The swing from positive to negative lines up with intuition, that higher inflation cuts into consumer travel budgets while also raising UAL's operating costs like fuel, wages, and leases. PAL doesn't show the same pattern; its average return is negative in both regimes (-0.07% and -0.03%), so inflation on its own doesn't appear to be a major driver for PAL specifically. Another important thing to note is how PAL's high-inflation bucket only has 349 real trading days against UAL's 545, since that stretch overlaps heavily with the suspension, so there's less to go on there. Volatility barely changes for either airline across the two regimes, so the real story is a shift in average return, not the stock getting riskier.

## SQL Investigation 7: Does the Fed Funds Rate Environment Change UAL's and PAL's Oil-Beta?

**Objective:** Determine how the prevailing interest rate environment (near-zero, hiking, or restrictive) affects both airlines' stock performance and their sensitivity to oil prices.

```{r bea-sql7}
fedfunds_daily <- tibble(Date = seq(min(df_list[["FEDFUNDS"]]$Date), max(fuel_leftjoin_df$Date), by = "day")) %>%
  left_join(df_list[["FEDFUNDS"]] %>% select(Date, FedFunds = FEDFUNDS), by = "Date") %>%
  arrange(Date) %>%
  fill(FedFunds, .direction = "down")

label_rate_regime <- function(df) {
  df %>%
    left_join(fedfunds_daily, by = "Date") %>%
    filter(!is.na(FedFunds)) %>%
    mutate(rate_regime = case_when(
      FedFunds < 1 ~ "1. Near-zero (<1%)",
      FedFunds < 4 ~ "2. Hiking (1-4%)",
      TRUE ~ "3. Restrictive (>4%)"
    ))
}

ual_rate_df <- fuel_leftjoin_df %>% label_rate_regime()
pal_rate_df <- fuel_leftjoin_df_masked %>% label_rate_regime()

ual_rate_summary <- bind_rows(
  ual_rate_df %>% mutate(rate_regime = "Overall (all regimes)"),
  ual_rate_df
) %>%
  group_by(rate_regime) %>%
  summarise(
    n_days_ual = n(),
    mean_UAL_ret = round(mean(UAL_ret), 5),
    UAL_oil_beta = round(cov(UAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

pal_rate_summary <- bind_rows(
  pal_rate_df %>% mutate(rate_regime = "Overall (all regimes)"),
  pal_rate_df
) %>%
  group_by(rate_regime) %>%
  summarise(
    n_days_pal = n(),
    mean_PAL_ret = round(mean(PAL_ret), 5),
    PAL_oil_beta = round(cov(PAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

rate_summary <- ual_rate_summary %>%
  left_join(pal_rate_summary, by = "rate_regime") %>%
  arrange(rate_regime)

print(rate_summary, width = Inf)
```

**Financial Interpretation:** Breaking the sample out by Fed policy stance shows UAL's relationship with oil fading rather than flipping: +0.131 when rates are near zero, down to +0.074 during the hiking phase, and basically flat at -0.009 once policy turns restrictive. Oil and UAL's stock gradually stop moving together as the Fed tightens, rather than starting to move in opposite directions. PAL tells almost the same story in reverse: its beta goes from -0.095 (near-zero) to -0.059 (hiking) to -0.015 (restrictive), meaning its fuel-cost relationship is strongest exactly when money is cheap and weakens as rates climb. Put the two together and both airlines' link to oil shrinks as rates rise, regardless of whether that link starts positive or negative, which matters for anyone using oil as a hedge or signal for either stock: that relationship isn't constant, it depends on where we are in the rate cycle.
