---
title: "Main"
author: "Roan Campo, Denzel Macayan, Bea Uy, Miguel Uy"
date: "2026-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-environment-dependencies}
required_packages <- c(
  "tidyr", "dplyr", "tibble", "purrr", "moments", "quantmod", "reticulate", "httr", "jsonlite", "xts", "readr","stringr", "ggplot2", "quantmod"
)

missing_packages <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]

if (length(missing_packages) > 0) {
  cat("Installing missing dependencies:", paste(missing_packages, collapse = ", "), "\n")
  install.packages(missing_packages)
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

```{r set-date}
start_date <- as.Date("2019-12-01")
end_date <- as.Date("2025-12-31")
```

```{r fetch-pse-data}
pal_raw <- read_csv("pal_stock.csv", show_col_types = FALSE) 

pal_df <- pal_raw %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y")) %>%
  
  filter(!is.na(Date)) %>%
  
  arrange(Date) %>%
  
  mutate(
    `Vol.` = case_when(
      str_detect(`Vol.`, "K") ~ as.numeric(str_remove(`Vol.`, "K")) * 1000,
      str_detect(`Vol.`, "M") ~ as.numeric(str_remove(`Vol.`, "M")) * 1000000,
      `Vol.` == "-" ~ 0,
      TRUE ~ as.numeric(`Vol.`)
    )
  ) %>%
  select(Date, Open, High, Low, Price, `Vol.`) %>% 
  rename(Close = Price, Volume = `Vol.`) %>%
  
  mutate(AdjClose = Close) 

pal_xts <- xts(pal_df[, -1], order.by = pal_df$Date)

colnames(pal_xts) <- c("PAL.Open", "PAL.High", "PAL.Low", "PAL.Close", "PAL.Volume", "PAL.Adjusted")

PAL <- pal_xts["2019-12/2025-12"]
```

```{r load-yahoo}
yahoo_tickers <- c(
  US_Equity = "UAL",     # US Equity: United Airlines
  Bitcoin = "BTC-USD",   # Bitcoin in USD
  SPY = "SPY",           # SPDR S&P 500 ETF Trust
  VIX = "^VIX"           # CBOE Volatility Index
)

getSymbols(yahoo_tickers, src = "yahoo", from = start_date, to = end_date)
```


```{r load-fred}
fred_tickers <- c(
  "DGS10",
  "FEDFUNDS",
  "CPIAUCSL",
  "DCOILWTICO"
)

getSymbols(fred_tickers, src = "FRED")

DGS10 <- window(DGS10, start = start_date, end = end_date)
FEDFUNDS <- window(FEDFUNDS, start = start_date, end = end_date)
CPIAUCSL <- window(CPIAUCSL, start = start_date, end = end_date)
DCOILWTICO <- window(DCOILWTICO, start = start_date, end = end_date)
```
```{r init-obs}
instruments <- c("UAL", "PAL", "BTC-USD", "SPY", "VIX", "DGS10", "FEDFUNDS", "CPIAUCSL", "DCOILWTICO")
init_obs <- function(ticker_list, sample_size = 3){
   for (sym in ticker_list) {
    data_obj <- get(sym)
    cat("INSTRUMENT:", sym, "| Total Observations:", nrow(data_obj), "\n")
    
    print(head(data_obj))
    cat("\n")
    
    na_indices <- which(rowSums(is.na(data_obj)) > 0)
    missing_count <- length(na_indices)
    
    cat("Total NAs:", missing_count, "\n\n")
    if (missing_count > 0) {
      for (idx in head(na_indices, sample_size)) {
        start_idx <- max(1, idx - 1)
        end_idx <- min(nrow(data_obj), idx + 1)
        print(data_obj[start_idx:end_idx, ])
        cat("------\n")
      }
      cat("\n")
    }
    
    dup_indices <- which(duplicated(zoo::index(data_obj)))
    dup_count <- length(dup_indices)
    cat("Total Duplicates:", dup_count, "\n\n")
    
    if (dup_count > 0) {
      for (idx in head(dup_indices, sample_size)) {
        start_idx <- max(1, idx - 1)
        end_idx <- min(nrow(data_obj), idx + 1)
        
        print(data_obj[start_idx:end_idx, ])
        cat("------\n")
      }
      cat("\n")
    }
  } 
}
init_obs(instruments)
```

```{r fix-local-na}
for (sym in instruments) {
  temp_data <- get(sym)
  temp_data <- na.locf(temp_data)
  assign(sym, temp_data)
}
```

```{r check-local-na-fix}
init_obs(instruments)
```

```{r convert-to-df}
df_list <- map(instruments, function(sym) {
  data_obj <- get(sym)
  
  # If it has an Adjusted column, extract it. Otherwise, use the raw data object.
  if (has.Ad(data_obj)) {
    extracted_data <- Ad(data_obj)
  } else {
    extracted_data <- data_obj
  }
  
  clean_df <- as.data.frame(extracted_data) %>% 
    rownames_to_column(var = "Date") %>% 
    mutate(Date = as.Date(Date)) %>%
    as_tibble()
  
  return(clean_df)
})
names(df_list) <- instruments

master_df <- df_list %>%
  compact() %>%
  reduce(full_join, by = "Date") %>%
  arrange(Date) %>%
  fill(-Date, .direction = "down")
```

This is to detect PAL's trading halt window. PAL was suspended by PSE due to audit rule violation.
```{r bea-pal-halt-detect}
pal_native <- df_list[["PAL"]] %>% 
  arrange(Date)

# find which dates have a gap greater than 10 -> will consider this as halt
pal_gap_idx <- which(as.numeric(diff(pal_native$Date)) > 10)

pal_halt_windows <- tibble(
  halt_start = pal_native$Date[pal_gap_idx] + 1,
  halt_end   = pal_native$Date[pal_gap_idx + 1] - 1
)

cat("Detected PAL trading halt window(s):\n")
print(pal_halt_windows)
cat("(Expected: 2021-06-17 to 2022-03-27 -- PSE audit-opinion suspension)\n")

in_pal_halt <- function(dates) {
  purrr::map_lgl(dates, function(d) any(d >= pal_halt_windows$halt_start & d <= pal_halt_windows$halt_end))
}
```

```{r log-returns}
master_df <- master_df %>%
  mutate(across(
    -Date, 
    ~ c(NA, diff(log(.x)))
  )) %>%
  rename_with(~ str_replace(.x, "\\.Adjusted$", ".LogAdj"), -Date)
```

```{r explore-master-df}
explore_instruments <- function(instruments, master_df) {
  for (sym in instruments) {
    cat("INSTRUMENT:", sym, "\n")
    
    instrument_df <- master_df %>% 
      select(Date, starts_with(sym))
    
    print(head(instrument_df), n = 3)
    print(tail(instrument_df), n = 3)
    cat("\n")
    
    cat("--- Missing Values per Column ---\n")
    instrument_df %>% 
      summarise(across(everything(), ~sum(is.na(.)))) %>% 
      print(width = Inf)
    cat("\n")
    
    missing_rows <- instrument_df %>%
      filter(if_any(starts_with(sym), is.na))
    
    total_missing <- nrow(missing_rows)
    cat("Total Rows with NAs:", total_missing, "\n")
    
    if (total_missing > 0) {
      cat("--- Top 10 Missing Rows ---\n")
      print(missing_rows, n = 10)
    }
    
    dup_count <- sum(duplicated(instrument_df$Date))
  }
  
  dplyr::glimpse(master_df)
}
explore_instruments(instruments, master_df)
```

```{r fix-missing-master-df}
master_df <- master_df %>%
  fill(-Date, .direction = "down") %>%
  filter(Date >= as.Date("2020-01-01"))

explore_instruments(instruments, master_df)
```

```{r bea-pal-masked-column}
# another column is created `PAL_masked.LogAdj` where its halted dates are masked to be NA
# this is based on PAL.LogAdj which all missing dates were forward filled,
# they will exist alongside each other, useful for more analysis later.
master_df <- master_df %>%
  mutate(PAL_masked.LogAdj = if_else(in_pal_halt(Date), NA_real_, PAL.LogAdj))

cat("PAL.LogAdj non-NA:", sum(!is.na(master_df$PAL.LogAdj)),
    "| PAL_masked.LogAdj non-NA:", sum(!is.na(master_df$PAL_masked.LogAdj)), "\n")
```

```{r function-descriptive-statistics}
analyze_instrument <- function(df, col_name) {
  inst_data <- df %>%
    select(Date, Value = all_of(col_name)) %>%
    filter(!is.na(Value))
  
  summary_stats <- tibble(
    Instrument = col_name,
    Mean = mean(inst_data$Value),
    Median = median(inst_data$Value),
    Variance = var(inst_data$Value),
    Std_Dev = sd(inst_data$Value),
    Min = min(inst_data$Value),
    Max = max(inst_data$Value),
    Range = Max - Min,
    Skewness = skewness(inst_data$Value), 
    Kurtosis = kurtosis(inst_data$Value)
  ) %>%
    mutate(across(where(is.numeric), ~round(., 4)))
  
  cat("\n======================================================\n")
  cat("### Daily Summary Statistics:", col_name, "\n")
  cat("======================================================\n")
  print(summary_stats)
  cat("\n")
  
  line_plot <- ggplot(inst_data, aes(x = Date, y = Value)) +
    geom_line(color = "darkblue", alpha = 0.8) +
    theme_minimal() +
    labs(
      title = paste("Daily Timeline of", col_name, "(2020 - 2025)"),
      x = "Date",
      y = "Value / Log Return"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold")
    )
  
  dist_plot <- ggplot(inst_data, aes(x = Value)) +
    geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "steelblue", color = "white", alpha = 0.7) +
    geom_density(color = "darkred", linewidth = 1) +
    theme_minimal() +
    labs(
      title = paste("Daily Distribution of", col_name, "(2020 - 2025)"),
      x = "Value / Log Return",
      y = "Density"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.title = element_text(face = "bold")
    )
  
  print(line_plot)
  print(dist_plot)
}

target_cols <- c(
  "UAL.LogAdj",
  "PAL.LogAdj",
  "PAL_masked.LogAdj",
  "BTC-USD.LogAdj",
  "SPY.LogAdj", 
  "VIX.LogAdj", 
  "DGS10", 
  "FEDFUNDS", 
  "CPIAUCSL",
  "DCOILWTICO"
)

for (col in target_cols) {
  analyze_instrument(master_df, col)
}
```
```{r name-df-list}
# df_list was built in instrument order in `convert-to-df` (pre full_join, pre fill).
# Name it so individual per-instrument tibbles can be pulled out by ticker for the
# join demonstrations below, instead of re-splitting the already-joined master_df.
names(df_list) <- instruments
```

```{r bea-leftjoin1-pre-obs fuel-pass-through}
cat("UAL:", nrow(df_list[["UAL"]]), "rows |",
    format(min(df_list[["UAL"]]$Date)), "to", format(max(df_list[["UAL"]]$Date)), "\n")
cat("PAL:", nrow(df_list[["PAL"]]), "rows |",
    format(min(df_list[["PAL"]]$Date)), "to", format(max(df_list[["PAL"]]$Date)), "\n")
cat("DCOILWTICO:", nrow(df_list[["DCOILWTICO"]]), "rows |",
    format(min(df_list[["DCOILWTICO"]]$Date)), "to", format(max(df_list[["DCOILWTICO"]]$Date)), "\n\n")
cat("Note: PSE (PAL) and NYMEX (WTI) calendars differ from NYSE (UAL).\n")
cat("A left_join on UAL's calendar will introduce NAs on days PAL/WTI didn't print,\n")
cat("which we carry forward via fill() rather than dropping UAL rows.\n")
```
The difference in rows could be explained by the following:
- UAL is based on NYSE weekday calendar, which is closed on weekends and US market holidays. 
- PAL is based on PSE weekday calendar, which is closed on weekends and PH holidays.
- PAL had a 285-day trading suspension imposed by the PSE between (2021-06-17 to 2022-03-28),
  as its financial statements had violated a rule. This was during COVID-19 travel restrictions.
- DCOILWTICO is from FRED instead of a stock market trading calendar. 
```{r bea-leftjoin1}
fuel_leftjoin_df <- df_list[["UAL"]] %>%
  select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  left_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  left_join(df_list[["DCOILWTICO"]] %>% select(Date, WTI = DCOILWTICO), by = "Date") %>%
  arrange(Date) %>%
  fill(PAL_Adjusted, WTI, .direction = "down") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted))),
    WTI_ret = c(NA, diff(log(WTI)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(WTI_ret))

cat("Rows kept (UAL trading calendar):", nrow(fuel_leftjoin_df), "\n\n")
print(fuel_leftjoin_df, n = 6)

ual_fuel_model <- lm(UAL_ret ~ WTI_ret, data = fuel_leftjoin_df)
pal_fuel_model <- lm(PAL_ret ~ WTI_ret, data = fuel_leftjoin_df)
print(summary(ual_fuel_model)$coefficients)
print(summary(pal_fuel_model)$coefficients)
```

```{r bea-leftjoin1-post-obs}
ual_oil_beta <- coef(ual_fuel_model)[["WTI_ret"]]
pal_oil_beta <- coef(pal_fuel_model)[["WTI_ret"]]
ual_oil_p <- summary(ual_fuel_model)$coefficients["WTI_ret", "Pr(>|t|)"]
pal_oil_p <- summary(pal_fuel_model)$coefficients["WTI_ret", "Pr(>|t|)"]

cat("Rows retained on UAL's calendar:", nrow(fuel_leftjoin_df), "\n")
cat("UAL oil-return coefficient:", round(ual_oil_beta, 4), "| p-value:", round(ual_oil_p, 4), "\n")
cat("PAL oil-return coefficient:", round(pal_oil_beta, 4), "| p-value:", round(pal_oil_p, 4), "\n")
```

- `fuel_leftjoin_df` - PAL's price forward-filled across every gap (including weekends, holidays, and the trading suspension)
- `fuel_leftjoin_df_masked` - weekends and holidays are forward-filled but the trading suspension is 
```{r bea-pal-masked-companion}

# dataframe where the trading halt (285-day suspension) are masked as NA
# this is built from the fuel_leftjoin_df which is forward-filled
fuel_leftjoin_df_masked <- fuel_leftjoin_df %>%
  mutate(PAL_ret = if_else(in_pal_halt(Date), NA_real_, PAL_ret)) %>%
  filter(!is.na(PAL_ret))

pal_fuel_model_masked <- lm(PAL_ret ~ WTI_ret, data = fuel_leftjoin_df_masked)
pal_oil_beta_masked <- coef(pal_fuel_model_masked)[["WTI_ret"]]
pal_oil_p_masked <- summary(pal_fuel_model_masked)$coefficients["WTI_ret", "Pr(>|t|)"]

cat("Naive (forward-filled) rows:", nrow(fuel_leftjoin_df), "| PAL oil-beta:", round(pal_oil_beta, 4), "\n")
cat("Masked (halt-excluded) rows:", nrow(fuel_leftjoin_df_masked), "| PAL oil-beta:", round(pal_oil_beta_masked, 4), "\n")
```

```{r bea-innerjoin-pre-obs BTC/airline correlation}
cat("BTC-USD:", nrow(df_list[["BTC-USD"]]), "rows | UAL:", nrow(df_list[["UAL"]]),
    "rows | PAL:", nrow(df_list[["PAL"]]), "rows\n")
cat("inner_join keeps only the intersection -- every BTC weekend/holiday move is\n")
cat("excluded by construction; loss is quantified explicitly below.\n")
```

```{r bea-innerjoin}
btc_innerjoin_df <- df_list[["BTC-USD"]] %>%
  select(Date, BTC_Adjusted = `BTC-USD.Adjusted`) %>%
  inner_join(df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`), by = "Date") %>%
  inner_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  arrange(Date) %>%
  mutate(
    BTC_ret = c(NA, diff(log(BTC_Adjusted))),
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(BTC_ret), !is.na(UAL_ret), !is.na(PAL_ret))

n_btc_total <- nrow(df_list[["BTC-USD"]])
n_common <- nrow(btc_innerjoin_df)
cor_matrix <- cor(btc_innerjoin_df %>% select(BTC_ret, UAL_ret, PAL_ret), use = "complete.obs")
print(round(cor_matrix, 4))
```

```{r bea-innerjoin-post-obs}
btc_ual_cor <- cor_matrix["BTC_ret", "UAL_ret"]
btc_pal_cor <- cor_matrix["BTC_ret", "PAL_ret"]
ual_pal_cor <- cor_matrix["UAL_ret", "PAL_ret"]

cat("Common trading days retained:", n_common, "of", n_btc_total, "BTC calendar days\n")
cat("BTC-UAL correlation:", round(btc_ual_cor, 4), "\n")
cat("BTC-PAL correlation:", round(btc_pal_cor, 4), "\n")
cat("UAL-PAL correlation:", round(ual_pal_cor, 4), "\n")
```

```{r bea-antijoin-pre-obs}
cat("PAL:", nrow(df_list[["PAL"]]), "rows | UAL:", nrow(df_list[["UAL"]]), "rows\n")
cat("anti_join surfaces every UAL trading day with no matching PAL date --\n")
cat("both routine PSE holidays and the 285-day suspension. Splitting the two\n")
cat("apart tests whether the suspension or routine calendar drift dominates.\n")
```

```{r bea-antijoin pal-ual-gap}
pal_gap_days <- df_list[["UAL"]] %>% select(Date) %>%
  anti_join(df_list[["PAL"]] %>% select(Date), by = "Date")

holiday_gap_days <- pal_gap_days %>% filter(!in_pal_halt(Date))
halted_gap_days  <- pal_gap_days %>% filter(in_pal_halt(Date))

cat("Total UAL trading days with no PAL match:", nrow(pal_gap_days), "\n")
cat("Routine PSE-holiday-only days:", nrow(holiday_gap_days), "\n")
cat("Suspension-only days:", nrow(halted_gap_days), "\n\n")
print(holiday_gap_days, n = 10)
```

```{r bea-antijoin-post-obs}
cat("Of", nrow(pal_gap_days), "total PAL gap days, which are dates that are in UAL but are not in the PAL dataset,", nrow(halted_gap_days),
    sprintf("(%.0f%%)", 100 * nrow(halted_gap_days) / nrow(pal_gap_days)),
    "are attributable to the 285-day suspension. Note that the 285-day suspension is the total days between 2021-06-17 to 2022-03-27, which includes weekends, but because PAL and UAL does not include weekends, which is why only 196 is attributable to the suspension. \n")
cat("Routine PSE-holiday mismatch accounts for only", nrow(holiday_gap_days), "days, which accounts for holidays starting from December 2019 to Decemer 2025. These are dates that PAL does not account for, but is included in the UAL dataset.\n")
```

# SQL-Based Financial Investigation

## SQL Investigation 1: Does Forward-Filling PAL's Suspension Bias Its Return Statistics?

**Objective:** Quantifies the how forward-filling PAL's 285-day suspension affects its measured return distribution, compared to excluding (masking) the halt window. Reuses the two parallel PAL series already built above (`fuel_leftjoin_df` naive, `fuel_leftjoin_df_masked` corrected).

```{r bea-sql1}
pal_stats_comparison <- bind_rows(
  fuel_leftjoin_df        %>% mutate(version = "Forward-filled (naive)") %>% select(version, PAL_ret),
  fuel_leftjoin_df_masked %>% mutate(version = "Halt-masked (corrected)") %>% select(version, PAL_ret)
) %>%
  group_by(version) %>%
  summarise(
    N = n(),
    Zero_Return_Pct = round(100 * mean(PAL_ret == 0), 1),
    Mean = round(mean(PAL_ret), 6),
    Median = round(median(PAL_ret), 6),
    Variance = round(var(PAL_ret), 6),
    SD = round(sd(PAL_ret), 4),
    Min = round(min(PAL_ret), 4),
    Max = round(max(PAL_ret), 4),
    Range = round(max(PAL_ret) - min(PAL_ret), 4),
    Skewness = round(skewness(PAL_ret), 3),
    Kurtosis = round(kurtosis(PAL_ret), 3),
    .groups = "drop"
  )

print(pal_stats_comparison, width = Inf)
```

**Financial Interpretation:** Forward-filling raises PAL's share of zero-return days from 25.8% to 35.5% (about 1.4x), this pulls its standard deviation down (0.0252 to 0.0235) and inflating kurtosis (13.1 to 15.1). The halt gets recorded as "196 calm trading days" compared to "no market existed." 

This mirrors a well-documented problem in the finance literature: stale/non-synchronous pricing biases variance and covariance estimates downward (Scholes & Williams, 1977; Dimson, 1979), and the same effect has been measured directly in markets with frequent long suspensions: correcting for suspension bias shrinks return-based factor estimates materially in Chinese A-share data (Tian, 2019). `fuel_leftjoin_df` (naive) stays the default elsewhere in this file; Investigations 2-4 below explicitly opt into `fuel_leftjoin_df_masked` where the corrected series matters to the result.

## SQL Investigation 2: Which Airline Is More Exposed to Oil: UAL or PAL?

**Objective:** Compare United Airlines' and Philippine Airlines' sensitivity to WTI crude oil returns, using the corrected (halt-masked) PAL series, to determine which airline's stock carries more fuel-cost risk.

```{r bea-sql2}
oil_direction_summary <- fuel_leftjoin_df_masked %>%
  mutate(oil_direction = if_else(WTI_ret > 0, "Oil Up", "Oil Down")) %>%
  group_by(oil_direction) %>%
  summarise(
    n_days = n(),
    mean_UAL_ret = round(mean(UAL_ret), 5),
    mean_PAL_ret = round(mean(PAL_ret), 5),
    .groups = "drop"
  ) %>%
  arrange(oil_direction)

print(oil_direction_summary)

cat("\nFull-sample regression betas (UAL from bea-leftjoin1; PAL from the masked companion):\n")
cat("UAL oil-beta:", round(ual_oil_beta, 4), "| p =", round(ual_oil_p, 4), "\n")
cat("PAL oil-beta:", round(pal_oil_beta_masked, 4), "| p =", round(pal_oil_p_masked, 4), "\n")
```

**Financial Interpretation:** Interestingly, although UAL's beta is *larger* in magnitude (+0.104) but *positively* signed, signifying that it moves with oil, consistent with oil and airline stocks both being pro-cyclical/demand-driven in a simple two-variable regression with no macro controls. PAL's beta is smaller (-0.078) but *negatively* signed, implying that the "rising fuel cost hurts an airline". So UAL shows the stronger statistical co-movement, but PAL shows the theoretically correct cost-channel relationship. 

## SQL Investigation 3: The Ukraine Invasion Oil Shock

**Objective:** Test whether UAL's oil-sensitivity shifts during the acute Feb 24-Mar 8, 2022 shock (WTI spiked from ~$92.77 to a $123.64 peak, +33%) relative to its full-sample beta. (Note: PAL has zero data for this exact window.)

```{r bea-sql3}
ukraine_shock_df_all_ual <- fuel_leftjoin_df %>%
  filter(between(Date, as.Date("2022-02-24"), as.Date("2022-03-08")))
ukraine_shock_df_pal <- fuel_leftjoin_df_masked %>%
  filter(between(Date, as.Date("2022-02-24"), as.Date("2022-03-08")))

cat("UAL trading days in Ukraine shock window:", nrow(ukraine_shock_df_all_ual), "\n")
cat("Of which PAL has real (non-halted) data:", nrow(ukraine_shock_df_pal), "\n\n")

shock_vs_full <- tibble(
  Period = c("Full sample", "Ukraine shock (2022-02-24 to 2022-03-08)"),
  N = c(nrow(fuel_leftjoin_df), nrow(ukraine_shock_df_all_ual)),
  UAL_oil_beta = c(
    round(cov(fuel_leftjoin_df$UAL_ret, fuel_leftjoin_df$WTI_ret, use = "complete.obs") / var(fuel_leftjoin_df$WTI_ret, na.rm = TRUE), 4),
    round(cov(ukraine_shock_df_all_ual$UAL_ret, ukraine_shock_df_all_ual$WTI_ret, use = "complete.obs") / var(ukraine_shock_df_all_ual$WTI_ret, na.rm = TRUE), 4)
  ),
  Mean_UAL_ret = c(round(mean(fuel_leftjoin_df$UAL_ret), 5), round(mean(ukraine_shock_df_all_ual$UAL_ret), 5)),
  Mean_WTI_ret = c(round(mean(fuel_leftjoin_df$WTI_ret), 5), round(mean(ukraine_shock_df_all_ual$WTI_ret), 5))
)
print(shock_vs_full, width = Inf)
```

**Financial Interpretation:** PAL has zero trading days in this window, its suspension (2021-06-17 to 2022-03-27) covers the entire shock. UAL investors could reprice the stock in real time against the oil spike; PAL investors could not trade at any price, favorable or not. That absence is itself the finding: a real-world illustration of market friction/information asymmetry, not just a data-cleaning footnote. UAL's own beta is also striking here: it *flips sign*, from +0.104 full-sample to -0.417 during the shock, alongside a -3.4% average daily return over those 9 days. That's consistent with the invasion being a broad risk-off/war-fear event that hit airline stocks directly (falling, alongside rising oil), temporarily overwhelming the smaller demand-channel co-movement seen in the full-sample average. (Small-sample caveat: only 9 trading days, so this is a noisy point estimate, but the direction and magnitude of the swing are large enough to be informative on their own.)

## SQL Investigation 4: Is UAL's Oil-Beta a Stable Relationship, or Just the Ukraine Shock?

**Objective:** Split the sample into pre-shock, shock-era, and recovery sub-periods and recompute UAL's and PAL's oil-beta in each, to test whether Investigation 2's relationship is a structural feature or an artifact of one event. UAL is grouped from the naive `fuel_leftjoin_df` (it has no gaps to mask, and restricting it to PAL's available days would silently exclude the actual shock dates, since they fall inside PAL's halt window); PAL is grouped from `fuel_leftjoin_df_masked`. The two are joined back together by period for a side-by-side table.

```{r bea-sql4}
period_label <- function(d) case_when(
  d < as.Date("2022-01-01") ~ "1. Pre-shock (2020-2021)",
  d < as.Date("2023-01-01") ~ "2. Shock era (2022)",
  TRUE ~ "3. Recovery (2023-2025)"
)

ual_period_beta <- fuel_leftjoin_df %>%
  mutate(period = period_label(Date)) %>%
  group_by(period) %>%
  summarise(
    n_days_ual = n(),
    UAL_beta = round(cov(UAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

pal_period_beta <- fuel_leftjoin_df_masked %>%
  mutate(period = period_label(Date)) %>%
  group_by(period) %>%
  summarise(
    n_days_pal = n(),
    PAL_beta = round(cov(PAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

period_beta <- ual_period_beta %>%
  left_join(pal_period_beta, by = "period") %>%
  arrange(period)

print(period_beta, width = Inf)
```

**Financial Interpretation:** UAL's beta is not stable. It swings from +0.178 (pre-shock) to -0.155 (shock era) to -0.020 (recovery), confirming that Investigation 2's full-sample beta (+0.104) is a blended average of genuinely different regimes, not a fixed structural relationship. The shock-era row is the most informative test of Investigation 3's finding: it includes the acute Ukraine-invasion days (unlike a PAL-restricted grouping would), and it does show a distinct, negative reading relative to the surrounding periods. PAL's beta stays negative throughout (-0.091, -0.083, -0.020), meaning its cost-channel sign (rising fuel prices squeezing PAL's margins, pulling its stock down) holds up more consistently than UAL's demand-channel co-movement (oil and airline stocks rising together with the broader economy) does. Its shock-era row does rest on only 193 real trading days against UAL's 251, though, a direct consequence of the suspension eating into exactly that period: the same missing-price-discovery issue from Investigation 3 showing up again.

## SQL Investigation 5: How Much Does PAL's Stagnant Period Deflate the UAL-PAL Correlation?

**Objective:** Directly measure how much a naive `full_join` + forward-fill deflates the UAL-PAL return correlation compared to a strict `inner_join`, closing the loop back to Investigation 1's methodology point and justifying `inner_join`'s use in `bea-innerjoin`.

```{r bea-sql5}
ual_pal_fulljoin_df <- df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  full_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  arrange(Date) %>%
  fill(UAL_Adjusted, PAL_Adjusted, .direction = "down") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(PAL_ret))

ual_pal_innerjoin_df <- df_list[["UAL"]] %>% select(Date, UAL_Adjusted = `UAL.Adjusted`) %>%
  inner_join(df_list[["PAL"]] %>% select(Date, PAL_Adjusted = `PAL.Adjusted`), by = "Date") %>%
  filter(format(Date, "%Y-%m") != "2019-12") %>%
  arrange(Date) %>%
  mutate(
    UAL_ret = c(NA, diff(log(UAL_Adjusted))),
    PAL_ret = c(NA, diff(log(PAL_Adjusted)))
  ) %>%
  filter(!is.na(UAL_ret), !is.na(PAL_ret))

naive_cor  <- cor(ual_pal_fulljoin_df$UAL_ret, ual_pal_fulljoin_df$PAL_ret, use = "complete.obs")
strict_cor <- cor(ual_pal_innerjoin_df$UAL_ret, ual_pal_innerjoin_df$PAL_ret, use = "complete.obs")

cat("Naive full_join + fill() correlation:", round(naive_cor, 4), "| N =", nrow(ual_pal_fulljoin_df), "\n")
cat("Strict inner_join correlation:", round(strict_cor, 4), "| N =", nrow(ual_pal_innerjoin_df), "\n")
```

**Financial Interpretation:** The naive version keeps a `full_join` across both calendars and forward-fills every gap, including the 285-day suspension, so N=1539 rows go into the naive correlation, many of them repeated/fabricated values rather than real paired price movements, pulling the coefficient toward whatever a flat series correlates at (near zero, by construction). The strict `inner_join` version drops to N=1206, using only days both stocks actually traded, so every observation is a real, paired price movement. This is the same deflation mechanism as Investigation 1, demonstrated on a pairwise statistic instead of a univariate one, and it's the direct justification for choosing `inner_join` over `full_join` for the BTC/UAL/PAL correlation work in `bea-innerjoin`.

## SQL Investigation 6: Do UAL and PAL Behave Differently in High-Inflation Periods?

**Objective:** Determine whether United and Philippine Airlines perform differently during periods of high inflation versus periods of low or moderate inflation.

```{r bea-sql6}
cpi_daily <- tibble(Date = seq(min(df_list[["CPIAUCSL"]]$Date), max(fuel_leftjoin_df$Date), by = "day")) %>%
  left_join(
    df_list[["CPIAUCSL"]] %>% arrange(Date) %>% mutate(CPI_YoY = CPIAUCSL / lag(CPIAUCSL, 12) - 1) %>% select(Date, CPI_YoY),
    by = "Date"
  ) %>%
  arrange(Date) %>%
  fill(CPI_YoY, .direction = "down")

label_inflation <- function(df) {
  df %>%
    left_join(cpi_daily, by = "Date") %>%
    filter(!is.na(CPI_YoY)) %>%
    mutate(inflation_regime = if_else(CPI_YoY > 0.04, "High inflation (>4% YoY)", "Low/moderate inflation (<=4% YoY)"))
}

ual_inflation_df <- fuel_leftjoin_df %>% label_inflation()
pal_inflation_df <- fuel_leftjoin_df_masked %>% label_inflation()

ual_inflation_summary <- bind_rows(
  ual_inflation_df %>% mutate(inflation_regime = "Overall (all regimes)"),
  ual_inflation_df
) %>%
  group_by(inflation_regime) %>%
  summarise(n_days_ual = n(), mean_UAL_ret = round(mean(UAL_ret), 5), sd_UAL_ret = round(sd(UAL_ret), 5), .groups = "drop")

pal_inflation_summary <- bind_rows(
  pal_inflation_df %>% mutate(inflation_regime = "Overall (all regimes)"),
  pal_inflation_df
) %>%
  group_by(inflation_regime) %>%
  summarise(n_days_pal = n(), mean_PAL_ret = round(mean(PAL_ret), 5), sd_PAL_ret = round(sd(PAL_ret), 5), .groups = "drop")

inflation_summary <- ual_inflation_summary %>%
  left_join(pal_inflation_summary, by = "inflation_regime") %>%
  arrange(inflation_regime)

print(inflation_summary, width = Inf)
```

**Financial Interpretation:** UAL's average daily return flips from +0.15% when inflation is low or moderate to -0.04% once inflation climbs above 4%. The swing from positive to negative lines up with intuition, that higher inflation cuts into consumer travel budgets while also raising UAL's operating costs like fuel, wages, and leases. PAL doesn't show the same pattern; its average return is negative in both regimes (-0.07% and -0.03%), so inflation on its own doesn't appear to be a major driver for PAL specifically. Another important thing to note is how PAL's high-inflation bucket only has 349 real trading days against UAL's 545, since that stretch overlaps heavily with the suspension, so there's less to go on there. Volatility barely changes for either airline across the two regimes, so the real story is a shift in average return, not the stock getting riskier.

## SQL Investigation 7: Does the Fed Funds Rate Environment Change UAL's and PAL's Oil-Beta?

**Objective:** Determine how the prevailing interest rate environment (near-zero, hiking, or restrictive) affects both airlines' stock performance and their sensitivity to oil prices.

```{r bea-sql7}
fedfunds_daily <- tibble(Date = seq(min(df_list[["FEDFUNDS"]]$Date), max(fuel_leftjoin_df$Date), by = "day")) %>%
  left_join(df_list[["FEDFUNDS"]] %>% select(Date, FedFunds = FEDFUNDS), by = "Date") %>%
  arrange(Date) %>%
  fill(FedFunds, .direction = "down")

label_rate_regime <- function(df) {
  df %>%
    left_join(fedfunds_daily, by = "Date") %>%
    filter(!is.na(FedFunds)) %>%
    mutate(rate_regime = case_when(
      FedFunds < 1 ~ "1. Near-zero (<1%)",
      FedFunds < 4 ~ "2. Hiking (1-4%)",
      TRUE ~ "3. Restrictive (>4%)"
    ))
}

ual_rate_df <- fuel_leftjoin_df %>% label_rate_regime()
pal_rate_df <- fuel_leftjoin_df_masked %>% label_rate_regime()

ual_rate_summary <- bind_rows(
  ual_rate_df %>% mutate(rate_regime = "Overall (all regimes)"),
  ual_rate_df
) %>%
  group_by(rate_regime) %>%
  summarise(
    n_days_ual = n(),
    mean_UAL_ret = round(mean(UAL_ret), 5),
    UAL_oil_beta = round(cov(UAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

pal_rate_summary <- bind_rows(
  pal_rate_df %>% mutate(rate_regime = "Overall (all regimes)"),
  pal_rate_df
) %>%
  group_by(rate_regime) %>%
  summarise(
    n_days_pal = n(),
    mean_PAL_ret = round(mean(PAL_ret), 5),
    PAL_oil_beta = round(cov(PAL_ret, WTI_ret, use = "complete.obs") / var(WTI_ret, na.rm = TRUE), 4),
    .groups = "drop"
  )

rate_summary <- ual_rate_summary %>%
  left_join(pal_rate_summary, by = "rate_regime") %>%
  arrange(rate_regime)

print(rate_summary, width = Inf)
```

**Financial Interpretation:** The overall row lines up exactly with what Investigation 2 already found (+0.104 for UAL, all 1504 rows), which is a good check that the regime split isn't dropping or duplicating anything. Break it out by Fed policy stance showcases how UAL's relationship with oil fades: +0.131 when rates are near zero, down to +0.074 during the hiking phase, and basically flat at -0.009 once policy turns restrictive. So oil and UAL's stock gradually stop moving together as the Fed tightens, rather than starting to move in opposite directions. PAL tells almost the same story in reverse: its beta goes from -0.095 (near-zero) to -0.059 (hiking) to -0.015 (restrictive), meaning its fuel-cost relationship is strongest exactly when money is cheap and weakens as rates climb. Put the two together and both airlines' link to oil shrinks as rates rise, regardless of whether that link starts positive or negative.
